### 2016-03-23 paper_ram.tex
first draft

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%2345678901234567890123456789012345678901234567890123456789012345678901234567890
%        1         2         3         4         5         6         7         8

\documentclass[letterpaper, 10 pt, conference]{ieeeconf}  % Comment this line out if you need a4paper

%\documentclass[a4paper, 10pt, conference]{ieeeconf}      % Use this line for a4 paper

\IEEEoverridecommandlockouts                              % This command is only needed if 
                                                          % you want to use the \thanks command

\overrideIEEEmargins                                      % Needed to meet printer requirements.

% See the \addtolength command later in the file to balance the column lengths
% on the last page of the document

% The following packages can be found on http:\\www.ctan.org
\usepackage{subfig}
\usepackage{graphicx} % for pdf, bitmapped graphics files
%\usepackage{epsfig} % for postscript graphics files
%\usepackage{mathptmx} % assumes new font selection scheme installed
%\usepackage{times} % assumes new font selection scheme installed
%\usepackage{amsmath} % assumes amsmath package installed
%\usepackage{amssymb}  % assumes amsmath package installed
\usepackage{url}
\usepackage{array}
\usepackage{tabularx}
\usepackage{tabulary}
\usepackage{booktabs}

\title{\LARGE \bf
The Thymio robot, open source hardware for formal education
}

\author{Francesco Mondada$^{1}$, Michael Bonani$^{2}$, Fanny Riedo$^{2}$, Philippe~Rétornaz and Stéphane Magnenat$^{1}$% <-this % stops a space
\thanks{*This research was supported by the Swiss National Center of Competence in Research ”Robotics”.}% <-this % stops a space
\thanks{$^{1}$Francesco Mondada and Stéphane Magnenat are with the Laboratoire de Syst\`emes Robotiques,
        Ecole Polytechnique F\'ed\'erale de Lausanne (EPFL), Switzerland
        {\tt\small francesco.mondada@epfl.ch} and {\tt\small stephane.magnenat@epfl.ch}}%
\thanks{$^{2}$Michael Bonani and Fanny Riedo are with Mobsya Association, Ecublens, Switzerland
        {\tt\small michael.bonani@mobsya.org} and {\tt\small fanny.riedo@mobsya.org}}%
}

\begin{document}

\maketitle
\thispagestyle{empty}
\pagestyle{empty}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\begin{abstract}
%Mobile robots are interesting tools for education because of both the enthusiasm they raise and the multidisciplinary nature of their technology.
%However, because of their price, complexity, bias towards specific groups (boys), and unproved educational value, they are still seldom used in schools and edutainment.
%This paper presents a system-design approach and the resulting Thymio robot that tackle these problems in a holistic way.
%We show how this strategy achieved a low price leading to a high demand and raising the interest of teachers.
%The robot and its development environment are open source and the whole project follows an open distribution and participative model.
%Finally, we show deployment statistics proving that the robot is gender- and age-neutral and preliminary results towards assessing the educational value of robots.
%\end{abstract}

\section{Introduction}

Mobile robots are interesting tools for education because of both the enthusiasm they raise and the multidisciplinary nature of their technology.
Because autonomous mobile robots sense the environment and take actions based on their perception, they seem to display \emph{intentions of their own}.
This fascinates the users of the robot and creates feelings of accomplishment and power in the creators.
The permeating presence of robots in science fiction and their projected use in our society increases these emotions by giving a sense of touching the future. % SM: not so happy with my sentence here
Children can spend hours looking at a robot interacting with its environment and adults can approach robots as a new hobby that replaces radio and electronics kits.
Robots embed various technologies and therefore give access to a wide range of fields, such as complex mechanics, sensors, wireless transmission, mathematics, computer science, etc.
The emotional potential of mobile robots and their varied technology makes them the ideal tool for education at large.

Despite this potential, robots are still not widespread in schools as they could be and this for several reasons. Among them we can mention:
\begin{enumerate}
\item Although there are many robotic projects developing educational robots, few reach a sufficient maturity to become distributed and accessible to schools. 
\item A robot performing interesting behaviors is a complex piece of technology and therefore expensive. 
Indeed, for having an educational value and providing an interesting level of interaction, a robot must embed a wide set of sensors and actuators.
Existing platforms with these features cost several hundred Euros.
This prevents most schools, which have a limited budget for equipment, to acquire interesting robots.
\item Introducing robotic tools into teaching activities requires investment in time and training for the teachers.
In Europe, despite a trend in including more and more technological tools into the learning process, teachers are still insufficiently trained and are reluctant to introduce these tools in their teaching activity~\cite{CERI2008}.
To be accepted by teachers, robots must be therefore both accessible with minimal effort and be associated with well prepared educational material. 
Moreover an infrastructure should allow the exchange of educational material among the teachers.
\item Robot construction, use, and programming is often perceived as a boyish activity in our society~\cite{leonard2009lego,nourbakhsh2009robot}.
% For instance, several authors report that girls have such a perception of the LEGO Minstorms~\cite{leonard2009lego,nourbakhsh2009robot}.
This strongly limits the potential of robots as general-purpose educational tools, especially in schools.
\item Finally, many teachers are reluctant to follow volatile trends, especially if based on purely commercial arguments. Teachers prefer to invest in stable tools, and this is the opposite of todays consumer technology, often very volatile and based on continuos renewal rather than on durability.
\end{enumerate}

Open source hardware projects can address several of these issues in a different way than pure commercial products. 
By open source hardware we understand, following the definition of the Open Source Hardware Association\footnote{\url{http://www.oshwa.org/definition/}}, ``hardware whose design is made publicly available so that anyone can study, modify, distribute, make, and sell the design or hardware based on that design''.
In this paper, we show that this concept, implemented as a project in a community of users, developers and producers, brings a very interesting added value to the robot and to the educational methods. 
This is illustrated by the results achieved with Thymio and is compared with the feedback of a group of engineers active in open source hardware design.
We can well observe both similarities and differences between our education-centered project and other open source hardware projects, pointing to challenges and opportunities.
More details on the robot and the related results can be found in conference papers~\cite{festival2011,thymio2012arsi}.

\section{Related Work}

There is a very large number of publications presenting educational robots with different prices and features, from very low-cost systems targeting, for instance, education in Africa~\cite{Rubenstein2015,Gyebi2015} to extremely sophisticated humanoids~\cite{Hood2015,Mazzoni2016}.
Among those, only very few are commercially available, limiting their validation by education scientists.
This results in 90\% of publications about validation of educational results made on LEGO Mindstorm~\cite{benitti2012exploring}.
The LEGO Mindstorm EV3\footnote{\url{http://mindstorms.lego.com}} is expensive ($\sim$400\,\$) but offers a wide range of possibilities, especially at the mechanical level using LEGO bricks and at the software level with its graphical programming environment. 
At a lower price, LEGO proposes the WeDo system ($\sim$80\,\$).
The WeDo is similar to EV3 but only reads one sensor and controls one actuator and must always be connected to a desktop computer.
The BeeBot robot\footnote{\url{http://www.beebot.org.uk}} is a cheap ($\sim$90\,\$) autonomous system.
However, it is not a real robot but rather an automaton: There is no perception of the environment, just a good encoder for precise movements and a simple programming interface based on buttons giving directions for the movements.

Among these widely used robots, none is part of an open source hardware project.
The only effort in this direction comes from LEGO, that made public all the schematics of the EV3 in an hardware developer kit, to allow the design of extensions. 
Otherwise, only few educational robots are both used in schools and are issued from an open hardware project:
Scribbler2, produced and sold by Parallax\footnote{\url{http://www.parallax.com}} ($\sim$180\,\$), is a 188~mm large robot, designed to run on the ground and equipped with few light sensors, one distance sensor, two ground sensors and few LEDs display.
It runs on standard AA batteries and has a hacker port for interfacing electronic extensions.
It is programmable with a graphic or text code interface.  
The main weakness of Scribbler2 is its performance per price ratio and the limited compatibility with other systems. 
The e-puck~\cite{mondada2009puck} robot is a robot targeting university-level education.
Well equipped with sensors and actuators, modular and compact, it can be programmed with professional environments.
Several simulators allow to run experiments of high complexity. 
It's main weakness is its high price ($\sim$870\,\$).
Finch\footnote{http://www.finchrobot.com/} ($\sim$99\,\$) is a very simple robot that has been designed around a wired connection to the computer. 
This connection reduces electronics requirements, such as batteries or wireless communication, and allows to implement the control entirely on the computer.
This results in a very broad set of possible programming languages available, the real force of this robot. 
On the other hand the cable does not allow real autonomy and mobility of the robot.
Finally the mBot\footnote{http://www.makeblock.cc/mbot/} is a mobile platform based on the Arduino board. 
The very cheap and simple electronics allow to drastically reduce the price of the system ($\sim$75\,\$) but also reduces the perception possibilities to a couple of sensors.

In respect to these robots, Thymio has a compact size (120~mm), many interaction possibilities, an affordable price and a large set of sensors.
It's design makes the best use of the recent cheap components coming from the explosive growth of the mobile devices market. 
To the best of our knowledge, beside Thymio there are no educational products exploiting this and providing many sensors and actuators at a price below 130\,\$.

For teachers, the motivation to use robotic tools in their classes depends on many factors \cite{chevalier2016}.
Among them, the availability of material and training plays a key role.
The BeeBot mentioned previously, because of its radical simplicity, is one of the tools that needs the least training. 
Moreover its use in learning activities is based on a simple grid, allowing fast and easy integration in existing lessons.
These are probably two important factors for its success in schools~\cite{2008highfield,2008demichele} and the availability of a rich set of educational material\footnote{\url{https://www.learningplace.com.au/deliver/content.asp?pid=38840} and \\ \url{http://beebots.skola.edu.mt/}}.
However, the total lack of sensors reduces the BeeBot to an automaton, easy to use but nearly useless to introduce computational concepts.

For more complex robots, both the development of educational material and the training of teachers requires a very important effort. 
Beside the development of the material from a pedagogical point of view, education is very dependent from local educational programs and languages. 
In the particular case of Switzerland, for instance, every of the 26 states composing the confederation has its own school program, and among the states there are three major official languages. 
Educational material and the related training has therefore to be adapted for each situation.
Even for large companies like LEGO this development is not affordable. 
LEGO itself is moving in the direction of promoting communities of users \cite{Hienerth2014} and a form of user-producer interaction that is close to those of open hardware projects.
The open source community regrouping developers, producers and final users is therefore a very interesting model to address the distributed development and sharing of educational material, and the diffusion of training sessions.
Communities around projects like Arduino are extremely good examples of this type of mechanism \cite{Jamieson2010}.

%Several initiatives have focused on increasing the rate of females in engineering in general and in robotics in particular~\cite{weinberg2007impact}.
%Roberta~\cite{2010bredenfeld} is one of the largest and uses the LEGO Minstorms.
%Recent work has shown that a different approach to robotics, for instance linking with arts and proposing creative workshops~\cite{nourbakhsh2009robot}, can improve the attractiveness for girls. 
%Some researchers have created hardware tailored to crafting activities, such as the PicoCricket~\cite{2008rusk} which is close to the NXT but does not require LEGO construction.
%Compared to these works, Thymio aims at being attractive for both genders.

%Several studies have measured the impact of the use of robots on children~\cite{weinberg2007impact,nourbakhsh2009robot,sinkele2010effectiveness}, but none of these provide quantitative data on the usage of the different features of the robot.
%On the contrary, Thymio can log the usage of its features, providing a tool to monitor the progress of children and to investigate the value of different curricula.

\section{Design choices}

\begin{figure}
   \centering
	\includegraphics[width=.6\columnwidth]{figures/front_2016}
	\vskip 1.5em
	\includegraphics[width=\columnwidth]{figures/Thymio_II_sensors-actuators-eng}
	\caption{The Thymio robot and its main components.}
	\label{fig:schematique}
\end{figure}

We designed the Thymio robot along seven main axes: a low price to address a larger number of users, a feature set that suits both genders and multiple ages from young children to adults, a combination of sensors, actuators, and programming features that facilitates learning, a mechanical design that allows creative extensions, a set of ready-to-use behaviors to get a demonstration right out of the box, an accessible programming environment, and an open source  community contributing to the design and disseminating the results.
The result is a miniature (10\,cm) differential-wheeled robot suited for use on a desktop.
The robot is robust enough to be mishandled by children, as it can fall from a table without breaking itself.
It features a translucent white hull and a wide range of sensors and actuators (see \ref{fig:schematique}).
The robot embeds a battery, rechargable by \textsc{usb}, that provides an autonomy of about 2 hours.

\subsection{Low Price}

Our main concern during the mechatronic design of Thymio was to keep a low production cost. 
Price is a primary factor for the adoption of the robot by schools.
In our previous experience with designing robots, most of the costs was due to the electronics and the sensors~\cite{thesis_michael}, not to the mechanical parts.
As we wanted Thymio to be an affordable robot, we selected electronic components with great care and explored different design solutions in order to limit the number of expensive components.
In particular, we focused on low-cost sensors that allow rich interaction possibilities, with both the environment and the user.
The resulting Thymio robot possesses a large number of sensors:
Seven horizontal infrared distance sensors, with a range of about 13\,cm.
They enable many interactions with the environment, like following an object or avoiding an obstacle.
Two ground infrared sensors allow to follow a line or to detect the edge of a table.
A three-axis accelerometer measures the orientation of the robot with respect to gravity, and can detect shocks and free falls.
Five capacitive touch buttons organized as an arrow form an intuitive user interface.
Compared to physical buttons, these simplify the plastic hull of the robot and make it more robust.
A remote control receiver provides additional distant buttons.
A microphone can record sounds and detect claps. 
Finally, a thermistor measures temperature.
Most of these sensors cost less then 0.2\,\$, the most expensive being the accelerometer with a cost of about 0.8\,\$, which is an acceptable price given the possibilities it brings to the robot.
The omnipresence of this type of sensors in smartphones and compact cameras has drastically lowered their prices over the last decade.
We also chose low-cost toy motors and control them in speed (max. 13\,cm/s) using a custom-designed electronics.
By measuring the back-electromotive force, speed control is possible without additional encoders.

However, choosing the optimal combination of sensors and processor is a difficult problem, as hardware interfaces are limited on a microcontroller and therefore compromises must be made.
We evaluated several microcontrollers and chose the PIC24F from Microchip because it integrates a USB interface and can drive capacitive touch buttons directly, saving additional components. 
This microcontroller controls all sensors and actuators, excepted the battery recharging logic, which uses a specific chip for safety reasons.
This choice of a single microcontroller led us to non-trivial design decisions like using low-cost shift registers to drive the many LEDs.
Moreover, instead of directly connecting the infrared sensors to an analogical input, we interface them using a comparator that transforms the analogical signal into a binary one with a pulse having a duration proportional to the intensity of the original signal~\cite{wanda}. 
This design saves \textsc{cpu}-time, reduces power consumption (it allows a shorter infrared pulse than with an analogical design) and allows to use the sensor for communication.

In addition, we conceived Thymio to fit the requirements of series production.
This is a very critical choice in open source hardware projects, and we will discuss this issue in detail in the next section.
We produced until now more than 10'000 robots by batches of 1'000 units.
Because manual work is expensive in western countries, we produce through subcontractors in China.
Indeed, while most of the electronics consists of \textsc{smd} components mounted by a robot, some are soldered (sensors, microphone, LEDs) by hand and the robot is also assembled by hand.
We streamlined the production of plastic parts (Thymio has only 5 parts while our previous prototype had 11) and opted for injection molding. 
The plastic hull consists of two main parts screwed together to facilitate the vertical assembly and disassembly of the robot.
Moreover, when selecting features we took into consideration the complexity of the injection mold.

\subsection{Multi-age and Gender-neutral Feature Set}

While we optimized Thymio for cost, we also targeted a large range of users.
The variety of sensors and ways of interacting with the robot make Thymio accessibly to girls and boys of different age groups.
The white hull is initially neutral and children can choose their own color using the \textsc{rgb} LEDs.
The robot can be used from children as young as 6 years old to adults:
The youngest play with the robot through a set of basic behaviors (see section \ref{sec:behaviors}) and the older can program it (see section \ref{sec:aseba}).
The different basic behaviors are selectable using the capacitive buttons and show some of the capabilities of the robot, such as obstacle avoidance, slope climbing, etc.
By interacting with the different behaviors, for instance by blocking the robot with the hand and seeing it avoiding it, children can discover the concept of the sensory-motor loop. 
As every behavior has a different color, children can also understand that different \emph{programs} run on the robot.
Children around 6 can use Thymio as a support for handicrafts or construction.
For example, they can create kinetic sculptures.
For children above 9 years old and adults, Thymio is suited to introduce programming.
Learners can start by programming new interactive behaviors and can go on with constructing around the robot and programming a behavior that interplays with the construction.
High-school and university students can use the robot to learn advanced robotics concepts through the integration with \textsc{ros}~\cite{quigley2009ros}.
Finally, because of the open-source and open-hardware nature of the project, advanced programmers can improve their skills in C by modifying the firmware or in electronics by disassembling the robot.

\subsection{Facilitating Learning}

When designing Thymio, we took care of providing many incentives for the users to learn new things throughout their direct interaction with the robot.
This translates into specific hardware and software choices.

At the hardware level, we render visible the activity of the sensors by adding a LED next to each, for a total of 39 LEDs.
These LEDs locally color the hull and allow the user to see immediately where and when the robot perceives a change in its environment.
There is a LED next to each proximity sensor that lights up as soon as an object is close enough to be seen, and shines brighter the closer the object gets.
Similarly, a combination of blue and red LEDs show the temperature, and the infrared remote control receiver and the microphone both have LEDs that flash when they detect something.
On the top of the robot, a circle of 8 LEDs shows the 3D inclination of the robot thanks to the accelerometer. 
This circle is also used to reflect some of the behaviors of Thymio.
Finally, two strong \textsc{rgb} LEDs color the whole top of the robot.
In addition to a visual feedback, the capacitive buttons also have an acoustic feedback.
The link between a sensor and its feedback can be turned off when programming the robot to use the LEDs and the loudspeaker for other purposes.

At the software level, our programming environment (see section \ref{sec:aseba}) provides a set of tools enabling beginners to discover progressively the basic rules of programming using a graphical interface, then the construction of a syntax by graphic blocs, and finally a full coding environment with advanced debugging tools, such as real-time inspection of the variables of the robot and plotting features, providing a visual way to understand time-related concepts.

\subsection{Promoting Creativity}

We wanted the robot to be a starting point for the users to invent their own creations.
Therefore, we thought Thymio as a support for handicrafts and constructions.
Its white color is meant to be seen as a blank page that can be decorated and drawn upon, and its shape allows easy integration into a structure.
The robot features a hole in its middle to insert a marker for drawing.
It also has a slot for an SD card, allowing users to add their own sounds or to load a pre-compiled code by the simple insertion of a dedicated card.

The square format of the hull allows to use the robot as a base for driving the user's own constructions.
To that end, Thymio is compatible with LEGO bricks:
There are four attachment positions on the top of the robot and two LEGO Technics crosses on each side.
Additionally, the two wheels have attach points, which permits to use them to actuate elements elsewhere in the structure (see Figure \ref{fig:example-construction}, bottom left) or to lift the robot's own weight (see Figure \ref{fig:example-construction}, bottom right).
Therefore we chose more powerful motors than strictly necessary to move the robot around.
Both the hull and the wheels can be easily disassembled and replaced or augmented with 3D printed customized parts.
Figure \ref{fig:3Dparts} shows an example of 3D printed accessory designed by a member of the community.
Finally, it is possible to link several Thymio by software, allowing the creation of complex robotic structures.

\begin{figure}
\centering
\includegraphics[width=.9\columnwidth]{figures/3Dparts}
\caption{Accessory (cable winder) built by 3D printing by a member of the community, Ezio Som\`a.}
\label{fig:3Dparts}
\end{figure}

\subsection{Short time to demo}
\label{sec:behaviors}

Many existing robots need to be built or configured before having the first operational behavior. 
This is a barrier for many users and we wanted to avoid it carefully, having a robot able to show interesting behaviors right out of the box.
Therefore Thymio provides 6 different basic behaviors (see table \ref{tbl:basic-behaviors}), which are accessible as soon as the robot is started.
These allow people encountering Thymio for the first time to immediately interact with it, while providing insights into the many possibilities of the robot.
The behaviors are interactive and oriented towards the user. % SM: FIXME: not clear what "oriented towards the user" means
For each behavior, the robot's body has a different color, allowing the user to easily recognize it.
The user can navigate between the colors with the buttons and select the behavior she wants to use.
Basic behaviors can be exploited in a construction to create reactivity without the use of programming (see Figure \ref{fig:example-construction}, top).

\begin{table}
\begin{tabularx}{\columnwidth}{@{}llll@{}}
\toprule
mode & color & sensors & behavior \\
\midrule
friendly & green & infrared (IR) & follows an object at distance \\
explorer & yellow & IR & moves avoiding obstacles \\
fearful & red & acc., IR & flees, notifies shocks and falls \\
investigator & cyan & IR & follows a black track \\
obedient & magenta & buttons, IR & follows moving orders \\
attentive & blue & mic. & moves following sound \\
\bottomrule
\end{tabularx}
\caption{The different basic behaviors with sensors used.}
\label{tbl:basic-behaviors}
\end{table}

\subsection{Programming}
\label{sec:aseba}

Thymio runs the Aseba open-source programming environment~\cite{aseba}.
Aseba is designed to enable novices to program robots easily.
On the robot side, it provides a lightweight virtual machine that runs on microcontrollers like the PIC24F.
A virtual machine allows instantaneous upload and safe execution of code.
On the desktop side, Aseba provides an IDE which allows to program in a purely graphical mode, or using a scripting language (see Figure \ref{fig:aseba-studio}) or in a mixed way, assembling code in a graphical way using the blockly environment\footnote{\url{https://developers.google.com/blockly/}}.

%The Aseba scripting language is a simple, imperative, Matlab-like language providing integers and arrays of integers as data types.
%Common logical and mathematical operations are available as operators, and additional mathematical functions such as trigonometric ones are available as native functions.
%These are functions implemented in native code and accessible from the virtual machine.
%In Aseba, code blocks are associated to events, simplifying the writing of real-time behaviors.
%Events are triggered by the robot's sensors or actuators.
%The value of the sensors and actuators are available through pre-defined variables and native functions, which are robot-dependent and enumerated dynamically by the robot when the IDE connects.

The IDE integrates a help documentation\footnote{currently English, French, and German}.
In addition, the wiki pages of the project provide tutorials on programming with Thymio and a detailed description of the programming interface.
%The latter was designed with consistency in mind: variables and functions are available in namespaces and similar features (such as the different LEDs) are accessed the same way.
%It also provides native functions for school-level mathematics such as vector operations, trigonometry, statistics and functions such as sorting and random.

%We expect the facilities that Aseba provides to allow text-based programming with reasonably young children, starting from around 10 years old, while still being interesting to older children and adults.
Aseba integrates with \textsc{ros}~\cite{quigley2009ros} through the \emph{asebaros}\footnote{\url{http://www.ros.org/wiki/asebaros}} bridge.
\textsc{Ros} is one of the most widely used software framework in robotics research, and this integration allows to run sophisticated algorithms, such as simultaneous localization and mapping, in conjunction with  Thymio.
This makes the robot suitable for university-level education.

\begin{figure}
\centering
\includegraphics[width=.9\columnwidth]{figures/aseba-studio}
\caption{Aseba Studio, the integrated development environment}
\label{fig:aseba-studio}
\end{figure}

\section{Open source hardware, choices and impact}

The choice of developing a robot within an open source hardware project has an impact on the robot design and the way this device is used in the community of users. 
In this section we analyze more in detail the implications of this choice in the context of educational robotics.
We compare these implications with the result of a survey we did among 25 people active in various open source hardware projects based on a worldwide call for contributions. 
Among these 25 people, 6 are project leaders, 10 are part of the core design team, 6 are contributors and 3 are enthusiastic users.
60\% of these respondents are between 25 and 35 years old and 80\% either work in an academic environment or are students.

\subsection{Motivation}

When asked about their motivation to participate to open source hardware projects, the people participating to the survey give as main motivation a link to their professional activity and to the specific project, followed by a more general motivation by the nature of the project and finally by the engineering practice. 
A more abstract goal of improving our society is placed at the last position, as mentioned only by few respondents.

\begin{figure}
\centering
\includegraphics[width=.7\columnwidth]{figures/motivation}
\caption{Motivation of the people participating to the survey.}
\label{fig:motivation}
\end{figure}

In the Thymio project, most contributors were hired for a research project developing the robot. 
Therefore the link with the professional activity and the particular project is evident.
On another side, the motivation of contributing to society is much stronger in our team, as developing a robot targeting education is a societal goal also.

It is more interesting to look at what people expect as benefits in participating in such a project.
Most of the answers mention impact, networking and learning. 
In these answers we can observe both a technological component in getting better results and impact, and a relational component that is not in the basic definition of open hardware but results from the community created around the project. 
People expect to work with other people to achieve more together, to create a network of competences or users, to learn through new experience and contacts.

In the Thymio project we had similar expectations, but with a very important additional element; we wanted a match between the non-profit nature of the project and the non-profit nature of education in general.

\begin{figure}
\centering
\includegraphics[width=.7\columnwidth]{figures/getout}
\caption{Motivation of the people participating to the survey.}
\label{fig:getout}
\end{figure}

\subsection{License of project and license of tools}

When starting an open source hardware project, one of the typical questions is about which license to use when disseminating the project source files. 
We will not discuss this matter here, as it is a very common and well covered issue.

There is another licensing issue which is not well known and that we discovered very late in our project: the constrains of the license of the mechanical and electronic CAD tools. 
Indeed, asked about this issue, people in the field seems generally not aware of the fact that CAD license can be very restrictive in the way the source files can be published.
Around the half or respondents to our survey, also among the project leaders, state that they were not aware that this could be an issue, and only one third checked the license of their CAD software. 
Among the respondents, some stated that they did not checked the license because they were sure they can freely publish their design, but half of them are using software not allowing the publication of source files when using educational licenses, and all these respondents are academic or students.
This issue is very serious, as most contributors of open source hardware projects are academic and use academic or educational licenses. 
Most licenses do not allow to use the design created with this version of the software for commercial activities. 
As producing or selling the product is part of the definition of open hardware, these licenses simply forbid publication under the standard open source hardware conditions.

\begin{figure}
\centering
\includegraphics[width=.9\columnwidth]{figures/aware-license}
\caption{Awarness about CAD tool licenses limitations by people participating to the survey.}
\label{fig:aware}
\end{figure}

To clarify this issue we contacted twelve of the major editors of mechanical CAD and PCB routing software.
We asked them if their educational license allows the publication of the source files, and we specified that the published files could have been downloaded by a company to produce the system.
Figure \ref{fig:editors} summarize the result of this survey.
Only three of the twelve editors have licenses allowing this type of publication. 
Two others explicitly mentioned the possibility to ask for permission before publication.
A large mechanical CAD editor was puzzled by our questions and after realizing the impact of the license, introduced a special condition allowing publication of files in clearly labeled open source hardware projects.
In previous situations of open source publication, the same editor asked the universities to purchase commercial licenses to permit publication.
This can multiply by a factor of some hundreds the price of the CAD license.

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{figures/table}
\caption{Publication possibilities as function of the CAD editors.}
\label{fig:editors}
\end{figure}

As a conclusion, this issue is totally underestimated by both people participating to the projects and by the CAD editors. 
In a period where editors are looking for additional revenues and are attacking universities for abuse of licenses\footnote{We are aware, just in Switzerland, of two situations where large amounts of money are asked to universities for non-respect of software licenses.}, this can be a very dangerous situation.

\subsection{Who produces the hardware}

Among the fundamental choices when starting an open source hardware project, there is the type of production. 
In the definition of open hardware we gave, it is stated that we should offer an ``hardware whose design is made publicly available so that anyone can ... make... the ... hardware ''.
Behind the ``anyone'' should we consider every single person or only every company who want to produce the product?
We opted for the second interpretation, for a reason of price and complexity of the product. 
When looking for the best performance per price ratio, we decided to use techniques that require heavy equipment.
For instance we decided to produce all mechanical parts by injection molds. 
This ensures a price per part of some cents, which cannot be achieved by others techniques.
Letting end-user produce their own parts would results in higher prices for less performances. 
When looking at the results of the survey (see Figure \ref{fig:production}), we can observe that this seems to be the choice of a majority of the projects represented in our survey.
Indeed, large open source hardware projects like Arduino are based on this model.

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{figures/production}
\caption{Production methods in the projects mentioned by the people participating to the survey.}
\label{fig:production}
\end{figure}

One could advocate that this choice of central production and sale of hardware can bring these robotic projects closer to standard commercial models, with users not feeling close to the developers.
In our case we decided to counterbalance this apparently business oriented model by creating a non-profit association in charge of producing and selling the robots. 
Of course anybody else could produce the Thymio based on more buiness-oriented models, but at least the initial production has been done respecting a coherent and global non-profit strategy, in line with educational values of the community of users.

Another strong difference between a standard consumer product and our vision of open hardware, is that Thymio should be durable.
Schools do not have large budget for technological tools and make strong investment in training of their teachers when adopting a new tool.
Therefore the life time of the products should be as long as possible. 
The open hardware approach gives to the user, or to a generic technician, better conditions to repair the system.
Supporting this type of operation has an impact on the robot design; for instance Thymio can be easily opened with four standard screws, and we introduced connectors between key elements such as motors, speaker, or the battery and the main printed circuit.

Another key element in supporting repair by the users is calibration. 
When choosing very low cost components, we get into large dispersion of characteristics. 
In our case, for instance, the motors can have very different characteristics, resulting in the robot not going straight when asked to apply the same speed on both wheels. 
To correct this, one interesting solution is to introduce factory calibration, storing correction factors that are specific for each robot, in the microcontroller memory . 
To allow the user to replace a broken motor, it is essential to give him also the possibility to re-calibrate the robot and adjust the parameters of the new motors.
This results in the design of calibration processes that can be performed by anyone, getting close to the original definition of open source hardware.

\subsection{Validated positive results}

Pushed by universities, like INRIA

Example of canton de Geneve

XXXXXXXXXXXXXXXXXX

\subsection{Problems}

Management of the community

Documentation in many languages, due to end-users that are child

Development of the project in the long term

We are building an open community around the robot.
A wiki\footnote{http://www.thymio.org} serves as a meeting point for the learners, the robot developers, and the teachers.
It introduces the robot, explains the basic behaviors, gives access to the programming environment and its documentation, and provides code samples and examples of constructions.
The wiki is open for editing by anyone, and although we initially provided most of the material, the other members of the community are starting to contribute.
We expect it to ease the sharing of educational materials from the teachers interested in the project.
To further contribute to the synergy between teachers and scientists, Thymio can write usage statistics on an \textsc{sd} card using a manual and voluntary operation.
This will allow us to assess the usefulness of the different features of the robot in a quantitative way.

We distribute the complete source code of both Thymio and the programming environment as \emph{open source} on the wiki.
Furthermore, the mechanical and electronic design is available as \emph{open hardware}.
Beside being of educational value, anyone can see how a robot works from inside, this further reduces the production cost as we did not deposit any expensive patent.
Finally, the Thymio robot is produced by a non-profit association, reducing the cost of the manufacturing and the production management.

\begin{figure}
\includegraphics[width=.49\columnwidth]{figures/poisson-thymio}\hfill\includegraphics[width=.49\columnwidth]{figures/top-traineau}
\vskip .5em
\includegraphics[width=.49\columnwidth]{figures/moving-heli}\hfill\includegraphics[width=.49\columnwidth]{figures/funi-overview}
\caption{Examples of constructions using Thymio}
\label{fig:example-construction}
\end{figure}

\subsection{Impact and validation}

\begin{table}
\centering
\begin{tabularx}{\columnwidth}{ll}
\toprule
Description & Price (USD)\\
\midrule
electronic components & 15\,\$ \\
microcontroller & 4 \$ \\
motors & 2 \$ \\
plastic parts (hull, wheels, and light guides) & 4 \$ \\
assembling & 8 \$ \\
transport & 2 \$ \\
\bottomrule
Total & 35 \$ \\
\end{tabularx}
\caption{Production cost of Thymio.}
\label{tbl:thymio-price}
\end{table}

We reached our price target thanks to our careful selection of components and our production in China.
Thymio costs 35\,\$ per unit to produce in batch of 1\,k units (see \ref{tbl:thymio-price}), with fixed fees of 15\,k\,\$ (13\,k\,\$ for the mold and 2\,k\,\$ for the \textsc{ce} certification).
This allows a selling price of 100\,CHF ($\approx$ 106\,\$).
This small ratio of selling-over-production price is made possible by our non-profit business model.

\begin{table}
\begin{tabularx}{\columnwidth}{@{}lll@{}}
\toprule
School & School type & Town in Switzerland \\
\midrule
HEPL (several units) & Teacher training & Lausanne  \\
EP Cossonay-Penthalaz & Primary school & Penthalaz  \\
EPS Ecublens & Primary and Secondary & Ecublens  \\
Collège de Vevey & Secondary & Vevey  \\
Collège Sainte Croix & Secondary & Fribourg  \\
Fondation de Verdeil & Specialized school & Lausanne  \\
Gymnase de Beaulieu & Baccalaureate school & Lausanne  \\
Gymnase de Provence & Baccalaureate school & Lausanne  \\
Lycée Blaise-Cendras & Baccalaureate school & La Chaux-de-Fonds  \\
edurobot.ch & Network of teachers & Vaud  \\
\bottomrule
\end{tabularx}
\caption{Schools and groups of teachers who develop educational material for the Thymio.}
\label{tbl:schools}
\end{table}

We deployed more than 800 Thymio in less than one year of activity.
The robot has been distributed in workshops at the occasion of public events like the \textsc{epfl} robotics festival~\cite{festival2011} or the \textsc{eth} Zürich Scientifica and in public schools.
The reception by the public was extremely positive: workshop slots (\textsc{epfl}:\,284, \textsc{eth}:\,125) were all completely filled and additional robots (\textsc{epfl}:\,155, \textsc{eth}:\,82) that we made available for purchase were sold out.
More than 40 teachers acquired a Thymio in their professional context.
Ten schools (see \ref{tbl:schools}) are exploiting hundred Thymio and are actively developing educational material in collaboration with us.
These schools target different ages and educational levels.
At the primary-school level, for instance, teachers use Thymio to explore the notion of space and the scientific methods.
They appreciate the integration of rational and spacial thinking that Thymio allows, but also its possibilities to concretely illustrate abstract concepts, such as directions in the teaching of foreign languages.
They use the sensors and their link to behavior changes as support for the methodological understanding of a phenomenon by children. %SM: FIXME: not so clear: methodological understanding 
Informal feedback from teachers shows that the direct visualization of sensor values on the robot's body dramatically increases the understanding of the working principles of the sensors themselves and of the robot's behavior in general. 
%This decreases the time spent in explaining the robot functionalities.
We attribute this effect to the good capacity of humans to detect correlations, allowing them to intuitively link variations in the environment, sensor responses and changes in the behavior. 
Most of the Thymio robots sold in schools are used in baccalaureate programming courses. 
In this context, Thymio offers an exceptional functionalities-by-price ratio and is used in computer-science classes to learn programming.
The aids of the programming environment, like real-time visualization of sensor data, are very appreciated.

%The current deployment speed is limited by our production capacity, and we will make the robot available through a web shop once we are able to produce fast enough.

% couleur + viewing of behaviour, immediatety, real-time understanding
% why robot detects table edge
% human capabilities to detect correlations

\begin{figure}
\includegraphics{figures/age-hist}
\includegraphics{figures/sex-wrt-age}
\includegraphics{figures/appreciation-wrt-age}
\caption{Distribution of ages (top), gender in function of age (middle), and appreciation in function of age (bottom).}
\label{fig:results-festival}
\end{figure}

At the occasion of the \textsc{epfl} robotics festival, we have conducted a survey to assess the appreciation of the robot and the educative contribution of a 80\,min programming workshop~\cite{thymio2012arsi}, whose goal was to introduce children to basic robotics concepts.
To quantitatively compare the different variables of the survey, we use the Spearman's rank correlation test.
A total of 116 participants took part in the workshop and filled the survey.
Most are children between 8 and 14 years old (\ref{fig:results-festival}, top).
91 participants are boys, 22 girls, and 3 unknown.
The distribution of gender slightly varies with age, with a trend of having more girls at younger ages (\ref{fig:results-festival}, middle, p-value=0.04).
The appreciation of girls does not correlate with age (p-value=0.48).
Therefore we attribute the decreasing number of girls with age to the prejudice in our western societies against girls in technical fields, which would affect the decision of older girls and their parents, although we might assume that the interest in robotics should a priori be even for boys and girls.
The appreciation is very high: the worst note is 4 (10 participants) on a scale of 1 to 6, while 67 participants (57\,\%) give 6.
The general appreciation correlates neither with age nor with gender (\ref{fig:results-festival}, bottom, p-values 0.44 and 0.63).
In this survey, we also measured the acquired concepts through a small test.
The results show that children have understood specific and practiced concepts such as the infrared sensor, but failed to understand fundamental robotics concepts such as the sensory-motor loop.
This workshop was run by scientists, rather than by professional teachers.

\section{Conclusion}

Most of the current robotic tools for education capitalize upon existing hardware platforms, programming environments, or robot architectures.
Because they do not consider all elements at stake, these tools fail to meet the challenges of the field.
On the contrary, in this paper we presented a system-design approach that tackles these questions in a holistic way.
The result is the Thymio robot and its surrounding development environment and distribution model.
Thymio is extremely low-cost with respect to its performances, reflecting our design choices and an open and non-profit development and distribution model.

% Thymio is well documented and allows simple deployment in classes, which simplifies the task of teachers and reduces their training needs.
% Both Thymio itself and its programming environment provide real-time inspection of the sensor values, allowing an intuitive understanding of robotic processes.
% These in turn can be used as examples or metaphors of complex real-life systems and interactions.
% Furthermore, the deployment of Thymio in many schools will allow a long-term analysis of the use of the robot to assess quantitatively the usefulness of the different features.
% 
We have validated that Thymio is useful to address the challenges of the field.
There is a high demand for the robot, proving that its price is sufficiently low to make it attractive to a large public.
Ten schools are using Thymio and developing curriculums for it, proving its ability to attract the interest of teachers.
Through public workshops, we have shown that children can learn new things using the robot in about one hour only.
We have shown statistics demonstrating that Thymio is a gender- and age-neutral robot.
Finally, we have shown preliminary results on the concepts that children can acquire through their interaction with the robot.
This builds a path towards a sound investigation of the educational value of robots.

\bibliography{ThymioEdu}
\bibliographystyle{ieeetr}

%-\addtolength{\textheight}{-12cm}   % This command serves to balance the column lengths
                                  % on the last page of the document manually. It shortens
                                  % the textheight of the last page by a suitable amount.
                                  % This command does not take effect until the next page
                                  % so it should come on the page before the last. Make
                                  % sure that you do not shorten the textheight too much.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\section*{APPENDIX}
%
%questionnaire?
%
%\section*{ACKNOWLEDGMENT}
%
%??

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\end{document}

### 2016-03-23 paper_ram.tex
draft

@@
 
-\subsection{Validated positive results}
+\subsection{Validated added value of open source hardware}
 
@@
 
-XXXXXXXXXXXXXXXXXX
+XXXXXXXXXXXXXXXXXX what is behind this line is historical....
 
@@
 This small ratio of selling-over-production price is made possible by our non-profit business model.
-
-\begin{table}
-\begin{tabularx}{\columnwidth}{@{}lll@{}}
-\toprule
-School & School type & Town in Switzerland \\
-\midrule
-HEPL (several units) & Teacher training & Lausanne  \\
-EP Cossonay-Penthalaz & Primary school & Penthalaz  \\
-EPS Ecublens & Primary and Secondary & Ecublens  \\
-Collège de Vevey & Secondary & Vevey  \\
-Collège Sainte Croix & Secondary & Fribourg  \\
-Fondation de Verdeil & Specialized school & Lausanne  \\
-Gymnase de Beaulieu & Baccalaureate school & Lausanne  \\
-Gymnase de Provence & Baccalaureate school & Lausanne  \\
-Lycée Blaise-Cendras & Baccalaureate school & La Chaux-de-Fonds  \\
-edurobot.ch & Network of teachers & Vaud  \\
-\bottomrule
-\end{tabularx}
-\caption{Schools and groups of teachers who develop educational material for the Thymio.}
-\label{tbl:schools}
-\end{table}
 

### 2016-03-25 paper_ram.tex
first draft version

@@
 \title{\LARGE \bf
-The Thymio robot, open source hardware for formal education
+Thymio, an example of open source hardware robot \\
+for formal education
 }
 
-\author{Francesco Mondada$^{1}$, Michael Bonani$^{2}$, Fanny Riedo$^{2}$, Philippe~Rétornaz and Stéphane Magnenat$^{1}$% <-this % stops a space
-\thanks{*This research was supported by the Swiss National Center of Competence in Research ”Robotics”.}% <-this % stops a space
-\thanks{$^{1}$Francesco Mondada and Stéphane Magnenat are with the Laboratoire de Syst\`emes Robotiques,
+\author{Francesco Mondada$^{1}$, Michael Bonani$^{2}$, Fanny Riedo$^{2}$, Manon Briod$^{1}$, \\
+L\'ea Pereyre$^{1}$, Philippe R\'etornaz$^{1}$ and St\'ephane Magnenat$^{1}$% <-this % stops a space
+\thanks{*This research was partially supported by the Swiss National Center of Competence in Research ``Robotics'' (Thymio robot development and deployment), partially by GebertRuf Stiftung (design of accessories), partially by the Swiss NSF project CRAGP2 151543 ``Robotics in schools'', and partially by the EU-FP7 project ASSISIbf, no.
+601074 (survey on open hardware).
+Many thank to Didier Roy at INRIA, Gordana Gerber for the educational material and all the Mobsya team.}% <-this % stops a space``anyone''
+\thanks{$^{1}$Francesco Mondada and St\'ephane Magnenat are with the Laboratoire de Syst\`emes Robotiques,
         Ecole Polytechnique F\'ed\'erale de Lausanne (EPFL), Switzerland
@@
 \subsection{Promoting Creativity}
+\label{sec:crea}
 
@@
 There are four attachment positions on the top of the robot and two LEGO Technics crosses on each side.
-Additionally, the two wheels have attach points, which permits to use them to actuate elements elsewhere in the structure (see Figure \ref{fig:example-construction}, bottom left) or to lift the robot's own weight (see Figure \ref{fig:example-construction}, bottom right).
+Additionally, the two wheels have attach points, which permits to use them to actuate elements elsewhere in the structure (see Figure~\ref{fig:example-construction}, third row left) or to lift the robot's own weight (see Figure~\ref{fig:example-construction}, third row right).
 Therefore we chose more powerful motors than strictly necessary to move the robot around.
-Both the hull and the wheels can be easily disassembled and replaced or augmented with 3D printed customized parts.
-Figure \ref{fig:3Dparts} shows an example of 3D printed accessory designed by a member of the community.
-Finally, it is possible to link several Thymio by software, allowing the creation of complex robotic structures.
+The same fixation points used for LEGO bricks can be used to attach paper structures.
+Paper can allow to simply change the body shape and add some body movements, as illustrated in Figure~\ref{fig:example-construction} by the orca, opening and closing its mouth while moving forward or by the bat, moving its wings.
+But paper and cardboard can also radically change the locomotion principle, as illustrated in the second row of Figure~\ref{fig:example-construction} by the zombie, where the robot's wheel activate the legs.
+The paper structure can be also used to interact with the sensors, as illustrated in the second row of Figure~\ref{fig:example-construction} by the bear, who extends his paw in front of the sensors to drive its iceberg.
+The same fixation points can be used to attach 3D printed customized parts, as illustrated by the winder shown in the fourth row of Figure~\ref{fig:example-construction}.
+Finally one can use paper to create environments, either flat with patterns that can be used in association to the ground sensors (see Figure~\ref{fig:example-construction} bottom) or 3D object, like the trees beside the zombie in Figure~\ref{fig:example-construction}.
+Finally, it is also possible to link several Thymio by software, allowing the creation of complex robotic structures.
 
 \begin{figure}
-\centering
-\includegraphics[width=.9\columnwidth]{figures/3Dparts}
-\caption{Accessory (cable winder) built by 3D printing by a member of the community, Ezio Som\`a.}
-\label{fig:3Dparts}
+\includegraphics[width=.49\columnwidth]{figures/orca}\hfill\includegraphics[width=.49\columnwidth]{figures/bat}
+\vskip .5em
+\includegraphics[width=.49\columnwidth]{figures/zombie}\hfill\includegraphics[width=.49\columnwidth]{figures/ours}
+\vskip .5em
+\includegraphics[width=.49\columnwidth]{figures/moving-heli}\hfill\includegraphics[width=.49\columnwidth]{figures/funi}
+\vskip .5em
+\includegraphics[width=.49\columnwidth]{figures/both_half_winder}\hfill\includegraphics[width=.49\columnwidth]{figures/winder_on_thymio2}
+\vskip .5em
+\includegraphics[width=\columnwidth]{figures/5senses}
+\caption{Examples of extensions of the Thymio basic robot with paper or cardboard body extensions (top four images), using LEGO$^{TM}$ structural extensions (third row), using 3D-printed extensions (fourth row) or using a printed environment (bottom).}
+\label{fig:example-construction}
 \end{figure}
@@
 The user can navigate between the colors with the buttons and select the behavior she wants to use.
-Basic behaviors can be exploited in a construction to create reactivity without the use of programming (see Figure \ref{fig:example-construction}, top).
+Basic behaviors can be exploited in a construction to create reactivity without the use of programming (see Figure~\ref{fig:example-construction}, top).
 
@@
 A virtual machine allows instantaneous upload and safe execution of code.
-On the desktop side, Aseba provides an IDE which allows to program in a purely graphical mode, or using a scripting language (see Figure \ref{fig:aseba-studio}) or in a mixed way, assembling code in a graphical way using the blockly environment\footnote{\url{https://developers.google.com/blockly/}}.
+On the desktop side, Aseba provides an IDE which allows to program in a purely graphical mode, or using a scripting language (see Figure~\ref{fig:aseba-studio}) or in a mixed way, assembling code in a graphical way using the blockly environment\footnote{\url{https://developers.google.com/blockly/}}.
 
@@
 
-In the Thymio project we had similar expectations, but with a very important additional element; we wanted a match between the non-profit nature of the project and the non-profit nature of education in general.
+In the Thymio project we had similar expectations. 
+Working together with several partners was for everybody a win-win situation, and creating a community of users was the only solution to allow the development of high quality accessories and educational material.
+A wiki\footnote{http://www.thymio.org} has been established as meeting point for the learners, the robot developers, and the teachers.
+It introduces the robot, explains the basic behaviors, gives access to the programming environment and its documentation, and provides code samples and examples of constructions.
+The wiki is open for editing by anyone, and although we initially provided most of the material, the other members of the community have started to contribute.
+
+In addition to those elements, in our project we expected another important benefit from open hardware in term of image: we wanted a match between the non-profit nature of the project and the non-profit nature of education in general.
 
@@
 We asked them if their educational license allows the publication of the source files, and we specified that the published files could have been downloaded by a company to produce the system.
-Figure \ref{fig:editors} summarize the result of this survey.
+Figure~\ref{fig:editors} summarize the result of this survey.
 Only three of the twelve editors have licenses allowing this type of publication. 
@@
 This can multiply by a factor of some hundreds the price of the CAD license.
+This blocking factor for open hardware also apply to the publication of scientific results, as promoted by many governments in the last decade and generally called ``open science''.
+The change of policy in CAD files publication that we reached is a sign of the very positive trend set by open hardware and by open science in general.
 
@@
 
-\subsection{Who produces the hardware}
+\subsection{Who designs and produces the hardware}
 
@@
 In the definition of open hardware we gave, it is stated that we should offer an ``hardware whose design is made publicly available so that anyone can ... make... the ... hardware ''.
-Behind the ``anyone'' should we consider every single person or only every company who want to produce the product?
-We opted for the second interpretation, for a reason of price and complexity of the product. 
+Behind the ``anyone'' should we consider every single person or only companies able to produce the product?
+This choice has implications on who designs the system and how.
+In our project we have two different types of hardware: the robot itself and the accessories used in specific activities.
+For the robot itself we opted for the second interpretation of the definition, considering under ``anyone'' only the professional structures able to produce hardware in a professional way, for a reason of price and complexity of the product.
 When looking for the best performance per price ratio, we decided to use techniques that require heavy equipment.
@@
 Letting end-user produce their own parts would results in higher prices for less performances. 
-When looking at the results of the survey (see Figure \ref{fig:production}), we can observe that this seems to be the choice of a majority of the projects represented in our survey.
-Indeed, large open source hardware projects like Arduino are based on this model.
+This choice has another implication: only a core team of highly skilled engineers can contribute to the project. 
+It is not trivial to design a mechanical part that can be molded win the simplest possible way.
+
+For the accessories, less technically challenging and stronger linked with creativity and educational value, we promoted techniques that are accessible to ``anyone'' in the broader sense: paper, cardboard, LEGO constructions and 3D printing, as explained in section~\ref{sec:crea} and illustrated in Figure~\ref{fig:example-construction}.
+This allows a much broader spectrum of contributor, especially among the teachers.
 
@@
 
-One could advocate that this choice of central production and sale of hardware can bring these robotic projects closer to standard commercial models, with users not feeling close to the developers.
-In our case we decided to counterbalance this apparently business oriented model by creating a non-profit association in charge of producing and selling the robots. 
+When looking at the results of the survey (see Figure~\ref{fig:production}), we can observe that the most common production method is a centralized one by commercial companies.
+Indeed, large open source hardware projects like Arduino are based on this model.
+One could advocate that this choice of central production and sale of hardware can bring these robotic projects closer to standard commercial models, ending in users not involved in the project and not feeling close to the developers.
+In our case we decided to counterbalance this apparently business oriented model by creating a non-profit association, called Mobsya, in charge of producing and selling the robots. 
+This is also a very well represented production method in our survey.
 Of course anybody else could produce the Thymio based on more buiness-oriented models, but at least the initial production has been done respecting a coherent and global non-profit strategy, in line with educational values of the community of users.
+Moreover the user are better integrated is the second layer of open source hardware consisting in accessories, where anybody can contribute with own designs.
 
@@
 
-Another key element in supporting repair by the users is calibration. 
+Another key element in supporting repair by the users are the calibration methods. 
 When choosing very low cost components, we get into large dispersion of characteristics. 
@@
 To allow the user to replace a broken motor, it is essential to give him also the possibility to re-calibrate the robot and adjust the parameters of the new motors.
-This results in the design of calibration processes that can be performed by anyone, getting close to the original definition of open source hardware.
+In Thymio, this results in the design and the documentation of calibration processes that can be performed by anyone, getting close to the original definition of open source hardware.
 
@@
 
-Pushed by universities, like INRIA
+Beside the sale of more than 10'000 robots by a non-profit organization producing it, there are several elements that show a specific advantage of having chosen an open source hardware model.
+The first element is the very high acceptance of the robot by universities and their offices in charge of promotion of science.
+Although these are not the primary user targeted by the robot, these are institutions with high visibility and influence on school programs. 
+In particular INRIA in France and the University of Cambridge in he United Kingdoms decided to use Thymio as tool for promotion of their activities and introduced it into schools.
+In France, INRIA developed several teaching units and supported the training of more than 1000 teachers to the use of this particular robot. 
+This finally resulted in having the Thymio robot included as one of the reference tools for the education of digital science in the reference books of the French Academy of Science and its foundation for promotion of science in education, ``La main \`a la p\^ate.''
+The fact that Thymio is open source allows these institution to have a better control on the tool they use, as they can ensure maintenance and even further development of it.
 
-Example of canton de Geneve
+In some cases, the open source nature of this project was a crucial political argument for the choice of this platform.
+In the state of Geneva, the government decided to move toward open source non proprietary tools in education.
+All primary school- and most secondary school computers have been installed with the Ubuntu distribution of Linux. 
+On this operating system the LEGO Mindstorms software tools are not available, and their proprietary nature is not compatible with the philosophy dictated by the government. 
+Thymio was therefore extremely welcome, and the state of Geneva organized own training courses for teachers, ported several educational materials into their curriculum and organized a lending system for all schools that cannot afford to buy the robots.
 
-XXXXXXXXXXXXXXXXXX what is behind this line is historical....
+Finally the non-profit nature of the project, including the production infrastructure, has contributed to reduce the obstacle toward the acquisition of this platform for many schools, especially at the beginning of the process.
 
-\subsection{Problems}
-
-Management of the community
-
-Documentation in many languages, due to end-users that are child
-
-Development of the project in the long term
-
-We are building an open community around the robot.
-A wiki\footnote{http://www.thymio.org} serves as a meeting point for the learners, the robot developers, and the teachers.
-It introduces the robot, explains the basic behaviors, gives access to the programming environment and its documentation, and provides code samples and examples of constructions.
-The wiki is open for editing by anyone, and although we initially provided most of the material, the other members of the community are starting to contribute.
-We expect it to ease the sharing of educational materials from the teachers interested in the project.
-To further contribute to the synergy between teachers and scientists, Thymio can write usage statistics on an \textsc{sd} card using a manual and voluntary operation.
-This will allow us to assess the usefulness of the different features of the robot in a quantitative way.
-
-We distribute the complete source code of both Thymio and the programming environment as \emph{open source} on the wiki.
-Furthermore, the mechanical and electronic design is available as \emph{open hardware}.
-Beside being of educational value, anyone can see how a robot works from inside, this further reduces the production cost as we did not deposit any expensive patent.
-Finally, the Thymio robot is produced by a non-profit association, reducing the cost of the manufacturing and the production management.
+\subsection{Problems of open-source project}
 
 \begin{figure}
-\includegraphics[width=.49\columnwidth]{figures/poisson-thymio}\hfill\includegraphics[width=.49\columnwidth]{figures/top-traineau}
-\vskip .5em
-\includegraphics[width=.49\columnwidth]{figures/moving-heli}\hfill\includegraphics[width=.49\columnwidth]{figures/funi-overview}
-\caption{Examples of constructions using Thymio}
-\label{fig:example-construction}
+\centering
+\includegraphics[width=\columnwidth]{figures/problems}
+\caption{Problems observed by participants to open source hardware projects.}
+\label{fig:problems}
 \end{figure}
 
-\subsection{Impact and validation}
+The very specific structure of an open project has also drawbacks. 
+Asked about problems met in the project, the participants to our survey mention the management of the community, the evolution of the project, the efforts in documentation, production and distribution and the limited resources.
+There are here two kind of problems: one of management of people, the other in the development of the project after the first design.
+The management of the project seems a recurrent problem and is related to the nature of this type of structure, mainly built by academic people with engineering background. 
+Most of the projects, following our survey, are managed by the founder of the initiative and not by the most expert in the field, which is perhaps not the best objective choice.
+It is very difficult to evolve from a phase where the engineering is the core of the project, to a new phase where communication and management play an increasing role.
+The same holds for several other factors, like the efforts in documenting, producing and distributing the product. 
+Also here there is a change of activity from the very appealing design phase, to a phase of documentation and, in some cases, production and distribution which is not at all the same work and requires different skills.
+This requires extra efforts and motivation, which are often missing, especially for jobs like documentation that are not appreciated by engineers.
+In general many respondents to our survey identified the development of the project as a general issue, which summarizes the several aspects mentioned above.
+
+In the Thymio project we had very similar issues and tried to addess them by having several partners, each of them specialized in one aspect: design, production and sale, creation of educational material, etc.
+Despite this approach, we had also problems of management of the community, and issues in finding the right long term vision, especially having so many partners. 
+The long term vision of a sales person is not the same of an engineer, and those confrontation result in conflicts in a community with a very flat hierarchy. 
+A key element that allowed to keep cohesion in our community was to share, among all partners, strong basic values about improvement of society through education.
+
+XXXXXXX historical, should we keep this?
 
@@
 
-We deployed more than 800 Thymio in less than one year of activity.
-The robot has been distributed in workshops at the occasion of public events like the \textsc{epfl} robotics festival~\cite{festival2011} or the \textsc{eth} Zürich Scientifica and in public schools.
-The reception by the public was extremely positive: workshop slots (\textsc{epfl}:\,284, \textsc{eth}:\,125) were all completely filled and additional robots (\textsc{epfl}:\,155, \textsc{eth}:\,82) that we made available for purchase were sold out.
-More than 40 teachers acquired a Thymio in their professional context.
-Ten schools (see \ref{tbl:schools}) are exploiting hundred Thymio and are actively developing educational material in collaboration with us.
-These schools target different ages and educational levels.
-At the primary-school level, for instance, teachers use Thymio to explore the notion of space and the scientific methods.
-They appreciate the integration of rational and spacial thinking that Thymio allows, but also its possibilities to concretely illustrate abstract concepts, such as directions in the teaching of foreign languages.
-They use the sensors and their link to behavior changes as support for the methodological understanding of a phenomenon by children. %SM: FIXME: not so clear: methodological understanding 
-Informal feedback from teachers shows that the direct visualization of sensor values on the robot's body dramatically increases the understanding of the working principles of the sensors themselves and of the robot's behavior in general.
-%This decreases the time spent in explaining the robot functionalities.
-We attribute this effect to the good capacity of humans to detect correlations, allowing them to intuitively link variations in the environment, sensor responses and changes in the behavior. 
-Most of the Thymio robots sold in schools are used in baccalaureate programming courses. 
-In this context, Thymio offers an exceptional functionalities-by-price ratio and is used in computer-science classes to learn programming.
-The aids of the programming environment, like real-time visualization of sensor data, are very appreciated.
-
-%The current deployment speed is limited by our production capacity, and we will make the robot available through a web shop once we are able to produce fast enough.
-
-% couleur + viewing of behaviour, immediatety, real-time understanding
-% why robot detects table edge
-% human capabilities to detect correlations
-
-\begin{figure}
-\includegraphics{figures/age-hist}
-\includegraphics{figures/sex-wrt-age}
-\includegraphics{figures/appreciation-wrt-age}
-\caption{Distribution of ages (top), gender in function of age (middle), and appreciation in function of age (bottom).}
-\label{fig:results-festival}
-\end{figure}
-
-At the occasion of the \textsc{epfl} robotics festival, we have conducted a survey to assess the appreciation of the robot and the educative contribution of a 80\,min programming workshop~\cite{thymio2012arsi}, whose goal was to introduce children to basic robotics concepts.
-To quantitatively compare the different variables of the survey, we use the Spearman's rank correlation test.
-A total of 116 participants took part in the workshop and filled the survey.
-Most are children between 8 and 14 years old (\ref{fig:results-festival}, top).
-91 participants are boys, 22 girls, and 3 unknown.
-The distribution of gender slightly varies with age, with a trend of having more girls at younger ages (\ref{fig:results-festival}, middle, p-value=0.04).
-The appreciation of girls does not correlate with age (p-value=0.48).
-Therefore we attribute the decreasing number of girls with age to the prejudice in our western societies against girls in technical fields, which would affect the decision of older girls and their parents, although we might assume that the interest in robotics should a priori be even for boys and girls.
-The appreciation is very high: the worst note is 4 (10 participants) on a scale of 1 to 6, while 67 participants (57\,\%) give 6.
-The general appreciation correlates neither with age nor with gender (\ref{fig:results-festival}, bottom, p-values 0.44 and 0.63).
-In this survey, we also measured the acquired concepts through a small test.
-The results show that children have understood specific and practiced concepts such as the infrared sensor, but failed to understand fundamental robotics concepts such as the sensory-motor loop.
-This workshop was run by scientists, rather than by professional teachers.
+XXXXXXX end of historical
 
@@
 
+The introduction of robots in formal education is a very challenging task, not only because of technical requirements such as low cost, performances and interactivity, but also because of factors depending on the school environment, such as the diversity of the educational programs, the dependence on local structures and languages, or the required training of teachers.
 Most of the current robotic tools for education capitalize upon existing hardware platforms, programming environments, or robot architectures.
-Because they do not consider all elements at stake, these tools fail to meet the challenges of the field.
-On the contrary, in this paper we presented a system-design approach that tackles these questions in a holistic way.
-The result is the Thymio robot and its surrounding development environment and distribution model.
-Thymio is extremely low-cost with respect to its performances, reflecting our design choices and an open and non-profit development and distribution model.
+Because they do not consider all elements at stake, these tools fail to meet some challenges of the field.
+On the contrary, in this paper we presented a global approach that tackles these questions in a holistic way, based on open source and non-profit principles.
+The result is the Thymio robot and its surrounding open source community including engineers, art designers, production and sales people, and teachers.
+The open source hardware and the related contributions are split in two categories: the robot itself, produced by a non-profit organization and requiring very advanced skills in both design and production, and the accessories for activities, based on techniques broadly accessible and with a larger panel of contributors.
 
-% Thymio is well documented and allows simple deployment in classes, which simplifies the task of teachers and reduces their training needs.
-% Both Thymio itself and its programming environment provide real-time inspection of the sensor values, allowing an intuitive understanding of robotic processes.
-% These in turn can be used as examples or metaphors of complex real-life systems and interactions.
-% Furthermore, the deployment of Thymio in many schools will allow a long-term analysis of the use of the robot to assess quantitatively the usefulness of the different features.
-% 
-We have validated that Thymio is useful to address the challenges of the field.
-There is a high demand for the robot, proving that its price is sufficiently low to make it attractive to a large public.
-Ten schools are using Thymio and developing curriculums for it, proving its ability to attract the interest of teachers.
-Through public workshops, we have shown that children can learn new things using the robot in about one hour only.
-We have shown statistics demonstrating that Thymio is a gender- and age-neutral robot.
-Finally, we have shown preliminary results on the concepts that children can acquire through their interaction with the robot.
-This builds a path towards a sound investigation of the educational value of robots.
+By making a survey among a panel of people contributing to open source hardware projects, we could observe that our project shares some characteristics with the majority of the projects represented in the survey.
+We also had problems of management of the community, we chose production and distributions methods that are broadly applied, and we share most of the motivation elements that are behind other projects.
+We identified an underestimated legal issue for open source hardware projects in the licensing term of CAD software.
+Indeed only very few editors have licenses allowing the publication of source files created with educational license.
+Two surveys among the users and the CAD editors shows that both people contributing or leading open source hardware projects and the editors themselves are now aware of this issue.
+Finally we could show some elements that differentiate our project from other open source hardware ones, being specific to educational robotics.
+In particular our project takes advantage from an alignment between the principles underlying open source project and the nature of education institutions. 
+This fits with a political trend toward open source in education and in science in general.
+We also found some solutions to the struggling question of production methods by splitting our hardware in two categories, enabling both very advanced hardware and inclusion of a broad community of end users in education.
 

### 2016-03-27 paper_ram.tex
modifs mineures

@@
 We can well observe both similarities and differences between our education-centered project and other open source hardware projects, pointing to challenges and opportunities.
-More details on the robot and the related results can be found in conference papers~\cite{festival2011,thymio2012arsi}.
+More details on the robot and the related results can be found in~\cite{RiedoPhD, magnenat2014}.
 
@@
 The BeeBot mentioned previously, because of its radical simplicity, is one of the tools that needs the least training. 
-Moreover its use in learning activities is based on a simple grid, allowing fast and easy integration in existing lessons.
+Moreover its use in learning activities is based on a simple grid on a printed mat on the ground, allowing fast and easy integration in existing lessons.
 These are probably two important factors for its success in schools~\cite{2008highfield,2008demichele} and the availability of a rich set of educational material\footnote{\url{https://www.learningplace.com.au/deliver/content.asp?pid=38840} and \\ \url{http://beebots.skola.edu.mt/}}.
-However, the total lack of sensors reduces the BeeBot to an automaton, easy to use but nearly useless to introduce computational concepts.
+However, the total lack of sensors reduces the BeeBot to an automaton, easy to use but limited to very simple activities for young kids, without much potential in teaching computational thinking or robotics concepts.
 
-For more complex robots, both the development of educational material and the training of teachers requires a very important effort. 
-Beside the development of the material from a pedagogical point of view, education is very dependent from local educational programs and languages. 
-In the particular case of Switzerland, for instance, every of the 26 states composing the confederation has its own school program, and among the states there are three major official languages. 
-Educational material and the related training has therefore to be adapted for each situation.
+For more complex robots, both the development of educational material and the training of teachers requires a very important effort, requiring a good mix between robotics and educational skills. 
+Moreover, educational material changes very much from school to school, as education is very dependent from local educational programs and languages, for instance. 
+In the particular case of Switzerland, for instance, every of the 26 states composing the confederation has its own school program and its own language among the four official ones. 
+Educational material and the related training has therefore to be developed for each situation, which requires a huge effort.
 Even for large companies like LEGO this development is not affordable. 
-LEGO itself is moving in the direction of promoting communities of users \cite{Hienerth2014} and a form of user-producer interaction that is close to those of open hardware projects.
+A solution is to have an active community of users contributing to the development of the material in a distributed manner, adapting the material to the local situation.
+LEGO itself is moving in the direction of promoting communities of users \cite{Hienerth2014} with a user-producer interaction close to those of open hardware projects.
 The open source community regrouping developers, producers and final users is therefore a very interesting model to address the distributed development and sharing of educational material, and the diffusion of training sessions.
@@
 
-We designed the Thymio robot along seven main axes: a low price to address a larger number of users, a feature set that suits both genders and multiple ages from young children to adults, a combination of sensors, actuators, and programming features that facilitates learning, a mechanical design that allows creative extensions, a set of ready-to-use behaviors to get a demonstration right out of the box, an accessible programming environment, and an open source  community contributing to the design and disseminating the results.
+We designed the Thymio robot along seven main axes: a low price to address a larger number of users, a feature set that suits both genders and multiple ages from young children to adults, a combination of sensors, actuators, and programming features that facilitates learning, a mechanical design that allows creative extensions by the users themselves, a set of ready-to-use behaviors to get a demonstration right out of the box, an accessible programming environment, and an open source  community contributing to the design and disseminating the results.
 The result is a miniature (10\,cm) differential-wheeled robot suited for use on a desktop.
@@
 
+We start by presenting here the first six main axes, before moving to the next section devoted the open hardware specificities.
+
 \subsection{Low Price}
@@
 Our main concern during the mechatronic design of Thymio was to keep a low production cost. 
-Price is a primary factor for the adoption of the robot by schools.
+Price is a primary factor for the adoption of the robot by schools~\cite{kradolfer2014sociological}.
 In our previous experience with designing robots, most of the costs was due to the electronics and the sensors~\cite{thesis_michael}, not to the mechanical parts.

### 2016-03-27 paper_ram.tex
details

@@
 601074 (survey on open hardware).
-Many thank to Didier Roy at INRIA, Gordana Gerber for the educational material and all the Mobsya team.}% <-this % stops a space``anyone''
+Many thank to Luc Bergeron and his team at écal.ch for the industrial design of Thymio; Didier Roy, David Sherman and all the INRIA team for their contributions and diffusion in France; Gordana Gerber for the educational material; all the Mobsya team for the effort in production and sale.}% <-this % stops a space``anyone''
 \thanks{$^{1}$Francesco Mondada and St\'ephane Magnenat are with the Laboratoire de Syst\`emes Robotiques,
@@
 
-While we optimized Thymio for cost, we also targeted a large range of users.
+While we optimized Thymio for cost, we also targeted a large range of users thanks to an important contribution by industrial designers of the University of Art and Design of Lausanne\footnote{\url{http://www.ecal.ch}}.
 The variety of sensors and ways of interacting with the robot make Thymio accessibly to girls and boys of different age groups.
@@
 High-school and university students can use the robot to learn advanced robotics concepts through the integration with \textsc{ros}~\cite{quigley2009ros}.
+Teachers can use this flexibility of use to adapt the look and the behavior of Thymio to their teaching activities, or even introduce this adaptation as activity in handicraft or computer science lessons.
 Finally, because of the open-source and open-hardware nature of the project, advanced programmers can improve their skills in C by modifying the firmware or in electronics by disassembling the robot.
@@
 In our project we have two different types of hardware: the robot itself and the accessories used in specific activities.
+The robot is the expensive part and has a very neutral design, allowing adaptation to specific situations.
+This adaptation is achieved by custom accessories that increase the attraction of the robot in its specific role, enabling activity for different ages and genders.
+
 For the robot itself we opted for the second interpretation of the definition, considering under ``anyone'' only the professional structures able to produce hardware in a professional way, for a reason of price and complexity of the product.
@@
 This fits with a political trend toward open source in education and in science in general.
-We also found some solutions to the struggling question of production methods by splitting our hardware in two categories, enabling both very advanced hardware and inclusion of a broad community of end users in education.
+We also found a solution to the struggling question of production methods by splitting our hardware in two categories, enabling both very advanced hardware and inclusion of a broad community of end users.
+This addresses durability of equipment and gender issues that are critical in education.
 

### 2016-03-27 paper_ram.tex
minor changes

@@
 It's main weakness is its high price ($\sim$870\,\$).
-Finch\footnote{http://www.finchrobot.com/} ($\sim$99\,\$) is a very simple robot that has been designed around a wired connection to the computer. 
+Finch\footnote{\url{http://www.finchrobot.com/}} ($\sim$99\,\$) is a very simple robot that has been designed around a wired connection to the computer. 
 This connection reduces electronics requirements, such as batteries or wireless communication, and allows to implement the control entirely on the computer.
@@
 On the other hand the cable does not allow real autonomy and mobility of the robot.
-Finally the mBot\footnote{http://www.makeblock.cc/mbot/} is a mobile platform based on the Arduino board. 
+Finally the mBot\footnote{\url{http://www.makeblock.cc/mbot/}} is a mobile platform based on the Arduino board. 
 The very cheap and simple electronics allow to drastically reduce the price of the system ($\sim$75\,\$) but also reduces the perception possibilities to a couple of sensors.
@@
 While we optimized Thymio for cost, we also targeted a large range of users thanks to an important contribution by industrial designers of the University of Art and Design of Lausanne\footnote{\url{http://www.ecal.ch}}.
-The variety of sensors and ways of interacting with the robot make Thymio accessibly to girls and boys of different age groups.
-The white hull is initially neutral and children can choose their own color using the \textsc{rgb} LEDs.
+The variety of sensors, the multiple ways of interacting with the robot, its hull design, and its customization make Thymio accessible to girls and boys of different age groups.
+The white hull is initially neutral and children can choose their own color using the \textsc{rgb} LEDs or attach accessories to the hull.
 The robot can be used from children as young as 6 years old to adults:
@@
 For example, they can create kinetic sculptures.
-For children above 9 years old and adults, Thymio is suited to introduce programming.
+For children above 9 years old and adults, Thymio is suited to introduce programming and robotics.
 Learners can start by programming new interactive behaviors and can go on with constructing around the robot and programming a behavior that interplays with the construction.
@@
 Teachers can use this flexibility of use to adapt the look and the behavior of Thymio to their teaching activities, or even introduce this adaptation as activity in handicraft or computer science lessons.
+Moreover users looking for additional hardware features can interface Thymio with other open hardware devices, such as computer boards\footnote{Interface with a Raspberry PI under \url{https://www.thymio.org/en:thymioexplorer}} or other robotic mechanics\footnote{Interaction with the Poppy open source hardware under \url{https://youtu.be/0otXtF8J_Z4}}.
 Finally, because of the open-source and open-hardware nature of the project, advanced programmers can improve their skills in C by modifying the firmware or in electronics by disassembling the robot.
@@
 
-At the hardware level, we render visible the activity of the sensors by adding a LED next to each, for a total of 39 LEDs.
+At the hardware level, we render visible the activity of the sensors by adding a LED next to each of them, for a total of 39 LEDs.
 These LEDs locally color the hull and allow the user to see immediately where and when the robot perceives a change in its environment.
@@
 
-At the software level, our programming environment (see section \ref{sec:aseba}) provides a set of tools enabling beginners to discover progressively the basic rules of programming using a graphical interface, then the construction of a syntax by graphic blocs, and finally a full coding environment with advanced debugging tools, such as real-time inspection of the variables of the robot and plotting features, providing a visual way to understand time-related concepts.
+At the software level, our programming environment (see section \ref{sec:aseba}) provides a set of tools enabling beginners to discover progressively the basic rules of programming using a graphical interface, then the construction of a syntax by graphic blocs, and finally a full text-based coding environment with advanced debugging tools, such as real-time inspection of the variables of the robot and plotting features, providing a visual way to understand time-related concepts.
 
@@
 Finally one can use paper to create environments, either flat with patterns that can be used in association to the ground sensors (see Figure~\ref{fig:example-construction} bottom) or 3D object, like the trees beside the zombie in Figure~\ref{fig:example-construction}.
-Finally, it is also possible to link several Thymio by software, allowing the creation of complex robotic structures.
+Finally, it is also possible to link several Thymio by software, allowing the coordination of complex multi-Thymio robotic structures.
 
@@
 For each behavior, the robot's body has a different color, allowing the user to easily recognize it.
-The user can navigate between the colors with the buttons and select the behavior she wants to use.
-Basic behaviors can be exploited in a construction to create reactivity without the use of programming (see Figure~\ref{fig:example-construction}, top).
+The user can navigate between the colors with the buttons and select the behavior she/he wants to use.
+Basic behaviors can be exploited in a construction to create reactivity without the use of programming (see Figure~\ref{fig:example-construction}, the top four paper creations can be used with basic behaviors).
 
@@
 When asked about their motivation to participate to open source hardware projects, the people participating to the survey give as main motivation a link to their professional activity and to the specific project, followed by a more general motivation by the nature of the project and finally by the engineering practice.
-A more abstract goal of improving our society is placed at the last position, as mentioned only by few respondents.
+The more abstract goal of improving our society is mentioned only by few respondents.
 
@@
 
-In the Thymio project, most contributors were hired for a research project developing the robot. 
+In the Thymio project, most engineering contributors were hired for a research project developing the robot. 
+Industrial designers also contributed as part of an institutional project.
 Therefore the link with the professional activity and the particular project is evident.
-On another side, the motivation of contributing to society is much stronger in our team, as developing a robot targeting education is a societal goal also.
+On another side, the motivation of contributing to society is much stronger in our team, as developing a robot targeting education has a strong societal component.
 
@@
 Most of the answers mention impact, networking and learning. 
-In these answers we can observe both a technological component in getting better results and impact, and a relational component that is not in the basic definition of open hardware but results from the community created around the project.
+In these answers we can observe both a technological component in getting better results and impact, and a human-relations component that is not in the basic definition of open hardware but results from the community created around the project.
 People expect to work with other people to achieve more together, to create a network of competences or users, to learn through new experience and contacts.
@@
 The wiki is open for editing by anyone, and although we initially provided most of the material, the other members of the community have started to contribute.
+One of the most important contribution to the wiki, in term of effort, has been its translation into four languages.
 

### 2016-03-29 paper_ram.tex
small modifs

@@
 There is another licensing issue which is not well known and that we discovered very late in our project: the constrains of the license of the mechanical and electronic CAD tools. 
-Indeed, asked about this issue, people in the field seems generally not aware of the fact that CAD license can be very restrictive in the way the source files can be published.
+Indeed, asked about this issue, the participants to our survey seem generally not aware of the fact that CAD licenses can be very restrictive about the way the source files can be published.
 Around the half or respondents to our survey, also among the project leaders, state that they were not aware that this could be an issue, and only one third checked the license of their CAD software. 
@@
 To clarify this issue we contacted twelve of the major editors of mechanical CAD and PCB routing software.
-We asked them if their educational license allows the publication of the source files, and we specified that the published files could have been downloaded by a company to produce the system.
-Figure~\ref{fig:editors} summarize the result of this survey.
-Only three of the twelve editors have licenses allowing this type of publication. 
+We asked them if their educational license allows the publication of the source files, and we specified that the published files could have been downloaded by a company to produce the system for commercial purposes.
+Figure~\ref{fig:editors} summarizes the result of this survey.
+Only three of the twelve editors have education licenses allowing this type of publication. 
 Two others explicitly mentioned the possibility to ask for permission before publication.
@@
 This blocking factor for open hardware also apply to the publication of scientific results, as promoted by many governments in the last decade and generally called ``open science''.
-The change of policy in CAD files publication that we reached is a sign of the very positive trend set by open hardware and by open science in general.
+The change of policy in CAD files publication that we obtained is a sign of the very positive trend set by open hardware and by open science in general.
 
@@
 \includegraphics[width=\columnwidth]{figures/table}
-\caption{Publication possibilities as function of the CAD editors.}
+\caption{Publication possibilities as function of the CAD editors, situation in February 2016.}
 \label{fig:editors}
@@
 
-As a conclusion, this issue is totally underestimated by both people participating to the projects and by the CAD editors. 
-In a period where editors are looking for additional revenues and are attacking universities for abuse of licenses\footnote{We are aware, just in Switzerland, of two situations where large amounts of money are asked to universities for non-respect of software licenses.}, this can be a very dangerous situation.
+As a conclusion, this legal issue is totally underestimated by both people participating to the projects and by the CAD editors. 
+In a period where editors are looking for additional revenues and are attacking universities for misuse of licenses\footnote{We are aware, just in Switzerland, of two situations where large amounts of money are asked to universities for non-respect of software licenses.}, this can be a very dangerous situation.
 
@@
 
-Among the fundamental choices when starting an open source hardware project, there is the type of production. 
-In the definition of open hardware we gave, it is stated that we should offer an ``hardware whose design is made publicly available so that anyone can ... make... the ... hardware ''.
+Among the fundamental choices when starting an open source hardware project, there is the choice of the type of production. 
+In the definition of open hardware we gave, it is stated that one should offer an ``hardware whose design is made publicly available so that anyone can ... make... the ... hardware ''.
 Behind the ``anyone'' should we consider every single person or only companies able to produce the product?

### 2016-03-29 paper_ram.tex
minor changes

@@
 
-For the robot itself we opted for the second interpretation of the definition, considering under ``anyone'' only the professional structures able to produce hardware in a professional way, for a reason of price and complexity of the product.
+For the robot itself we opted for the second interpretation of the definition, considering under ``anyone'' only the professional structures able to mass-produce hardware, for a reason of price and complexity of the product.
 When looking for the best performance per price ratio, we decided to use techniques that require heavy equipment.
@@
 This choice has another implication: only a core team of highly skilled engineers can contribute to the project. 
-It is not trivial to design a mechanical part that can be molded win the simplest possible way.
+It is not trivial to design a mechanical part that can be molded in the simplest possible way.
 
@@
 Another key element in supporting repair by the users are the calibration methods. 
-When choosing very low cost components, we get into large dispersion of characteristics. 
-In our case, for instance, the motors can have very different characteristics, resulting in the robot not going straight when asked to apply the same speed on both wheels. 
-To correct this, one interesting solution is to introduce factory calibration, storing correction factors that are specific for each robot, in the microcontroller memory . 
+When choosing very low cost components, one gets into large dispersion of the characteristics. 
+In our case, for instance, the right and left wheel motors can have very different electrical characteristics, resulting in different speed evaluations and therefore in the robot not going straight when the two motors regulators have the same set speed.
+To correct this problem, one interesting solution is to introduce factory calibration, storing correction factors that are specific for each robot, in the microcontroller memory . 
 To allow the user to replace a broken motor, it is essential to give him also the possibility to re-calibrate the robot and adjust the parameters of the new motors.
@@
 
-\subsection{Validated added value of open source hardware}
+\subsection{Added value of open source hardware in distribution}
 
@@
 In particular INRIA in France and the University of Cambridge in he United Kingdoms decided to use Thymio as tool for promotion of their activities and introduced it into schools.
-In France, INRIA developed several teaching units and supported the training of more than 1000 teachers to the use of this particular robot. 
-This finally resulted in having the Thymio robot included as one of the reference tools for the education of digital science in the reference books of the French Academy of Science and its foundation for promotion of science in education, ``La main \`a la p\^ate.''
+In France, INRIA developed several teaching modules and supported the training of more than 1000 teachers to the use of this particular robot. 
+This finally resulted in having the Thymio robot included as one of the tools chosen for the education of digital science in the reference books of the French Academy of Science and its foundation for promotion of science in education, called ``La main \`a la p\^ate.''
 The fact that Thymio is open source allows these institution to have a better control on the tool they use, as they can ensure maintenance and even further development of it. 
+The impulsion given by the academic institutions has been followed by many teachers who contributed with own teaching material. 
+Up to date one can find more than 50 teaching modules, one large tutorial, and many reports on school activities using Thymio.
 
@@
 Thymio was therefore extremely welcome, and the state of Geneva organized own training courses for teachers, ported several educational materials into their curriculum and organized a lending system for all schools that cannot afford to buy the robots.
+Others states are following this model.
 
@@
 
-\subsection{Problems of open-source project}
+\subsection{Problems of open-source projects}
 
@@
 The very specific structure of an open project has also drawbacks. 
-Asked about problems met in the project, the participants to our survey mention the management of the community, the evolution of the project, the efforts in documentation, production and distribution and the limited resources.
+Asked about problems met in their project, the participants to our survey mention the management of the community, the evolution of the project, the efforts in documentation, production and distribution and the limited resources.
 There are here two kind of problems: one of management of people, the other in the development of the project after the first design.
-The management of the project seems a recurrent problem and is related to the nature of this type of structure, mainly built by academic people with engineering background. 
-Most of the projects, following our survey, are managed by the founder of the initiative and not by the most expert in the field, which is perhaps not the best objective choice.
-It is very difficult to evolve from a phase where the engineering is the core of the project, to a new phase where communication and management play an increasing role.
-The same holds for several other factors, like the efforts in documenting, producing and distributing the product. 
-Also here there is a change of activity from the very appealing design phase, to a phase of documentation and, in some cases, production and distribution which is not at all the same work and requires different skills.
-This requires extra efforts and motivation, which are often missing, especially for jobs like documentation that are not appreciated by engineers.
-In general many respondents to our survey identified the development of the project as a general issue, which summarizes the several aspects mentioned above.
+The management of the project seems a recurrent problem and is related to the nature of this type of structure, mainly very flat and built on academic people with engineering background. 
+Most of the projects, following our survey, are managed by the founder of the initiative and not by the most expert in the field or by a manager, which is perhaps not the best objective choice.
+It is indeed very difficult to evolve from a phase where the engineering is the core of the project, to a new phase where communication and management play an increasing role.
+The same holds for several other factors, like the efforts in documentation, production and distribution of the product. 
+Also here there is a change of activity from the very creative and appealing design phase, to a phase of documentation and, in some cases, production and distribution which is not at all the same type of activity and requires different skills.
+This shift requires extra efforts and motivation, which are often missing, especially for activities like documentation that are not appreciated by engineers.
+In general many respondents to our survey identified the evolution of the project as a general problematic issue, which summarizes the several aspects mentioned above.
 

### 2016-03-31 paper_ram.tex
improved version

@@
 This results in 90\% of publications about validation of educational results made on LEGO Mindstorm~\cite{benitti2012exploring}.
-The LEGO Mindstorm EV3\footnote{\url{http://mindstorms.lego.com}} is expensive ($\sim$400\,\$) but offers a wide range of possibilities, especially at the mechanical level using LEGO bricks and at the software level with its graphical programming environment.
-At a lower price, LEGO proposes the WeDo system ($\sim$80\,\$).
+The LEGO Mindstorm EV3\footnote{\url{http://mindstorms.lego.com}} is expensive ($\approx$ 400\,\$) but offers a wide range of possibilities, especially at the mechanical level using LEGO bricks and at the software level with its graphical programming environment.
+At a lower price, LEGO proposes the WeDo system ($\approx$ 80\,\$).
 The WeDo is similar to EV3 but only reads one sensor and controls one actuator and must always be connected to a desktop computer.
-The BeeBot robot\footnote{\url{http://www.beebot.org.uk}} is a cheap ($\sim$90\,\$) autonomous system.
+The BeeBot robot\footnote{\url{http://www.beebot.org.uk}} is a cheap ($\approx$ 90\,\$) autonomous system.
 However, it is not a real robot but rather an automaton: There is no perception of the environment, just a good encoder for precise movements and a simple programming interface based on buttons giving directions for the movements.
@@
 Otherwise, only few educational robots are both used in schools and are issued from an open hardware project:
-Scribbler2, produced and sold by Parallax\footnote{\url{http://www.parallax.com}} ($\sim$180\,\$), is a 188~mm large robot, designed to run on the ground and equipped with few light sensors, one distance sensor, two ground sensors and few LEDs display.
+Scribbler2, produced and sold by Parallax\footnote{\url{http://www.parallax.com}} ($\approx$ 180\,\$), is a 188~mm large robot, designed to run on the ground and equipped with few light sensors, one distance sensor, two ground sensors and few LEDs display.
 It runs on standard AA batteries and has a hacker port for interfacing electronic extensions.
@@
 Several simulators allow to run experiments of high complexity. 
-It's main weakness is its high price ($\sim$870\,\$).
-Finch\footnote{\url{http://www.finchrobot.com/}} ($\sim$99\,\$) is a very simple robot that has been designed around a wired connection to the computer. 
+It's main weakness is its high price ($\approx$ 870\,\$).
+Finch\footnote{\url{http://www.finchrobot.com/}} ($\approx$ 99\,\$) is a very simple robot that has been designed around a wired connection to the computer. 
 This connection reduces electronics requirements, such as batteries or wireless communication, and allows to implement the control entirely on the computer.
@@
 Finally the mBot\footnote{\url{http://www.makeblock.cc/mbot/}} is a mobile platform based on the Arduino board. 
-The very cheap and simple electronics allow to drastically reduce the price of the system ($\sim$75\,\$) but also reduces the perception possibilities to a couple of sensors.
+The very cheap and simple electronics allow to drastically reduce the price of the system ($\approx$ 75\,\$) but also reduces the perception possibilities to a couple of sensors.
 
@@
 Moreover, when selecting features we took into consideration the complexity of the injection mold.
+
+\begin{table}
+\centering
+\begin{tabularx}{\columnwidth}{ll}
+\toprule
+Description & Price (USD)\\
+\midrule
+electronic components & 15\,\$ \\
+microcontroller & 4 \$ \\
+motors & 2 \$ \\
+plastic parts (hull, wheels, and light guides) & 4 \$ \\
+assembling & 8 \$ \\
+transport & 2 \$ \\
+\bottomrule
+Total & 35 \$ \\
+\end{tabularx}
+\caption{Production cost of Thymio.}
+\label{tbl:thymio-price}
+\end{table}
+
+Thymio finally costs 35\,\$ per unit when produced in a batch of 1\,k units (see \ref{tbl:thymio-price}), with fixed fees of 15\,k\,\$ (13\,k\,\$ for the mold and 2\,k\,\$ for the \textsc{ce} certification).
+This allows a selling price of 129\,CHF ($\approx$ 130\,\$).
 
@@
 In this section we analyze more in detail the implications of this choice in the context of educational robotics.
-We compare these implications with the result of a survey we did among 25 people active in various open source hardware projects based on a worldwide call for contributions. 
-Among these 25 people, 6 are project leaders, 10 are part of the core design team, 6 are contributors and 3 are enthusiastic users.
-60\% of these respondents are between 25 and 35 years old and 80\% either work in an academic environment or are students.
+We compare these implications with the result of a survey that got 35 answers from people active in various open source hardware projects based on a worldwide call for contributions. 
+Among these 35 answers, 11 come from project leaders, 13 from core design team members, 8 from contributors and 3 from enthusiastic users.
+54\% of these respondents are between 25 and 35 years old, 26\% between 35 and 50 years old, 9\% are less than 25 years old and 11\% are older than 50.  
+67\% either work in an academic environment or are students. 
+34 of the 35 answers come from male repondents.
+This last point characterizes strongly the hardware developers community.
 
@@
 
-When asked about their motivation to participate to open source hardware projects, the people participating to the survey give as main motivation a link to their professional activity and to the specific project, followed by a more general motivation by the nature of the project and finally by the engineering practice.
+When asked about their motivation to participate to open source hardware projects, the people participating to the survey give as main motivation a link to their professional activity and to the specific project, followed by a more general motivation by the nature of the project and finally by the engineering practice (see Figure~\ref{fig:motivation}).
 The more abstract goal of improving our society is mentioned only by few respondents.
@@
 \centering
-\includegraphics[width=.7\columnwidth]{figures/motivation}
+\includegraphics[width=\columnwidth]{figures/motivation}
 \caption{Motivation of the people participating to the survey.}
@@
 
-It is more interesting to look at what people expect as benefits in participating in such a project.
+It is more interesting to look at what people expect as benefits in participating in such a project (see Figure~\ref{fig:getout}).
 Most of the answers mention impact, networking and learning. 
@@
 \centering
-\includegraphics[width=.7\columnwidth]{figures/getout}
-\caption{Motivation of the people participating to the survey.}
+\includegraphics[width=\columnwidth]{figures/advantages}
+\caption{Advantages that people participating to the survey perceive by contributing to an open source hardware project.}
 \label{fig:getout}
@@
 There is another licensing issue which is not well known and that we discovered very late in our project: the constrains of the license of the mechanical and electronic CAD tools. 
-Indeed, asked about this issue, the participants to our survey seem generally not aware of the fact that CAD licenses can be very restrictive about the way the source files can be published.
-Around the half or respondents to our survey, also among the project leaders, state that they were not aware that this could be an issue, and only one third checked the license of their CAD software. 
-Among the respondents, some stated that they did not checked the license because they were sure they can freely publish their design, but half of them are using software not allowing the publication of source files when using educational licenses, and all these respondents are academic or students.
+Indeed, asked about this issue, the participants to our survey seem generally not aware of the fact that CAD licenses can be very restrictive about the way the source files can be published (see Figure~\ref{fig:aware}).
+More than the half or respondents to our survey, and more than 60\% among the project leaders, state that they were not aware that this could be an issue, and only one third checked the license of their CAD software.
+Among the respondents, some stated that they did not checked the license because they were sure they can freely publish their design, but one third of them are using software not allowing the publication of source files when using educational licenses, and all these respondents are academic or students.
 This issue is very serious, as most contributors of open source hardware projects are academic and use academic or educational licenses. 
@@
 \centering
-\includegraphics[width=.9\columnwidth]{figures/aware-license}
+\includegraphics[width=\columnwidth]{figures/checklicense}
 \caption{Awarness about CAD tool licenses limitations by people participating to the survey.}
@@
 \includegraphics[width=\columnwidth]{figures/table}
-\caption{Publication possibilities as function of the CAD editors, situation in February 2016.}
+\caption{Publication possibilities as function of the CAD editors, situation end of March 2016.}
 \label{fig:editors}
@@
 \centering
-\includegraphics[width=\columnwidth]{figures/production}
+\includegraphics[width=\columnwidth]{figures/manufacturing}
 \caption{Production methods in the projects mentioned by the people participating to the survey.}
@@
 \centering
-\includegraphics[width=\columnwidth]{figures/problems}
-\caption{Problems observed by participants to open source hardware projects.}
+\includegraphics[width=\columnwidth]{figures/limitations}
+\caption{Limitations of open source hardware projects observed by participants to open source hardware projects.}
 \label{fig:problems}
@@
 The very specific structure of an open project has also drawbacks. 
-Asked about problems met in their project, the participants to our survey mention the management of the community, the evolution of the project, the efforts in documentation, production and distribution and the limited resources.
+Asked about problems met in their project (see Figure~\ref{fig:problems}), the participants to our survey mention the management of the community, the evolution of the project, the efforts in documentation, production and distribution and the limited resources.
 There are here two kind of problems: one of management of people, the other in the development of the project after the first design.
@@
 
-XXXXXXX historical, should we keep this?
-
-\begin{table}
-\centering
-\begin{tabularx}{\columnwidth}{ll}
-\toprule
-Description & Price (USD)\\
-\midrule
-electronic components & 15\,\$ \\
-microcontroller & 4 \$ \\
-motors & 2 \$ \\
-plastic parts (hull, wheels, and light guides) & 4 \$ \\
-assembling & 8 \$ \\
-transport & 2 \$ \\
-\bottomrule
-Total & 35 \$ \\
-\end{tabularx}
-\caption{Production cost of Thymio.}
-\label{tbl:thymio-price}
-\end{table}
-
-We reached our price target thanks to our careful selection of components and our production in China.
-Thymio costs 35\,\$ per unit to produce in batch of 1\,k units (see \ref{tbl:thymio-price}), with fixed fees of 15\,k\,\$ (13\,k\,\$ for the mold and 2\,k\,\$ for the \textsc{ce} certification).
-This allows a selling price of 100\,CHF ($\approx$ 106\,\$).
-This small ratio of selling-over-production price is made possible by our non-profit business model.
-
-XXXXXXX end of historical
-
 \section{Conclusion}
@@
 The open source hardware and the related contributions are split in two categories: the robot itself, produced by a non-profit organization and requiring very advanced skills in both design and production, and the accessories for activities, based on techniques broadly accessible and with a larger panel of contributors.
+
+The open source hardware development and dissemination of Thymio addresses several of the issues found in educational robotics.
+It allowed to broadly distributed the robot with minimal changes dues to management of intellectual properties, royalties, financial support and so on.
+This was achieved with an excellent match between the philosophy of the project and the one of the community of users in education.
+In particular, the open source approach allow to provide a durable robot, easy to maintain and repair, with at the same time a community of users providing educational material and mutual support.
 

### 2016-03-31 paper_ram.tex
modifs Michael

@@
    \centering
-	\includegraphics[width=.6\columnwidth]{figures/front_2016}
+	\includegraphics[width=.8\columnwidth]{figures/front_2016}
 	\vskip 1.5em
-	\includegraphics[width=\columnwidth]{figures/Thymio_II_sensors-actuators-eng}
-	\caption{The Thymio robot and its main components.}
+	\includegraphics[width=\columnwidth]{figures/wirelessThymioVecto}
+	\caption{The Thymio robot and its main components for the Wireless and the plain version.}
 	\label{fig:schematique}
@@
 \end{tabularx}
-\caption{Production cost of Thymio.}
+\caption{Production cost of basic Thymio (without wireless modules).}
 \label{tbl:thymio-price}
@@
 This is a barrier for many users and we wanted to avoid it carefully, having a robot able to show interesting behaviors right out of the box.
-Therefore Thymio provides 6 different basic behaviors (see table \ref{tbl:basic-behaviors}), which are accessible as soon as the robot is started.
-These allow people encountering Thymio for the first time to immediately interact with it, while providing insights into the many possibilities of the robot.
+Therefore Thymio provides 6 different basic behaviors (see table \ref{tbl:basic-behaviors}), which are accessible as soon as the robot is started and are stored in flash permanently, also if the Thymio is reprogrammed by the user.
+These basic behaviors allow people starting Thymio to immediately interact with it, while illustrating the many possibilities of the robot.
 The behaviors are interactive and oriented towards the user. % SM: FIXME: not clear what "oriented towards the user" means
@@
 The management of the project seems a recurrent problem and is related to the nature of this type of structure, mainly very flat and built on academic people with engineering background. 
-Most of the projects, following our survey, are managed by the founder of the initiative and not by the most expert in the field or by a manager, which is perhaps not the best objective choice.
+More than 62\% of the answers to our survey report about projects that are managed by the founder of the initiative and not by the most expert in the field or by a manager, which is perhaps not the best objective choice.
 It is indeed very difficult to evolve from a phase where the engineering is the core of the project, to a new phase where communication and management play an increasing role.

### 2016-03-31 paper_ram.tex
changes pour commentaires de steph

@@
 \subsection{Multi-age and Gender-neutral Feature Set}
+\label{sec:multi}
 
@@
 
-\subsection{Short time to demo}
+\subsection{Fast access to robotics behaviors}
 \label{sec:behaviors}
@@
 A virtual machine allows instantaneous upload and safe execution of code.
-On the desktop side, Aseba provides an IDE which allows to program in a purely graphical mode, or using a scripting language (see Figure~\ref{fig:aseba-studio}) or in a mixed way, assembling code in a graphical way using the blockly environment\footnote{\url{https://developers.google.com/blockly/}}.
+On the desktop side, Aseba provides an IDE which allows to program in a purely graphical mode (called VPL), or using a scripting language (see Figure~\ref{fig:aseba-studio}) or in a mixed way, assembling code in a graphical way using the blockly environment\footnote{\url{https://developers.google.com/blockly/}}.
+This allows the IDE to match with the multi-age goal described in section~\ref{sec:multi}.
 
@@
 
-The IDE integrates a help documentation\footnote{currently English, French, and German}.
+The IDE integrates real-time feedback both graphically under VPL~\cite{Magnenat2015} and displaying variables or graphs in the text mode.
+The IDE also provides a help documentation\footnote{currently English, French, and German}.
 In addition, the wiki pages of the project provide tutorials on programming with Thymio and a detailed description of the programming interface.

### 2016-03-31 paper_ram.tex
modifs

@@
 \title{\LARGE \bf
-Thymio, an example of open source hardware robot \\
-for formal education
+Bringing robotics in formal education \\
+using the Thymio open source hardware robot
 }
@@
 Many thank to Luc Bergeron and his team at écal.ch for the industrial design of Thymio; Didier Roy, David Sherman and all the INRIA team for their contributions and diffusion in France; Gordana Gerber for the educational material; all the Mobsya team for the effort in production and sale.}% <-this % stops a space``anyone''
-\thanks{$^{1}$Francesco Mondada and St\'ephane Magnenat are with the Laboratoire de Syst\`emes Robotiques,
+\thanks{$^{1}$Francesco Mondada, Manon Briod, L\'ea Pereyre, Philippe R\'etornaz and St\'ephane Magnenat are with the Laboratoire de Syst\`emes Robotiques,
         Ecole Polytechnique F\'ed\'erale de Lausanne (EPFL), Switzerland
-        {\tt\small francesco.mondada@epfl.ch} and {\tt\small stephane.magnenat@epfl.ch}}%
+        {\tt\small firstname.lastname@epfl.ch}}%
 \thanks{$^{2}$Michael Bonani and Fanny Riedo are with Mobsya Association, Ecublens, Switzerland
@@
 In Europe, despite a trend in including more and more technological tools into the learning process, teachers are still insufficiently trained and are reluctant to introduce these tools in their teaching activity~\cite{CERI2008}.
-To be accepted by teachers, robots must be therefore both accessible with minimal effort and be associated with well prepared educational material. 
+To be accepted by teachers, robots must be therefore both accessible with minimal effort and be associated with-well prepared educational material. 
 Moreover an infrastructure should allow the exchange of educational material among the teachers.
@@
 
-For more complex robots, both the development of educational material and the training of teachers requires a very important effort, requiring a good mix between robotics and educational skills. 
+For more complex robots, both the development of educational material and the training of teachers require a very important effort, requiring a good mix between robotics and educational skills. 
 Moreover, educational material changes very much from school to school, as education is very dependent from local educational programs and languages, for instance. 
 In the particular case of Switzerland, for instance, every of the 26 states composing the confederation has its own school program and its own language among the four official ones. 
-Educational material and the related training has therefore to be developed for each situation, which requires a huge effort.
+Educational material and the related training have therefore to be developed for each situation, which requires a huge effort.
 Even for large companies like LEGO this development is not affordable. 
@@
 Learners can start by programming new interactive behaviors and can go on with constructing around the robot and programming a behavior that interplays with the construction.
-High-school and university students can use the robot to learn advanced robotics concepts through the integration with \textsc{ros}~\cite{quigley2009ros}.
+High school and university students can use the robot to learn advanced robotics concepts through the integration with \textsc{ros}~\cite{quigley2009ros}.
 Teachers can use this flexibility of use to adapt the look and the behavior of Thymio to their teaching activities, or even introduce this adaptation as activity in handicraft or computer science lessons.
@@
 There are four attachment positions on the top of the robot and two LEGO Technics crosses on each side.
-Additionally, the two wheels have attach points, which permits to use them to actuate elements elsewhere in the structure (see Figure~\ref{fig:example-construction}, third row left) or to lift the robot's own weight (see Figure~\ref{fig:example-construction}, third row right).
+Additionally, the two wheels have attachment points, which permits to use them to actuate elements elsewhere in the structure (see Figure~\ref{fig:example-construction}, third row left) or to lift the robot's own weight (see Figure~\ref{fig:example-construction}, third row right).
 Therefore we chose more powerful motors than strictly necessary to move the robot around.
@@
 A virtual machine allows instantaneous upload and safe execution of code.
-On the desktop side, Aseba provides an IDE which allows to program in a purely graphical mode (called VPL), or using a scripting language (see Figure~\ref{fig:aseba-studio}) or in a mixed way, assembling code in a graphical way using the blockly environment\footnote{\url{https://developers.google.com/blockly/}}.
+On the desktop side, Aseba provides an IDE which allows to program in a purely graphical mode (called VPL, see Figure~\ref{fig:vpl}), or using a scripting language (see Figure~\ref{fig:aseba-studio}) or in a mixed way, assembling code in a graphical way using the blockly environment\footnote{\url{https://developers.google.com/blockly/}} (see Figure~\ref{fig:blockly}).
 This allows the IDE to match with the multi-age goal described in section~\ref{sec:multi}.
@@
 This makes the robot suitable for university-level education.
+
+\begin{figure}
+\centering
+\includegraphics[width=.9\columnwidth]{figures/vpl}
+\caption{The graphic programming environment VPL.}
+\label{fig:vpl}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=.9\columnwidth]{figures/blockly}
+\caption{The blockly programming environment.}
+\label{fig:blockly}
+\end{figure}
 
@@
 On another side, the motivation of contributing to society is much stronger in our team, as developing a robot targeting education has a strong societal component.
+Moreover this project has a strong scientific motivation, with several ongoing studies on the acceptance by teachers and the impact on children.
+Also this motivation is therefore stronger in our team.
+Although this seems a minor issue, sharing a strong fundamental motivation such as education or scientific achievements, is a key element for a strong community~\cite{Stahlbrost2011}, especially if interdisciplinary like ours.
 
-It is more interesting to look at what people expect as benefits in participating in such a project (see Figure~\ref{fig:getout}).
+It is also interesting to look at what people expect as benefits in participating in such a project (see Figure~\ref{fig:getout}).
 Most of the answers mention impact, networking and learning. 
@@
 The wiki is open for editing by anyone, and although we initially provided most of the material, the other members of the community have started to contribute.
-One of the most important contribution to the wiki, in term of effort, has been its translation into four languages.
+One of the most important contributions to the wiki, in term of effort, has been its translation into four languages.
 
@@
 When starting an open source hardware project, one of the typical questions is about which license to use when disseminating the project source files. 
-We will not discuss this matter here, as it is a very common and well covered issue.
+We will not discuss this matter here, as it is a very common and well-covered issue.
 
@@
 \includegraphics[width=\columnwidth]{figures/checklicense}
-\caption{Awarness about CAD tool licenses limitations by people participating to the survey.}
+\caption{Awareness about CAD tool licenses limitations by people participating to the survey.}
 \label{fig:aware}
@@
 This can multiply by a factor of some hundreds the price of the CAD license.
-This blocking factor for open hardware also apply to the publication of scientific results, as promoted by many governments in the last decade and generally called ``open science''.
+This blocking factor for open hardware also applies to the publication of scientific results, as promoted by many governments in the last decade and generally called ``open science''.
 The change of policy in CAD files publication that we obtained is a sign of the very positive trend set by open hardware and by open science in general.
@@
 Schools do not have large budget for technological tools and make strong investment in training of their teachers when adopting a new tool.
-Therefore the life time of the products should be as long as possible. 
+Therefore the lifetime of the products should be as long as possible. 
 The open hardware approach gives to the user, or to a generic technician, better conditions to repair the system.
@@
 In our case, for instance, the right and left wheel motors can have very different electrical characteristics, resulting in different speed evaluations and therefore in the robot not going straight when the two motors regulators have the same set speed.
-To correct this problem, one interesting solution is to introduce factory calibration, storing correction factors that are specific for each robot, in the microcontroller memory . 
+To correct this problem, one interesting solution is to introduce factory calibration, storing correction factors that are specific for each robot, in the microcontroller memory. 
 To allow the user to replace a broken motor, it is essential to give him also the possibility to re-calibrate the robot and adjust the parameters of the new motors.
@@
 Beside the sale of more than 10'000 robots by a non-profit organization producing it, there are several elements that show a specific advantage of having chosen an open source hardware model.
-The first element is the very high acceptance of the robot by universities and their offices in charge of promotion of science.
+
+The first element is linked to the financial support of the project.
+Although the non-profit nature of the project blocked the possibility to have shareholders injecting capital, this same nature enabled a lot of institutional donations linked with the societal goal of the project.
+The financial support through donations instead of acquisition of shares, allows keeping a total decisional freedom and void debts at the same time.
+The main disadvantage of this approach is the amount of financial support, limited in our case of several hundreds of thousands of dollars.
+
+The second interesting element of the open source hardware appraoch is the very high acceptance of the resulting robot by universities and their offices in charge of promotion of science.
 Although these are not the primary user targeted by the robot, these are institutions with high visibility and influence on school programs. 
@@
 In some cases, the open source nature of this project was a crucial political argument for the choice of this platform.
-In the state of Geneva, the government decided to move toward open source non proprietary tools in education.
+In the state of Geneva, for instance, the government decided to move toward open source non-proprietary tools in education.
 All primary school- and most secondary school computers have been installed with the Ubuntu distribution of Linux. 
@@
 Thymio was therefore extremely welcome, and the state of Geneva organized own training courses for teachers, ported several educational materials into their curriculum and organized a lending system for all schools that cannot afford to buy the robots.
-Others states are following this model.
+Others states are starting following this model.
 
-Finally the non-profit nature of the project, including the production infrastructure, has contributed to reduce the obstacle toward the acquisition of this platform for many schools, especially at the beginning of the process.
+Finally the non-profit nature of the project, including the production infrastructure, has contributed to reduce the obstacle toward the acquisition of this platform for many schools, where teachers have a strong societal motivation and are reluctant to use tax-payers money to pay the divident of some shareholders.
 
@@
 
-In the Thymio project we had very similar issues and tried to addess them by having several partners, each of them specialized in one aspect: design, production and sale, creation of educational material, etc.
+In the Thymio project we had very similar issues and tried to address them by having several partners, each of them specialized in one aspect: design, production and sale, creation of educational material, etc.
 Despite this approach, we had also problems of management of the community, and issues in finding the right long term vision, especially having so many partners. 
@@
 The open source hardware development and dissemination of Thymio addresses several of the issues found in educational robotics.
-It allowed to broadly distributed the robot with minimal changes dues to management of intellectual properties, royalties, financial support and so on.
+It allowed to broadly distribute the robot with minimal changes dues to management of intellectual properties, royalties, financial support and so on.
 This was achieved with an excellent match between the philosophy of the project and the one of the community of users in education.
-In particular, the open source approach allow to provide a durable robot, easy to maintain and repair, with at the same time a community of users providing educational material and mutual support.
+In particular, the open source approach allows to provide a durable robot, easy to maintain and repair, with at the same time a community of users providing educational material and mutual support.
 

### 2016-04-01 paper_ram.tex
added discussion with e-puck and khepera, + minor issues

@@
 \begin{enumerate}
-\item Although there are many robotic projects developing educational robots, few reach a sufficient maturity to become distributed and accessible to schools. 
+\item Although there are many robotic projects developing innovative and interesting educational robots, few reach a sufficient maturity to become distributed and accessible to schools. 
 \item A robot performing interesting behaviors is a complex piece of technology and therefore expensive. 
@@
 Among those, only very few are commercially available, limiting their validation by education scientists.
-This results in 90\% of publications about validation of educational results made on LEGO Mindstorm~\cite{benitti2012exploring}.
-The LEGO Mindstorm EV3\footnote{\url{http://mindstorms.lego.com}} is expensive ($\approx$ 400\,\$) but offers a wide range of possibilities, especially at the mechanical level using LEGO bricks and at the software level with its graphical programming environment.
-At a lower price, LEGO proposes the WeDo system ($\approx$ 80\,\$).
-The WeDo is similar to EV3 but only reads one sensor and controls one actuator and must always be connected to a desktop computer.
+This results in 90\% of publications about validation of educational results made on LEGO\textsuperscript{\textregistered} Mindstorms\textsuperscript{\textregistered}~\cite{benitti2012exploring}.
+The LEGO\textsuperscript{\textregistered} Mindstorms\textsuperscript{\textregistered} EV3\footnote{\url{http://mindstorms.lego.com}} is expensive ($\approx$ 400\,\$) but offers a wide range of possibilities, especially at the mechanical level using LEGO\textsuperscript{\textregistered} bricks and at the software level with its graphical programming environment.
+At a lower price, LEGO\textsuperscript{\textregistered} proposes the WeDo\textsuperscript{\textregistered} system ($\approx$ 80\,\$).
+The WeDo\textsuperscript{\textregistered} is similar to EV3 but only reads one sensor and controls one actuator and must always be connected to a desktop computer.
 The BeeBot robot\footnote{\url{http://www.beebot.org.uk}} is a cheap ($\approx$ 90\,\$) autonomous system.
@@
 Among these widely used robots, none is part of an open source hardware project.
-The only effort in this direction comes from LEGO, that made public all the schematics of the EV3 in an hardware developer kit, to allow the design of extensions. 
+The only effort in this direction comes from LEGO\textsuperscript{\textregistered}, that made public all the schematics of the EV3 in an hardware developer kit, to allow the design of extensions. 
 Otherwise, only few educational robots are both used in schools and are issued from an open hardware project:
@@
 The main weakness of Scribbler2 is its performance per price ratio and the limited compatibility with other systems. 
+Moreover there seems to be no real community around its development.
 The e-puck~\cite{mondada2009puck} robot is a robot targeting university-level education.
@@
 Several simulators allow to run experiments of high complexity. 
+There is an active community of users and developers around it.
 It's main weakness is its high price ($\approx$ 870\,\$).
@@
 
+Among the main target users there are teachers.
 For teachers, the motivation to use robotic tools in their classes depends on many factors \cite{chevalier2016}.
@@
 
-For more complex robots, both the development of educational material and the training of teachers require a very important effort, requiring a good mix between robotics and educational skills. 
+For more complex robots, both the development of educational material and the training of teachers require a very important effort, needing a good mix between robotics and educational skills. 
 Moreover, educational material changes very much from school to school, as education is very dependent from local educational programs and languages, for instance. 
-In the particular case of Switzerland, for instance, every of the 26 states composing the confederation has its own school program and its own language among the four official ones. 
-Educational material and the related training have therefore to be developed for each situation, which requires a huge effort.
-Even for large companies like LEGO this development is not affordable. 
+In the particular case of Switzerland, for instance, every of the 26 states composing the confederation has its own school program and its own language among the four national ones. 
+Educational material and the related training have therefore to be adapted to each situation, which requires a huge effort.
+Even for large companies like LEGO\textsuperscript{\textregistered} this development is not affordable. 
 A solution is to have an active community of users contributing to the development of the material in a distributed manner, adapting the material to the local situation.
-LEGO itself is moving in the direction of promoting communities of users \cite{Hienerth2014} with a user-producer interaction close to those of open hardware projects.
+LEGO\textsuperscript{\textregistered} itself is moving in the direction of promoting communities of users \cite{Hienerth2014} with a user-producer interaction close to those of open hardware projects.
 The open source community regrouping developers, producers and final users is therefore a very interesting model to address the distributed development and sharing of educational material, and the diffusion of training sessions.
-Communities around projects like Arduino are extremely good examples of this type of mechanism \cite{Jamieson2010}.
+This model is an extension to education of what already practiced in projects like Arduino~\cite{Jamieson2010}.
 
@@
 We designed the Thymio robot along seven main axes: a low price to address a larger number of users, a feature set that suits both genders and multiple ages from young children to adults, a combination of sensors, actuators, and programming features that facilitates learning, a mechanical design that allows creative extensions by the users themselves, a set of ready-to-use behaviors to get a demonstration right out of the box, an accessible programming environment, and an open source  community contributing to the design and disseminating the results.
-The result is a miniature (10\,cm) differential-wheeled robot suited for use on a desktop.
+The result is a miniature (10\,cm in diameter, see Figure~\ref{fig:schematique}) differential-wheeled robot suited for use on a desktop.
 The robot is robust enough to be mishandled by children, as it can fall from a table without breaking itself.
-It features a translucent white hull and a wide range of sensors and actuators (see \ref{fig:schematique}).
+It features a translucent white hull and a wide range of sensors and actuators (see Figure~\ref{fig:schematique} bottom).
 The robot embeds a battery, rechargable by \textsc{usb}, that provides an autonomy of about 2 hours.
@@
 Our main concern during the mechatronic design of Thymio was to keep a low production cost. 
-Price is a primary factor for the adoption of the robot by schools~\cite{kradolfer2014sociological}.
+Price is a key factor for the adoption of the robot by schools~\cite{kradolfer2014sociological}.
 In our previous experience with designing robots, most of the costs was due to the electronics and the sensors~\cite{thesis_michael}, not to the mechanical parts.
@@
 
-Thymio finally costs 35\,\$ per unit when produced in a batch of 1\,k units (see \ref{tbl:thymio-price}), with fixed fees of 15\,k\,\$ (13\,k\,\$ for the mold and 2\,k\,\$ for the \textsc{ce} certification).
+Thymio finally costs 35\,\$ per unit when produced in a batch of 1\,k units (see Table~\ref{tbl:thymio-price}), with fixed fees of 15\,k\,\$ (13\,k\,\$ for the mold and 2\,k\,\$ for the \textsc{ce} certification).
 This allows a selling price of 129\,CHF ($\approx$ 130\,\$).
@@
 The square format of the hull allows to use the robot as a base for driving the user's own constructions.
-To that end, Thymio is compatible with LEGO bricks:
-There are four attachment positions on the top of the robot and two LEGO Technics crosses on each side.
+To that end, Thymio is compatible with LEGO\textsuperscript{\textregistered} bricks:
+There are four attachment positions on the top of the robot and two LEGO\textsuperscript{\textregistered} Technics crosses on each side.
 Additionally, the two wheels have attachment points, which permits to use them to actuate elements elsewhere in the structure (see Figure~\ref{fig:example-construction}, third row left) or to lift the robot's own weight (see Figure~\ref{fig:example-construction}, third row right).
 Therefore we chose more powerful motors than strictly necessary to move the robot around.
-The same fixation points used for LEGO bricks can be used to attach paper structures.
+The same fixation points used for LEGO\textsuperscript{\textregistered} bricks can be used to attach paper structures.
 Paper can allow to simply change the body shape and add some body movements, as illustrated in Figure~\ref{fig:example-construction} by the orca, opening and closing its mouth while moving forward or by the bat, moving its wings.
@@
 \includegraphics[width=\columnwidth]{figures/5senses}
-\caption{Examples of extensions of the Thymio basic robot with paper or cardboard body extensions (top four images), using LEGO$^{TM}$ structural extensions (third row), using 3D-printed extensions (fourth row) or using a printed environment (bottom).}
+\caption{Examples of extensions of the Thymio basic robot with paper or cardboard body extensions (top four images), using LEGO\textsuperscript{\textregistered} structural extensions (third row), using 3D-printed extensions (fourth row) or using a printed environment (bottom).}
 \label{fig:example-construction}
@@
 67\% either work in an academic environment or are students. 
-34 of the 35 answers come from male repondents.
-This last point characterizes strongly the hardware developers community.
+34 of the 35 answers come from male respondents.
 
@@
 
-When asked about their motivation to participate to open source hardware projects, the people participating to the survey give as main motivation a link to their professional activity and to the specific project, followed by a more general motivation by the nature of the project and finally by the engineering practice (see Figure~\ref{fig:motivation}).
+We can distinguish two levels of motivation, the institutional and the personal one.
+The institution initiating a project is mainly looking for recognition and/or money. 
+Our group has a good experience in disseminating robotic hardware with the Khepera~\cite{MonFraIen93} and e-puck~\cite{mondada2009puck} robots.
+Khepera was disseminated with a proprietary strategy, e-puck with an open source hardware one.
+Both were targeting similar users and have been sold in similar quantities.
+What we can observe after more than 10 years is that Khepera generated many royalties for the university, but less relative academic visibility than e-puck (see Figure~\ref{fig:khepuck}) .
+The e-puck robot, being open hardware and with an image better linked to the university, generated no income for the university but much more visibility in respect to his diffusion.
+Khepera was more linked with the name of the company producing it, the e-puck robot was nearly not linked with the producer when mentioned in the literature.
+In the case of Thymio, developed in a research project, the institutional motivation was to get visibility more than money. 
+Therefore, based on the past experience, the open source hardware strategy seemed more adapted.
+
+\begin{figure}
+\centering
+\includegraphics[width=\columnwidth]{figures/others_robots}
+\caption{Comparison between mentions in the literature, mention of the manufacturer and citation of the reference paper for two robots disseminated by our lab in the past.
+It appears that the Khepera robot, distributed with a proprietary strategy, was more linked to the company producing it than the original designers.
+This is the opposite for the e-puck robot, distributed with an open source license.}
+\label{fig:khepuck}
+\end{figure}
+
+The personal motivation to participate to such a project is very different than the institutional one. 
+When asked about their personal motivation, the people participating to the survey give as main motivation a link to their professional activity and to the specific project, followed by a more general motivation by the nature of the project and finally by the engineering practice (see Figure~\ref{fig:motivation}).
 The more abstract goal of improving our society is mentioned only by few respondents.
@@
 
-For the accessories, less technically challenging and stronger linked with creativity and educational value, we promoted techniques that are accessible to ``anyone'' in the broader sense: paper, cardboard, LEGO constructions and 3D printing, as explained in section~\ref{sec:crea} and illustrated in Figure~\ref{fig:example-construction}.
+For the accessories, less technically challenging and stronger linked with creativity and educational value, we promoted techniques that are accessible to ``anyone'' in the broader sense: paper, cardboard, LEGO\textsuperscript{\textregistered} constructions and 3D printing, as explained in section~\ref{sec:crea} and illustrated in Figure~\ref{fig:example-construction}.
 This allows a much broader spectrum of contributor, especially among the teachers.
@@
 All primary school- and most secondary school computers have been installed with the Ubuntu distribution of Linux. 
-On this operating system the LEGO Mindstorms software tools are not available, and their proprietary nature is not compatible with the philosophy dictated by the government. 
+On this operating system the LEGO\textsuperscript{\textregistered} Mindstorms\textsuperscript{\textregistered} software tools are not available, and their proprietary nature is not compatible with the philosophy dictated by the government.
 Thymio was therefore extremely welcome, and the state of Geneva organized own training courses for teachers, ported several educational materials into their curriculum and organized a lending system for all schools that cannot afford to buy the robots.

### 2016-04-09 paper_ram.tex
shorter version

@@
 In this paper, we show that this concept, implemented as a project in a community of users, developers and producers, brings a very interesting added value to the robot and to the educational methods.
-This is illustrated by the results achieved with Thymio and is compared with the feedback of a group of engineers active in open source hardware design.
-We can well observe both similarities and differences between our education-centered project and other open source hardware projects, pointing to challenges and opportunities.
+This is illustrated by the results achieved with Thymio and is compared with the feedback of a group of robotics engineers active in open source hardware design.
+We can well observe both similarities and differences between our education-centered project and other open source robotic hardware projects, pointing to challenges and opportunities.
 More details on the robot and the related results can be found in~\cite{RiedoPhD, magnenat2014}.
@@
 Most of these sensors cost less then 0.2\,\$, the most expensive being the accelerometer with a cost of about 0.8\,\$, which is an acceptable price given the possibilities it brings to the robot.
-The omnipresence of this type of sensors in smartphones and compact cameras has drastically lowered their prices over the last decade.
+%The omnipresence of this type of sensors in smartphones and compact cameras has drastically lowered their prices over the last decade.
 We also chose low-cost toy motors and control them in speed (max. 13\,cm/s) using a custom-designed electronics.
@@
 This microcontroller controls all sensors and actuators, excepted the battery recharging logic, which uses a specific chip for safety reasons.
-This choice of a single microcontroller led us to non-trivial design decisions like using low-cost shift registers to drive the many LEDs.
-Moreover, instead of directly connecting the infrared sensors to an analogical input, we interface them using a comparator that transforms the analogical signal into a binary one with a pulse having a duration proportional to the intensity of the original signal~\cite{wanda}.
-This design saves \textsc{cpu}-time, reduces power consumption (it allows a shorter infrared pulse than with an analogical design) and allows to use the sensor for communication.
+%This choice of a single microcontroller led us to non-trivial design decisions like using low-cost shift registers to drive the many LEDs.
+%Moreover, instead of directly connecting the infrared sensors to an analogical input, we interface them using a comparator that transforms the analogical signal into a binary one with a pulse having a duration proportional to the intensity of the original signal~\cite{wanda}.
+%This design saves \textsc{cpu}-time, reduces power consumption (it allows a shorter infrared pulse than with an analogical design) and allows to use the sensor for communication.
 
@@
 The robot can be used from children as young as 6 years old to adults:
-The youngest play with the robot through a set of basic behaviors (see section \ref{sec:behaviors}) and the older can program it (see section \ref{sec:aseba}).
-The different basic behaviors are selectable using the capacitive buttons and show some of the capabilities of the robot, such as obstacle avoidance, slope climbing, etc.
-By interacting with the different behaviors, for instance by blocking the robot with the hand and seeing it avoiding it, children can discover the concept of the sensory-motor loop. 
-As every behavior has a different color, children can also understand that different \emph{programs} run on the robot.
-Children around 6 can use Thymio as a support for handicrafts or construction.
-For example, they can create kinetic sculptures.
+The youngest play with the robot through a set of basic behaviors (see section \ref{sec:behaviors}) and can use Thymio as a support for handicrafts or construction.
+%and the older can program it (see section \ref{sec:aseba}).
+%The different basic behaviors are selectable using the capacitive buttons and show some of the capabilities of the robot, such as obstacle avoidance, slope climbing, etc.
+%By interacting with the different behaviors, for instance by blocking the robot with the hand and seeing it avoiding it, children can discover the concept of the sensory-motor loop. 
+%As every behavior has a different color, children can also understand that different \emph{programs} run on the robot.
+%Children around 6 can use Thymio as a support for handicrafts or construction.
+%For example, they can create kinetic sculptures.
 For children above 9 years old and adults, Thymio is suited to introduce programming and robotics.
-Learners can start by programming new interactive behaviors and can go on with constructing around the robot and programming a behavior that interplays with the construction.
+Learners can start by programming new interactive behaviors (see section \ref{sec:aseba}) and can go on with constructing around the robot and programming a behavior that interplays with the construction.
 High school and university students can use the robot to learn advanced robotics concepts through the integration with \textsc{ros}~\cite{quigley2009ros}.

### 2016-04-09 paper_ram.tex
small type

@@
 Price is a key factor for the adoption of the robot by schools~\cite{kradolfer2014sociological}.
-In our previous experience with designing robots, most of the costs was due to the electronics and the sensors~\cite{thesis_michael}, not to the mechanical parts.
+In our previous experience with designing robots, most of the costs were due to the electronics and the sensors~\cite{thesis_michael}, not to the mechanical parts.
 As we wanted Thymio to be an affordable robot, we selected electronic components with great care and explored different design solutions in order to limit the number of expensive components.

### 2016-04-11 paper_ram.tex
Passe de correction

@@
 \usepackage{booktabs}
+\usepackage{microtype}
+\usepackage{acronym}
+
+\acrodef{ide}[\textsc{ide}]{integrated development environment}
+\acrodef{vpl}[\textsc{vpl}]{visual programming language}
 
 \title{\LARGE \bf
-Bringing robotics in formal education \\
+Bringing robotics into formal education \\
 using the Thymio open source hardware robot
@@
 601074 (survey on open hardware).
-Many thank to Luc Bergeron and his team at écal.ch for the industrial design of Thymio; Didier Roy, David Sherman and all the INRIA team for their contributions and diffusion in France; Gordana Gerber for the educational material; all the Mobsya team for the effort in production and sale.}% <-this % stops a space``anyone''
+Many thank to Luc Bergeron and his team at écal.ch for the industrial design of Thymio; Didier Roy, David Sherman and all the INRIA team for their contributions and diffusion in France; Gordana Gerber for the educational material; all the Mobsya team for the effort in production and sale; and Fabian Hahn for the port of Blockly.}% <-this % stops a space``anyone''
 \thanks{$^{1}$Francesco Mondada, Manon Briod, L\'ea Pereyre, Philippe R\'etornaz and St\'ephane Magnenat are with the Laboratoire de Syst\`emes Robotiques,
@@
 %Mobile robots are interesting tools for education because of both the enthusiasm they raise and the multidisciplinary nature of their technology.
-%However, because of their price, complexity, bias towards specific groups (boys), and unproved educational value, they are still seldom used in schools and edutainment.
+%However, because of their price, complexity, bias toward specific groups (boys), and unproved educational value, they are still seldom used in schools and edutainment.
 %This paper presents a system-design approach and the resulting Thymio robot that tackle these problems in a holistic way.
@@
 %The robot and its development environment are open source and the whole project follows an open distribution and participative model.
-%Finally, we show deployment statistics proving that the robot is gender- and age-neutral and preliminary results towards assessing the educational value of robots.
+%Finally, we show deployment statistics proving that the robot is gender- and age-neutral and preliminary results toward assessing the educational value of robots.
 %\end{abstract}
@@
 
-Mobile robots are interesting tools for education because of both the enthusiasm they raise and the multidisciplinary nature of their technology.
+Mobile robots are valuable tools for education because of both the enthusiasm they raise and the multidisciplinary nature of their technology.
 Because autonomous mobile robots sense the environment and take actions based on their perception, they seem to display \emph{intentions of their own}.
-This fascinates the users of the robot and creates feelings of accomplishment and power in the creators.
-The permeating presence of robots in science fiction and their projected use in our society increases these emotions by giving a sense of touching the future. % SM: not so happy with my sentence here
+This fascinates the users of the robot and creates feelings of accomplishment and power when these users engage in creative activities with the robots.
+The permeating presence of robots in science fiction and their projected use in our society increases these emotions by giving a sense of touching the future.
 Children can spend hours looking at a robot interacting with its environment and adults can approach robots as a new hobby that replaces radio and electronics kits.
 Robots embed various technologies and therefore give access to a wide range of fields, such as complex mechanics, sensors, wireless transmission, mathematics, computer science, etc.
-The emotional potential of mobile robots and their varied technology makes them the ideal tool for education at large.
+The emotional potential of mobile robots and their varied technology makes them a potential ideal tool for education.
 
-Despite this potential, robots are still not widespread in schools as they could be and this for several reasons. Among them we can mention:
+However, robots are still not as widespread in schools as they could be, for several reasons:
 \begin{enumerate}
-\item Although there are many robotic projects developing innovative and interesting educational robots, few reach a sufficient maturity to become distributed and accessible to schools. 
+\item Although there are many robotics projects developing innovative and interesting educational robots, few reach a sufficient maturity to become distributed and accessible to schools. 
 \item A robot performing interesting behaviors is a complex piece of technology and therefore expensive. 
@@
 \item Introducing robotic tools into teaching activities requires investment in time and training for the teachers.
-In Europe, despite a trend in including more and more technological tools into the learning process, teachers are still insufficiently trained and are reluctant to introduce these tools in their teaching activity~\cite{CERI2008}.
-To be accepted by teachers, robots must be therefore both accessible with minimal effort and be associated with-well prepared educational material. 
-Moreover an infrastructure should allow the exchange of educational material among the teachers.
+In Europe, despite a trend to increase the use of technological tools in the learning process, teachers are still insufficiently trained and are reluctant to introduce these tools in their teaching activity~\cite{CERI2008}.
+Therefore, to be accepted by teachers, robots must be both accessible with minimal effort and accompanied by well-prepared educational material. 
+Moreover, infrastructure should be provided to allow the sharing of educational material among teachers.
 \item Robot construction, use, and programming is often perceived as a boyish activity in our society~\cite{leonard2009lego,nourbakhsh2009robot}.
@@
 This strongly limits the potential of robots as general-purpose educational tools, especially in schools.
-\item Finally, many teachers are reluctant to follow volatile trends, especially if based on purely commercial arguments.
-Teachers prefer to invest in stable tools, and this is the opposite of todays consumer technology, often very volatile and based on continuos renewal rather than on durability.
+\item Finally, many teachers are reluctant to follow volatile trends, especially if these are based on purely commercial arguments.
+Teachers prefer to invest in stable tools, and this is the opposite as the trend of current consumer technology, which is often very volatile and based on continuous renewal rather than on durability.
 \end{enumerate}
 
-Open source hardware projects can address several of these issues in a different way than pure commercial products. 
-By open source hardware we understand, following the definition of the Open Source Hardware Association\footnote{\url{http://www.oshwa.org/definition/}}, ``hardware whose design is made publicly available so that anyone can study, modify, distribute, make, and sell the design or hardware based on that design''.
-In this paper, we show that this concept, implemented as a project in a community of users, developers and producers, brings a very interesting added value to the robot and to the educational methods.
+Open source hardware projects can address several of these issues in a different way than closed-source purely commercial products. 
+By \textit{open source hardware} we mean, following the definition of the Open Source Hardware Association\footnote{\url{http://www.oshwa.org/definition/}}, ``hardware whose design is made publicly available so that anyone can study, modify, distribute, make, and sell the design or hardware based on that design''.
+In this paper, we show that this concept, implemented as a project gathering a community of users, developers and manufacturers, brings a strong added value to the robot and to the educational methods.
 This is illustrated by the results achieved with Thymio and is compared with the feedback of a group of robotics engineers active in open source hardware design.
+% FIXME: SM: I do not understand this link, why do we compare with other open source hardware products, as it seems that until now the arguing is against closed source products?
 We can well observe both similarities and differences between our education-centered project and other open source robotic hardware projects, pointing to challenges and opportunities.
@@
 
-There is a very large number of publications presenting educational robots with different prices and features, from very low-cost systems targeting, for instance, education in Africa~\cite{Rubenstein2015,Gyebi2015} to extremely sophisticated humanoids~\cite{Hood2015,Mazzoni2016}.
-Among those, only very few are commercially available, limiting their validation by education scientists.
-This results in 90\% of publications about validation of educational results made on LEGO\textsuperscript{\textregistered} Mindstorms\textsuperscript{\textregistered}~\cite{benitti2012exploring}.
+There is a large number of publications presenting educational robots with different prices and features, from low-cost systems for instance gtargeting education in Africa~\cite{Rubenstein2015,Gyebi2015} to extremely sophisticated humanoids~\cite{Hood2015,Mazzoni2016}.
+Among those, only a handful are commercially available, limiting their validation by educational scientists.
+This results in 90\,\% of publications about validation of educational results made on LEGO\textsuperscript{\textregistered} Mindstorms\textsuperscript{\textregistered}~\cite{benitti2012exploring}.
+% FIXME: SM: is that in general or only about STEM/programming?
 The LEGO\textsuperscript{\textregistered} Mindstorms\textsuperscript{\textregistered} EV3\footnote{\url{http://mindstorms.lego.com}} is expensive ($\approx$ 400\,\$) but offers a wide range of possibilities, especially at the mechanical level using LEGO\textsuperscript{\textregistered} bricks and at the software level with its graphical programming environment.
@@
 The WeDo\textsuperscript{\textregistered} is similar to EV3 but only reads one sensor and controls one actuator and must always be connected to a desktop computer.
-The BeeBot robot\footnote{\url{http://www.beebot.org.uk}} is a cheap ($\approx$ 90\,\$) autonomous system.
+% FIXME: SM: The WeDo falls a bit from the sky here, although I see the underlying idea...
+The BeeBot robot\footnote{\url{http://www.beebot.org.uk}} is a cheap ($\approx$ 90\,\$) mobile system.
 However, it is not a real robot but rather an automaton: There is no perception of the environment, just a good encoder for precise movements and a simple programming interface based on buttons giving directions for the movements.
@@
 Among these widely used robots, none is part of an open source hardware project.
-The only effort in this direction comes from LEGO\textsuperscript{\textregistered}, that made public all the schematics of the EV3 in an hardware developer kit, to allow the design of extensions. 
-Otherwise, only few educational robots are both used in schools and are issued from an open hardware project:
-Scribbler2, produced and sold by Parallax\footnote{\url{http://www.parallax.com}} ($\approx$ 180\,\$), is a 188~mm large robot, designed to run on the ground and equipped with few light sensors, one distance sensor, two ground sensors and few LEDs display.
+%The only move in this direction comes from LEGO\textsuperscript{\textregistered}, that made public all the schematics of the EV3 in an hardware developer kit, to allow the design of extensions.
+% FIXME: if the LEGO license is not a true open hardware one, I would drop the previous sentence, it does not bring much
+In niche projects, only few open source hardware educational robots are used in schools:
+Scribbler2, produced and sold by Parallax\footnote{\url{http://www.parallax.com}} ($\approx$ 180\,\$), is a 188\,mm large robot, designed to run on the ground and equipped with few light sensors, one distance sensor, two ground sensors and few \textsc{led} displays.
 It runs on standard AA batteries and has a hacker port for interfacing electronic extensions.
-It is programmable with a graphic or text code interface.  
+It is programmable with a graphical or a textual code interface.  
 The main weakness of Scribbler2 is its performance per price ratio and the limited compatibility with other systems. 
-Moreover there seems to be no real community around its development.
+Moreover, there seems to be no active community around its development.
 The e-puck~\cite{mondada2009puck} robot is a robot targeting university-level education.
-Well equipped with sensors and actuators, modular and compact, it can be programmed with professional environments.
+Well equipped with sensors and actuators, modular and compact, it can be programmed with industry-standard environments.
 Several simulators allow to run experiments of high complexity. 
 There is an active community of users and developers around it.
-It's main weakness is its high price ($\approx$ 870\,\$).
+Its main weakness is its high price ($\approx$ 870\,\$).
 Finch\footnote{\url{http://www.finchrobot.com/}} ($\approx$ 99\,\$) is a very simple robot that has been designed around a wired connection to the computer. 
 This connection reduces electronics requirements, such as batteries or wireless communication, and allows to implement the control entirely on the computer.
-This results in a very broad set of possible programming languages available, the real force of this robot. 
-On the other hand the cable does not allow real autonomy and mobility of the robot.
-Finally the mBot\footnote{\url{http://www.makeblock.cc/mbot/}} is a mobile platform based on the Arduino board. 
-The very cheap and simple electronics allow to drastically reduce the price of the system ($\approx$ 75\,\$) but also reduces the perception possibilities to a couple of sensors.
+This results in a very broad set of possible programming languages available, which is the real force of this robot. 
+However, the cable does not allow real autonomy and mobility.
+Finally, the mBot\footnote{\url{http://www.makeblock.cc/mbot/}} is a mobile platform based on an Arduino board.
+Its electronics is simple and inexpensive and the robot only features a couple of sensors, which allows to drastically reduce its price ($\approx$ 75\,\$), but also limits the perception possibilities and therefore the span of use.
 
-In respect to these robots, Thymio has a compact size (120~mm), many interaction possibilities, an affordable price and a large set of sensors.
-It's design makes the best use of the recent cheap components coming from the explosive growth of the mobile devices market. 
-To the best of our knowledge, beside Thymio there are no educational products exploiting this and providing many sensors and actuators at a price below 130\,\$.
+With respect to these robots, Thymio has a compact size (120~mm), many interaction possibilities, an affordable price and a large set of sensors.
+Its design makes the best use of the recent cheap components coming from the explosive growth of the mobile devices market. 
+To the best of our knowledge, beside Thymio there are no educational products providing a similar integration of sensors and actuators at a price below 130\,\$.
 
-Among the main target users there are teachers.
-For teachers, the motivation to use robotic tools in their classes depends on many factors \cite{chevalier2016}.
+Teachers are one of the main target user group of the Thymio project.
+For teachers, the motivation to use robotic tools in their classes depends on many factors~\cite{chevalier2016}.
 Among them, the availability of material and training plays a key role.
 The BeeBot mentioned previously, because of its radical simplicity, is one of the tools that needs the least training. 
-Moreover its use in learning activities is based on a simple grid on a printed mat on the ground, allowing fast and easy integration in existing lessons.
+Moreover its use in learning activities is based on a simple grid on a printed map on the ground, allowing fast and easy integration in existing lessons.
 These are probably two important factors for its success in schools~\cite{2008highfield,2008demichele} and the availability of a rich set of educational material\footnote{\url{https://www.learningplace.com.au/deliver/content.asp?pid=38840} and \\ \url{http://beebots.skola.edu.mt/}}.
 However, the total lack of sensors reduces the BeeBot to an automaton, easy to use but limited to very simple activities for young kids, without much potential in teaching computational thinking or robotics concepts.
+% FIXME: SM: I would even say that the BeeBot targets a specfic concept for kids, which is the learning of sequences, and is very good at it.
+I think we could find a strong educational argument to show how limited it is.
+maybe Morgane can help.
 
-For more complex robots, both the development of educational material and the training of teachers require a very important effort, needing a good mix between robotics and educational skills. 
+For more complex robots, both the development of educational material and the training of teachers require a huge work, demanding a good mix between robotics and educational skills. 
 Moreover, educational material changes very much from school to school, as education is very dependent from local educational programs and languages, for instance. 
-In the particular case of Switzerland, for instance, every of the 26 states composing the confederation has its own school program and its own language among the four national ones. 
-Educational material and the related training have therefore to be adapted to each situation, which requires a huge effort.
-Even for large companies like LEGO\textsuperscript{\textregistered} this development is not affordable. 
-A solution is to have an active community of users contributing to the development of the material in a distributed manner, adapting the material to the local situation.
-LEGO\textsuperscript{\textregistered} itself is moving in the direction of promoting communities of users \cite{Hienerth2014} with a user-producer interaction close to those of open hardware projects.
-The open source community regrouping developers, producers and final users is therefore a very interesting model to address the distributed development and sharing of educational material, and the diffusion of training sessions.
-This model is an extension to education of what already practiced in projects like Arduino~\cite{Jamieson2010}.
+In the particular case of Switzerland, for instance, every of the 26 states composing the country has its own school program and its own language among the four national ones. 
+Educational material and the related training must therefore be adapted to each situation.
+Even for large companies like LEGO\textsuperscript{\textregistered}, such a development is not sustainable. 
+To solve this problem, a crowd-sourcing approach can be used.
+An active community of users can contribute to the development of the material in a distributed manner, adapting the material to the local situation.
+LEGO\textsuperscript{\textregistered} itself is moving toward this direction by promoting communities of users~\cite{Hienerth2014} with a user-producer interaction close to those of open hardware projects.
+An open source community regrouping developers, manufacturers and end users is therefore a very interesting model to address the distributed development and sharing of educational material, and the diffusion of training sessions.
+This model is an extension to education of what is already practiced in projects such as Arduino~\cite{Jamieson2010}.
+
+% FIXME: SM: I am lacking a link, what is the contribution of the paper in that context?
 
@@
 	\includegraphics[width=.8\columnwidth]{figures/front_2016}
+	%FIXME: SM: maybe we should have a ruler on the side to show the scale?
 	\vskip 1.5em
 	\includegraphics[width=\columnwidth]{figures/wirelessThymioVecto}
-	\caption{The Thymio robot and its main components for the Wireless and the plain version.}
+	\caption{The Thymio robot and its main components for the Wireless and the \textsc{usb}-connected versions.}
 	\label{fig:schematique}
@@
 
-We designed the Thymio robot along seven main axes: a low price to address a larger number of users, a feature set that suits both genders and multiple ages from young children to adults, a combination of sensors, actuators, and programming features that facilitates learning, a mechanical design that allows creative extensions by the users themselves, a set of ready-to-use behaviors to get a demonstration right out of the box, an accessible programming environment, and an open source  community contributing to the design and disseminating the results.
-The result is a miniature (10\,cm in diameter, see Figure~\ref{fig:schematique}) differential-wheeled robot suited for use on a desktop.
+We designed the Thymio robot along seven main axes: a low price to address a larger number of users, a feature set that suits both genders and multiple ages from young children to adults, a combination of sensors, actuators, and programming features that facilitates learning, a mechanical design that allows creative extensions by the users themselves, a set of ready-to-use behaviors to get a demonstration right out of the box, an accessible programming environment, and an open source community contributing to the design and the dissemination.
+The result is a miniature (10\,cm in side) differential-wheeled robot suited for use on a desktop (Fig.~\ref{fig:schematique}, top).
 The robot is robust enough to be mishandled by children, as it can fall from a table without breaking itself.
-It features a translucent white hull and a wide range of sensors and actuators (see Figure~\ref{fig:schematique} bottom).
+It features a translucent white hull and a wide range of sensors and actuators (Fig.~\ref{fig:schematique}, bottom).
 The robot embeds a battery, rechargable by \textsc{usb}, that provides an autonomy of about 2 hours.
 
-We start by presenting here the first six main axes, before moving to the next section devoted the open hardware specificities.
+%We start by presenting here the first six main axes, before moving to the next section devoted the open hardware specificities.
+% FIXME: SM: what do you mean by specificities here?
 
@@
 A three-axis accelerometer measures the orientation of the robot with respect to gravity, and can detect shocks and free falls.
-Five capacitive touch buttons organized as an arrow form an intuitive user interface.
+Five capacitive touch buttons organized as a direction pad form an intuitive user interface.
 Compared to physical buttons, these simplify the plastic hull of the robot and make it more robust.
@@
 However, choosing the optimal combination of sensors and processor is a difficult problem, as hardware interfaces are limited on a microcontroller and therefore compromises must be made.
-We evaluated several microcontrollers and chose the PIC24F from Microchip because it integrates a USB interface and can drive capacitive touch buttons directly, saving additional components. 
+We evaluated several microcontrollers and chose the PIC24F from Microchip because it integrates a \textsc{usb} interface and can drive capacitive touch buttons directly, saving additional components. 
 This microcontroller controls all sensors and actuators, excepted the battery recharging logic, which uses a specific chip for safety reasons.
-%This choice of a single microcontroller led us to non-trivial design decisions like using low-cost shift registers to drive the many LEDs.
+%This choice of a single microcontroller led us to non-trivial design decisions like using low-cost shift registers to drive the many \textsc{led}s.
 %Moreover, instead of directly connecting the infrared sensors to an analogical input, we interface them using a comparator that transforms the analogical signal into a binary one with a pulse having a duration proportional to the intensity of the original signal~\cite{wanda}.
@@
 Because manual work is expensive in western countries, we produce through subcontractors in China.
-Indeed, while most of the electronics consists of \textsc{smd} components mounted by a robot, some are soldered (sensors, microphone, LEDs) by hand and the robot is also assembled by hand.
+Indeed, while most of the electronics consists of \textsc{smd} components mounted by a robot, some are soldered (sensors, microphone, \textsc{led}s) by hand and the robot is also assembled by hand.
 We streamlined the production of plastic parts (Thymio has only 5 parts while our previous prototype had 11) and opted for injection molding. 
@@
 \end{tabularx}
-\caption{Production cost of basic Thymio (without wireless modules).}
+\caption{Production cost of \textsc{usb}-connected Thymio.}
 \label{tbl:thymio-price}
@@
 
-Thymio finally costs 35\,\$ per unit when produced in a batch of 1\,k units (see Table~\ref{tbl:thymio-price}), with fixed fees of 15\,k\,\$ (13\,k\,\$ for the mold and 2\,k\,\$ for the \textsc{ce} certification).
-This allows a selling price of 129\,CHF ($\approx$ 130\,\$).
+Thymio finally costs 35\,\$ per unit when produced in a batch of 1\,k units (see Table~\ref{tbl:thymio-price}), with fixed fees of 15\,k\$ (13\,k\$ for the mold and 2\,k\$ for the \textsc{ce} certification).
+This allows a selling price of \textsc{chf}\,129 ($\approx$ 130\,\$).
 
@@
 The variety of sensors, the multiple ways of interacting with the robot, its hull design, and its customization make Thymio accessible to girls and boys of different age groups.
-The white hull is initially neutral and children can choose their own color using the \textsc{rgb} LEDs or attach accessories to the hull.
+The white hull is initially neutral and children can choose their own color using the \textsc{rgb} \textsc{led}s or attach accessories to the hull.
 The robot can be used from children as young as 6 years old to adults:
@@
 
-At the hardware level, we render visible the activity of the sensors by adding a LED next to each of them, for a total of 39 LEDs.
-These LEDs locally color the hull and allow the user to see immediately where and when the robot perceives a change in its environment.
-There is a LED next to each proximity sensor that lights up as soon as an object is close enough to be seen, and shines brighter the closer the object gets.
-Similarly, a combination of blue and red LEDs show the temperature, and the infrared remote control receiver and the microphone both have LEDs that flash when they detect something.
-On the top of the robot, a circle of 8 LEDs shows the 3D inclination of the robot thanks to the accelerometer. 
+At the hardware level, we render visible the activity of the sensors by adding a \textsc{led} next to each of them, for a total of 39 \textsc{led}s.
+These \textsc{led}s locally color the hull and allow the user to see immediately where and when the robot perceives a change in its environment.
+There is a \textsc{led} next to each proximity sensor that lights up as soon as an object is close enough to be seen, and shines brighter the closer the object gets.
+Similarly, a combination of blue and red \textsc{led}s show the temperature, and the infrared remote control receiver and the microphone both have \textsc{led}s that flash when they detect something.
+On the top of the robot, a circle of 8 \textsc{led}s shows the 3D inclination of the robot thanks to the accelerometer. 
 This circle is also used to reflect some of the behaviors of Thymio.
-Finally, two strong \textsc{rgb} LEDs color the whole top of the robot.
+Finally, two strong \textsc{rgb} \textsc{led}s color the whole top of the robot.
 In addition to a visual feedback, the capacitive buttons also have an acoustic feedback.
-The link between a sensor and its feedback can be turned off when programming the robot to use the LEDs and the loudspeaker for other purposes.
+The link between a sensor and its feedback can be turned off when programming the robot to use the \textsc{led}s and the loudspeaker for other purposes.
 
-At the software level, our programming environment (see section \ref{sec:aseba}) provides a set of tools enabling beginners to discover progressively the basic rules of programming using a graphical interface, then the construction of a syntax by graphic blocs, and finally a full text-based coding environment with advanced debugging tools, such as real-time inspection of the variables of the robot and plotting features, providing a visual way to understand time-related concepts.
+At the software level, we provide a set of programming environments (see section \ref{sec:aseba}) that enable beginners to discover programming progressively.
+First, we teach them the basic rules of programming using a purely visual interface, then they discover the construction of syntax trees by assembling graphical blocks, and finally we provide a full text-based coding environment with advanced debugging tools, such as real-time inspection of the variables of the robot and plotting features, providing a visual way to understand time-related concepts.
 
@@
 
-We wanted the robot to be a starting point for the users to invent their own creations.
+We wanted the robot to be a starting point for users to invent their own creations.
 Therefore, we thought Thymio as a support for handicrafts and constructions.
@@
 The robot features a hole in its middle to insert a marker for drawing.
-It also has a slot for an SD card, allowing users to add their own sounds or to load a pre-compiled code by the simple insertion of a dedicated card.
+It also has a slot for a microSD card, allowing users to add their own sounds or to load a pre-compiled code by the simple insertion of a dedicated card.
 
@@
 There are four attachment positions on the top of the robot and two LEGO\textsuperscript{\textregistered} Technics crosses on each side.
-Additionally, the two wheels have attachment points, which permits to use them to actuate elements elsewhere in the structure (see Figure~\ref{fig:example-construction}, third row left) or to lift the robot's own weight (see Figure~\ref{fig:example-construction}, third row right).
+Additionally, the two wheels have attachment points, which permits to use them to actuate elements elsewhere in the structure (Fig.~\ref{fig:example-construction}, third row left) or to lift the robot's own weight (Fig.~\ref{fig:example-construction}, third row right).
 Therefore we chose more powerful motors than strictly necessary to move the robot around.
 The same fixation points used for LEGO\textsuperscript{\textregistered} bricks can be used to attach paper structures.
-Paper can allow to simply change the body shape and add some body movements, as illustrated in Figure~\ref{fig:example-construction} by the orca, opening and closing its mouth while moving forward or by the bat, moving its wings.
-But paper and cardboard can also radically change the locomotion principle, as illustrated in the second row of Figure~\ref{fig:example-construction} by the zombie, where the robot's wheel activate the legs.
-The paper structure can be also used to interact with the sensors, as illustrated in the second row of Figure~\ref{fig:example-construction} by the bear, who extends his paw in front of the sensors to drive its iceberg.
-The same fixation points can be used to attach 3D printed customized parts, as illustrated by the winder shown in the fourth row of Figure~\ref{fig:example-construction}.
-Finally one can use paper to create environments, either flat with patterns that can be used in association to the ground sensors (see Figure~\ref{fig:example-construction} bottom) or 3D object, like the trees beside the zombie in Figure~\ref{fig:example-construction}.
+Paper can allow to simply change the body shape and add some body movements, as illustrated in Fig.~\ref{fig:example-construction} by the orca, opening and closing its mouth while moving forward or by the bat, moving its wings.
+But paper and cardboard can also radically change the locomotion principle, as illustrated in the second row of Fig.~\ref{fig:example-construction} by the zombie, where the wheels of the robot activate the legs.
+The paper structure can be also used to interact with the sensors, as illustrated in the second row of Fig.~\ref{fig:example-construction} by the bear, who extends his paw in front of the sensors to drive its iceberg (the robot).
+The same fixation points can be used to attach 3D printed customized parts, as illustrated by the winder shown in the fourth row of Fig.~\ref{fig:example-construction}.
+Moreover, one can use paper to create environments, either flat with patterns that can be used in association with the ground sensors (Fig.~\ref{fig:example-construction}, bottom) or 3D object, such as the trees beside the zombie in Fig.~\ref{fig:example-construction}.
 Finally, it is also possible to link several Thymio by software, allowing the coordination of complex multi-Thymio robotic structures.
@@
 These basic behaviors allow people starting Thymio to immediately interact with it, while illustrating the many possibilities of the robot.
-The behaviors are interactive and oriented towards the user. % SM: FIXME: not clear what "oriented towards the user" means
+The behaviors are interactive and oriented toward the user. % SM: FIXME: not clear what "oriented toward the user" means
 For each behavior, the robot's body has a different color, allowing the user to easily recognize it.
 The user can navigate between the colors with the buttons and select the behavior she/he wants to use.
-Basic behaviors can be exploited in a construction to create reactivity without the use of programming (see Figure~\ref{fig:example-construction}, the top four paper creations can be used with basic behaviors).
+These behaviors can be exploited in a construction to create reactivity without the use of programming, such as in the paper creations shown in the top four images of Fig.~\ref{fig:example-construction}.
 
@@
 Aseba is designed to enable novices to program robots easily.
-On the robot side, it provides a lightweight virtual machine that runs on microcontrollers like the PIC24F.
-A virtual machine allows instantaneous upload and safe execution of code.
-On the desktop side, Aseba provides an IDE which allows to program in a purely graphical mode (called VPL, see Figure~\ref{fig:vpl}), or using a scripting language (see Figure~\ref{fig:aseba-studio}) or in a mixed way, assembling code in a graphical way using the blockly environment\footnote{\url{https://developers.google.com/blockly/}} (see Figure~\ref{fig:blockly}).
-This allows the IDE to match with the multi-age goal described in section~\ref{sec:multi}.
+On the robot side, it provides a lightweight virtual machine that runs on microcontrollers such as the PIC24F inside Thymio.
+A virtual machine allows instantaneous upload and safe execution of programs.
+On the desktop side, Aseba provides an \ac{ide} featuring a \ac{vpl} (Fig.~\ref{fig:vpl}), a scripting language (Fig.~\ref{fig:aseba-studio}), and a mixed language, Blockly\footnote{\url{https://developers.google.com/blockly/}}, to assemble scripts graphically (Fig.~\ref{fig:blockly}).
+These different languages cover the abilities of children of different ages and the progression of skills of learners.
 
@@
 
-The IDE integrates real-time feedback both graphically under VPL~\cite{Magnenat2015} and displaying variables or graphs in the text mode.
-The IDE also provides a help documentation\footnote{currently English, French, and German}.
+The \ac{ide} integrates real-time feedback of the status of program execution, as this feature was recognized of critical importance to properly learn programming~\cite{sorva2013notional}.
+This capability is provided both with the \ac{vpl}~\cite{Magnenat2015} and in the scripting environment, by displaying the content of variables in real time through texts or plots.
+The \ac{ide} also provides a help documentation\footnote{currently in English, French, German, Italian and Spanish}.
 In addition, the wiki pages of the project provide tutorials on programming with Thymio and a detailed description of the programming interface.
-%The latter was designed with consistency in mind: variables and functions are available in namespaces and similar features (such as the different LEDs) are accessed the same way.
+%The latter was designed with consistency in mind: variables and functions are available in namespaces and similar features (such as the different \textsc{led}s) are accessed the same way.
 %It also provides native functions for school-level mathematics such as vector operations, trigonometry, statistics and functions such as sorting and random.
@@
 
+% FIXME: SM: is there a reason to limit images to 90% width?
 \begin{figure}
@@
 \includegraphics[width=.9\columnwidth]{figures/vpl}
-\caption{The graphic programming environment VPL.}
+\caption{The visual programming language.}
 \label{fig:vpl}
@@
 \includegraphics[width=.9\columnwidth]{figures/blockly}
-\caption{The blockly programming environment.}
+\caption{The Blockly programming environment.}
 \label{fig:blockly}
@@
 Both were targeting similar users and have been sold in similar quantities.
-What we can observe after more than 10 years is that Khepera generated many royalties for the university, but less relative academic visibility than e-puck (see Figure~\ref{fig:khepuck}) .
-The e-puck robot, being open hardware and with an image better linked to the university, generated no income for the university but much more visibility in respect to his diffusion.
+What we can observe after more than 10 years is that Khepera generated many royalties for the university, but less relative academic visibility than e-puck (Fig.~\ref{fig:khepuck}).
+% FIXME: SM: "many royalties" is unclear
+The e-puck robot, being open hardware and with an image better linked to the university, generated no income for the university but much more visibility.
 Khepera was more linked with the name of the company producing it, the e-puck robot was nearly not linked with the producer when mentioned in the literature.
@@
 \includegraphics[width=\columnwidth]{figures/others_robots}
-\caption{Comparison between mentions in the literature, mention of the manufacturer and citation of the reference paper for two robots disseminated by our lab in the past.
-It appears that the Khepera robot, distributed with a proprietary strategy, was more linked to the company producing it than the original designers.
-This is the opposite for the e-puck robot, distributed with an open source license.}
+\caption{Comparison of the impact of e-puck and Khepera.
+%NOTE: SM: this is an analysis and should not be in caption
+%between mentions in the literature, mention of the manufacturer and citation of the reference paper for two robots disseminated by our lab in the past.
+%It appears that the Khepera robot, distributed with a proprietary strategy, was more linked to the company producing it than the original designers.
+This is the opposite for the e-puck robot, distributed with an open source license.
+}
 \label{fig:khepuck}
@@
 The personal motivation to participate to such a project is very different than the institutional one. 
-When asked about their personal motivation, the people participating to the survey give as main motivation a link to their professional activity and to the specific project, followed by a more general motivation by the nature of the project and finally by the engineering practice (see Figure~\ref{fig:motivation}).
+When asked about their personal motivation, the people participating to the survey give as main motivation a link to their professional activity and to the specific project, followed by a more general motivation by the nature of the project and finally by the engineering practice (Fig.~\ref{fig:motivation}).
 The more abstract goal of improving our society is mentioned only by few respondents.
@@
 \includegraphics[width=\columnwidth]{figures/motivation}
-\caption{Motivation of the people participating to the survey.}
+\caption{Motivation of contributing to open source hardware.}
 \label{fig:motivation}
@@
 Also this motivation is therefore stronger in our team.
+% FIXME: SM: duplication about "our team"
 Although this seems a minor issue, sharing a strong fundamental motivation such as education or scientific achievements, is a key element for a strong community~\cite{Stahlbrost2011}, especially if interdisciplinary like ours.
 
-It is also interesting to look at what people expect as benefits in participating in such a project (see Figure~\ref{fig:getout}).
-Most of the answers mention impact, networking and learning. 
-In these answers we can observe both a technological component in getting better results and impact, and a human-relations component that is not in the basic definition of open hardware but results from the community created around the project.
-People expect to work with other people to achieve more together, to create a network of competences or users, to learn through new experience and contacts.
+It is also interesting to look at what people expect as benefits in participating in such a project (Fig.~\ref{fig:getout}).
+We can observe both a technological component in getting better results and impact, and a human-relations component that is not in the basic definition of open hardware but results from the community created around the project.
+People expect to work with other people to achieve more together, to create a network, and to learn through new experience and contacts.
 
@@
 Working together with several partners was for everybody a win-win situation, and creating a community of users was the only solution to allow the development of high quality accessories and educational material.
-A wiki\footnote{http://www.thymio.org} has been established as meeting point for the learners, the robot developers, and the teachers.
-It introduces the robot, explains the basic behaviors, gives access to the programming environment and its documentation, and provides code samples and examples of constructions.
-The wiki is open for editing by anyone, and although we initially provided most of the material, the other members of the community have started to contribute.
+We established a wiki\footnote{http://www.thymio.org} as the meeting point for the learners, the robot developers, and the teachers.
+This wiki introduces the robot, explains the basic behaviors, gives access to the programming environment and its documentation, and provides code samples and examples of constructions.
+It is open for editing by anyone, and although we initially provided most of the material, the other members of the community have started to contribute.
 One of the most important contributions to the wiki, in term of effort, has been its translation into four languages.
@@
 \includegraphics[width=\columnwidth]{figures/advantages}
-\caption{Advantages that people participating to the survey perceive by contributing to an open source hardware project.}
+\caption{Advantages of contributing to open source hardware.}
 \label{fig:getout}
@@
 
-When starting an open source hardware project, one of the typical questions is about which license to use when disseminating the project source files. 
+When starting an open source hardware project, one of the typical questions is which license to use when disseminating the project source files. 
 We will not discuss this matter here, as it is a very common and well-covered issue.
 
-There is another licensing issue which is not well known and that we discovered very late in our project: the constrains of the license of the mechanical and electronic CAD tools. 
-Indeed, asked about this issue, the participants to our survey seem generally not aware of the fact that CAD licenses can be very restrictive about the way the source files can be published (see Figure~\ref{fig:aware}).
-More than the half or respondents to our survey, and more than 60\% among the project leaders, state that they were not aware that this could be an issue, and only one third checked the license of their CAD software.
-Among the respondents, some stated that they did not checked the license because they were sure they can freely publish their design, but one third of them are using software not allowing the publication of source files when using educational licenses, and all these respondents are academic or students.
-This issue is very serious, as most contributors of open source hardware projects are academic and use academic or educational licenses. 
+There is another licensing issue which is not well known and that we discovered very late in our project: the constraints of the license of the mechanics and electronics \textsc{cad} tools. 
+Indeed, when asked about this issue, the participants to our survey seem generally not aware of the fact that \textsc{cad} licenses can be very restrictive about the way the source files can be published (Fig.~\ref{fig:aware}).
+More than the half or respondents to our survey state that they were not aware that this could be an issue, and only one third checked the license of their \textsc{cad} software. 
+%More than the half or respondents to our survey, and more than 60\% among the project leaders, state that they were not aware that this could be an issue, and only one third checked the license of their \textsc{cad} software.
+Some stated that they did not checked the license because they were sure they can freely publish their design, but one third of them are using software not allowing the publication of source files when using educational licenses, and all these respondents are academics or students.
+% FIXME: SM: there is a bit of duplication with previous sentence, I think these can be compressed
+This issue is very serious, as most contributors of open source hardware projects are academics and use academic or educational licenses. 
 Most licenses do not allow to use the design created with this version of the software for commercial activities. 
@@
 \includegraphics[width=\columnwidth]{figures/checklicense}
-\caption{Awareness about CAD tool licenses limitations by people participating to the survey.}
+\caption{Awareness about \textsc{cad} tool licenses limitations.}
 \label{fig:aware}
@@
 
-To clarify this issue we contacted twelve of the major editors of mechanical CAD and PCB routing software.
+To clarify this issue we contacted twelve of the major editors of mechanical \textsc{cad} and \textsc{pcb} routing software.
 We asked them if their educational license allows the publication of the source files, and we specified that the published files could have been downloaded by a company to produce the system for commercial purposes.
-Figure~\ref{fig:editors} summarizes the result of this survey.
+Fig.~\ref{fig:editors} summarizes the result of this survey.
 Only three of the twelve editors have education licenses allowing this type of publication. 
 Two others explicitly mentioned the possibility to ask for permission before publication.
-A large mechanical CAD editor was puzzled by our questions and after realizing the impact of the license, introduced a special condition allowing publication of files in clearly labeled open source hardware projects.
+A large mechanical \textsc{cad} editor was puzzled by our questions and after realizing the impact of the license, introduced a special condition allowing publication of files in clearly-labeled open source hardware projects.
 In previous situations of open source publication, the same editor asked the universities to purchase commercial licenses to permit publication.
-This can multiply by a factor of some hundreds the price of the CAD license.
+This can multiply by factors of hundreds the price of the \textsc{cad} license.
 This blocking factor for open hardware also applies to the publication of scientific results, as promoted by many governments in the last decade and generally called ``open science''.
-The change of policy in CAD files publication that we obtained is a sign of the very positive trend set by open hardware and by open science in general.
+The change of policy in \textsc{cad} files publication that we obtained is a sign of the very positive trend set by open hardware and by open science in general.
 
@@
 \includegraphics[width=\columnwidth]{figures/table}
-\caption{Publication possibilities as function of the CAD editors, situation end of March 2016.}
+\caption{Publication possibilities as function of the \textsc{cad} editors, situation at the end of March 2016.}
 \label{fig:editors}
@@
 
-As a conclusion, this legal issue is totally underestimated by both people participating to the projects and by the CAD editors. 
+As a conclusion, this legal issue is totally underestimated by both people participating to the projects and by the \textsc{cad} editors. 
 In a period where editors are looking for additional revenues and are attacking universities for misuse of licenses\footnote{We are aware, just in Switzerland, of two situations where large amounts of money are asked to universities for non-respect of software licenses.}, this can be a very dangerous situation.
@@
 In the definition of open hardware we gave, it is stated that one should offer an ``hardware whose design is made publicly available so that anyone can ... make... the ... hardware ''.
+% FIXME: SM: I do not understand the dots
 Behind the ``anyone'' should we consider every single person or only companies able to produce the product?
@@
 The robot is the expensive part and has a very neutral design, allowing adaptation to specific situations.
-This adaptation is achieved by custom accessories that increase the attraction of the robot in its specific role, enabling activity for different ages and genders.
+This adaptation is achieved by custom accessories that increase the attraction of the robot in its specific role, enabling activities for different ages and genders.
 
-For the robot itself we opted for the second interpretation of the definition, considering under ``anyone'' only the professional structures able to mass-produce hardware, for a reason of price and complexity of the product.
+For the robot itself, we opted for the interpretation of the definition that considers under ``anyone'' only the professional structures able to mass-produce hardware, for a reason of price and complexity of the product.
 When looking for the best performance per price ratio, we decided to use techniques that require heavy equipment.
@@
 
-For the accessories, less technically challenging and stronger linked with creativity and educational value, we promoted techniques that are accessible to ``anyone'' in the broader sense: paper, cardboard, LEGO\textsuperscript{\textregistered} constructions and 3D printing, as explained in section~\ref{sec:crea} and illustrated in Figure~\ref{fig:example-construction}.
-This allows a much broader spectrum of contributor, especially among the teachers.
+For the accessories, less technically challenging and stronger linked with creativity and educational value, we promoted techniques that are accessible to ``anyone'' in the broader sense: paper, cardboard, LEGO\textsuperscript{\textregistered} constructions and 3D printing.
+%, as explained in section~\ref{sec:crea} and illustrated in Fig.~\ref{fig:example-construction}.
+This allows a much broader spectrum of contributors, including teachers and lay people.
 
@@
 \includegraphics[width=\columnwidth]{figures/manufacturing}
-\caption{Production methods in the projects mentioned by the people participating to the survey.}
+\caption{Production methods in the mentioned projects.}
 \label{fig:production}
@@
 
-When looking at the results of the survey (see Figure~\ref{fig:production}), we can observe that the most common production method is a centralized one by commercial companies.
+Our survey shows that the most common production method is a centralized one by commercial companies (Fig.~\ref{fig:production}).
 Indeed, large open source hardware projects like Arduino are based on this model.
-One could advocate that this choice of central production and sale of hardware can bring these robotic projects closer to standard commercial models, ending in users not involved in the project and not feeling close to the developers.
-In our case we decided to counterbalance this apparently business oriented model by creating a non-profit association, called Mobsya, in charge of producing and selling the robots. 
-This is also a very well represented production method in our survey.
-Of course anybody else could produce the Thymio based on more buiness-oriented models, but at least the initial production has been done respecting a coherent and global non-profit strategy, in line with educational values of the community of users.
-Moreover the user are better integrated is the second layer of open source hardware consisting in accessories, where anybody can contribute with own designs.
+One could argue that this model makes these robotic projects close to standard commercial models, leading to a lack of involvement from users and loosing the open source spirit.
+In our case, we decided to alleviate this effect by creating a non-profit association, called Mobsya, in charge of producing and selling the robots. 
+%While anybody else could produce the Thymio follwing a more buiness-oriented model, the initial production has been done respecting a coherent and global non-profit strategy, in line with educational values of the community of users.
+Moreover, the users are better integrated in the second layer of open source hardware consisting of accessories, where anybody can contribute their own designs and put them on the wiki.
 
-Another strong difference between a standard consumer product and our vision of open hardware, is that Thymio should be durable.
-Schools do not have large budget for technological tools and make strong investment in training of their teachers when adopting a new tool.
-Therefore the lifetime of the products should be as long as possible. 
+Another strong difference between a standard consumer product and our vision of open hardware is that Thymio should be durable.
+Schools do not have large budgets for technological tools and make strong investment in training of their teachers when adopting a new tool.
+Therefore, the lifetime of the products should be as long as possible. 
 The open hardware approach gives to the user, or to a generic technician, better conditions to repair the system.
-Supporting this type of operation has an impact on the robot design; for instance Thymio can be easily opened with four standard screws, and we introduced connectors between key elements such as motors, speaker, or the battery and the main printed circuit.
+Supporting this type of operation has an impact on the robot design; for instance Thymio can be easily opened with four standard screws, and we introduced connectors between key elements such as motors, speaker, the battery and the main \textsc{pcb}.
 
+% FIXME: SM: in case of lack of space, we can strip this paragraph away
 Another key element in supporting repair by the users are the calibration methods. 
@@
 To correct this problem, one interesting solution is to introduce factory calibration, storing correction factors that are specific for each robot, in the microcontroller memory. 
-To allow the user to replace a broken motor, it is essential to give him also the possibility to re-calibrate the robot and adjust the parameters of the new motors.
+To allow the user to replace a broken motor, it is essential to give her also the possibility to re-calibrate the robot and adjust the parameters of the new motors.
 In Thymio, this results in the design and the documentation of calibration processes that can be performed by anyone, getting close to the original definition of open source hardware.
@@
 
-Beside the sale of more than 10'000 robots by a non-profit organization producing it, there are several elements that show a specific advantage of having chosen an open source hardware model.
+Beside the sale of more than 10'000 robots by Mobsya, there are several elements that show a specific advantage of having chosen an open source hardware model.
 
 The first element is linked to the financial support of the project.
-Although the non-profit nature of the project blocked the possibility to have shareholders injecting capital, this same nature enabled a lot of institutional donations linked with the societal goal of the project.
+Although the non-profit nature of the project blocked the possibility of having shareholders injecting capital, this same nature enabled a lot of institutional donations linked with the societal goal of the project.
 The financial support through donations instead of acquisition of shares, allows keeping a total decisional freedom and void debts at the same time.
-The main disadvantage of this approach is the amount of financial support, limited in our case of several hundreds of thousands of dollars.
+The main disadvantage of this approach is the amount of financial support, limited in our case to several hundreds of thousands of dollars.
 
-The second interesting element of the open source hardware appraoch is the very high acceptance of the resulting robot by universities and their offices in charge of promotion of science.
-Although these are not the primary user targeted by the robot, these are institutions with high visibility and influence on school programs. 
-In particular INRIA in France and the University of Cambridge in he United Kingdoms decided to use Thymio as tool for promotion of their activities and introduced it into schools.
-In France, INRIA developed several teaching modules and supported the training of more than 1000 teachers to the use of this particular robot. 
-This finally resulted in having the Thymio robot included as one of the tools chosen for the education of digital science in the reference books of the French Academy of Science and its foundation for promotion of science in education, called ``La main \`a la p\^ate.''
-The fact that Thymio is open source allows these institution to have a better control on the tool they use, as they can ensure maintenance and even further development of it. 
-The impulsion given by the academic institutions has been followed by many teachers who contributed with own teaching material. 
-Up to date one can find more than 50 teaching modules, one large tutorial, and many reports on school activities using Thymio.
+The second interesting element of the open source hardware approach is the very high acceptance of the resulting robot by universities and their offices in charge of promotion of science.
+Although these are not the primary users targeted by the robot, these are institutions with high visibility and influence on school programs. 
+In particular INRIA in France and the University of Cambridge in the United Kingdoms decided to use Thymio as tool for promoting their activities and introduced it into schools.
+In France, INRIA developed several teaching modules and supported the training of more than 1000 teachers to the use of Thymio. 
+This resulted in Thymio being one of the tools chosen for the education of digital science in the reference books of the French Academy of Science and its foundation for promotion of science in education, called ``La main \`a la p\^ate.''
+The open source nature of Thymio gives these institutions a large control, as they can ensure maintenance and even further develop the robot.
+The impulsion given by the academic institutions has been followed by many teachers who contributed their own teaching material. 
+Currently one can find more than 50 teaching modules, a large tutorial, and many reports on school activities using Thymio.
 
 In some cases, the open source nature of this project was a crucial political argument for the choice of this platform.
-In the state of Geneva, for instance, the government decided to move toward open source non-proprietary tools in education.
-All primary school- and most secondary school computers have been installed with the Ubuntu distribution of Linux. 
+For instance, in Geneva, Switzerland, the state government decided to move toward open source non-proprietary tools in education.
+All primary school and most secondary school computers have been installed with the Ubuntu distribution of Linux. 
 On this operating system the LEGO\textsuperscript{\textregistered} Mindstorms\textsuperscript{\textregistered} software tools are not available, and their proprietary nature is not compatible with the philosophy dictated by the government.
-Thymio was therefore extremely welcome, and the state of Geneva organized own training courses for teachers, ported several educational materials into their curriculum and organized a lending system for all schools that cannot afford to buy the robots.
-Others states are starting following this model.
+Thymio was therefore extremely welcome, and the state of Geneva organized ther own training courses for teachers, ported several educational materials into their curriculum and organized a lending system for all schools that cannot afford to buy the robots.
+%FIXME: SM: what do you mean by "ported several educational materials into their curriculum", which direction?
+%Others states are starting following this model.
 
-Finally the non-profit nature of the project, including the production infrastructure, has contributed to reduce the obstacle toward the acquisition of this platform for many schools, where teachers have a strong societal motivation and are reluctant to use tax-payers money to pay the divident of some shareholders.
+% FIXME: SM: in case of space limit, I would remove the following paragraph
+Finally, the non-profit nature of the project, including the production infrastructure, has contributed to reduce the obstacle toward the acquisition of this platform for many schools, where teachers have a strong societal motivation and are reluctant to use tax-payers money to pay the divident of some shareholders.
 
@@
 \includegraphics[width=\columnwidth]{figures/limitations}
-\caption{Limitations of open source hardware projects observed by participants to open source hardware projects.}
+\caption{Limitations of open source hardware projects.}
 \label{fig:problems}
@@
 The very specific structure of an open project has also drawbacks. 
-Asked about problems met in their project (see Figure~\ref{fig:problems}), the participants to our survey mention the management of the community, the evolution of the project, the efforts in documentation, production and distribution and the limited resources.
-There are here two kind of problems: one of management of people, the other in the development of the project after the first design.
+Asked about problems met in their project (Fig.~\ref{fig:problems}), the participants to our survey mention the management of the community, the evolution of the project, the efforts in documentation, production and distribution and the limited resources.
+There are here two kinds of problems: one of management of people, the other in the development of the project after the first design.
 The management of the project seems a recurrent problem and is related to the nature of this type of structure, mainly very flat and built on academic people with engineering background. 
 More than 62\% of the answers to our survey report about projects that are managed by the founder of the initiative and not by the most expert in the field or by a manager, which is perhaps not the best objective choice.
+% FIXME: SM: I do not totally agree, successful open source software project are heavily engineering-oriented, maybe it is interesting to highlight the effect of hardware here
 It is indeed very difficult to evolve from a phase where the engineering is the core of the project, to a new phase where communication and management play an increasing role.
@@
 Despite this approach, we had also problems of management of the community, and issues in finding the right long term vision, especially having so many partners. 
-The long term vision of a sales person is not the same of an engineer, and those confrontation result in conflicts in a community with a very flat hierarchy. 
+% FIXME: SM: I would highlight the problem of sustained funding
+The long term vision of a sales person is not the same of an engineer, and those confrontations result in conflicts. 
 A key element that allowed to keep cohesion in our community was to share, among all partners, strong basic values about improvement of society through education.
@@
 
+% FIXME: SM: it is not clear what you mean by "performances" in the following sentence
 The introduction of robots in formal education is a very challenging task, not only because of technical requirements such as low cost, performances and interactivity, but also because of factors depending on the school environment, such as the diversity of the educational programs, the dependence on local structures and languages, or the required training of teachers.
@@
 Because they do not consider all elements at stake, these tools fail to meet some challenges of the field.
+% FIXME: SM: I do not totally agree, Mindstorms was very successful in schools in the US, our paper is kind of weak with respect to such work
 On the contrary, in this paper we presented a global approach that tackles these questions in a holistic way, based on open source and non-profit principles.
@@
 By making a survey among a panel of people contributing to open source hardware projects, we could observe that our project shares some characteristics with the majority of the projects represented in the survey.
-We also had problems of management of the community, we chose production and distributions methods that are broadly applied, and we share most of the motivation elements that are behind other projects.
-We identified an underestimated legal issue for open source hardware projects in the licensing term of CAD software.
+We also had problems of management of the community, we chose production and distribution methods that are broadly applied, and we share most of the motivational elements that are behind other projects.
+We identified an underestimated legal issue for open source hardware projects in the licensing term of \textsc{cad} software.
 Indeed only very few editors have licenses allowing the publication of source files created with educational license.
-Two surveys among the users and the CAD editors shows that both people contributing or leading open source hardware projects and the editors themselves are now aware of this issue.
-Finally we could show some elements that differentiate our project from other open source hardware ones, being specific to educational robotics.
-In particular our project takes advantage from an alignment between the principles underlying open source project and the nature of education institutions. 
-This fits with a political trend toward open source in education and in science in general.
-We also found a solution to the struggling question of production methods by splitting our hardware in two categories, enabling both very advanced hardware and inclusion of a broad community of end users.
+Two surveys among the users and the \textsc{cad} editors show that both people contributing or leading open source hardware projects and the editors themselves are not aware of this issue.
+Finally, we could show some elements, specific to educational robotics, that differentiate our project from other open source hardware projects.
+In particular, our project takes advantage of an alignment between the principles underlying open source and the nature of education institutions. 
+This fits a political trend toward open source in education and in science in general.
+We also found a solution to the problem of production methods by splitting our hardware in two categories, enabling both advanced technology and the inclusion of a broad community of end users.
+% FIXME: SM: although I see what you want to say, I feel that the gender issue falls a bit from the sky here
 This addresses durability of equipment and gender issues that are critical in education.
@@
 %
-%\section*{ACKNOWLEDGMENT}
+%\section*{ACKNOW\textsc{led}GMENT}
 %

### 2016-04-13 paper_ram.tex
Corrected all FIXME, still need to agree on paragraph on problems of open source.

@@
 By \textit{open source hardware} we mean, following the definition of the Open Source Hardware Association\footnote{\url{http://www.oshwa.org/definition/}}, ``hardware whose design is made publicly available so that anyone can study, modify, distribute, make, and sell the design or hardware based on that design''.
-In this paper, we show that this concept, implemented as a project gathering a community of users, developers and manufacturers, brings a strong added value to the robot and to the educational methods.
-This is illustrated by the results achieved with Thymio and is compared with the feedback of a group of robotics engineers active in open source hardware design.
-% FIXME: SM: I do not understand this link, why do we compare with other open source hardware products, as it seems that until now the arguing is against closed source products?
-We can well observe both similarities and differences between our education-centered project and other open source robotic hardware projects, pointing to challenges and opportunities.
-More details on the robot and the related results can be found in~\cite{RiedoPhD, magnenat2014}.
+In this paper, we show that this concept, implemented in the Thymio project gathering a community of users, developers and manufacturers, brings a strong added value to the robot and to the educational methods.
+In addition, we compare our experience with other robotics open source hardware projects not focused on education, and highlight challenges and opportunities specific to education.
 
@@
 There is a large number of publications presenting educational robots with different prices and features, from low-cost systems for instance gtargeting education in Africa~\cite{Rubenstein2015,Gyebi2015} to extremely sophisticated humanoids~\cite{Hood2015,Mazzoni2016}.
-Among those, only a handful are commercially available, limiting their validation by educational scientists.
-This results in 90\,\% of publications about validation of educational results made on LEGO\textsuperscript{\textregistered} Mindstorms\textsuperscript{\textregistered}~\cite{benitti2012exploring}.
-% FIXME: SM: is that in general or only about STEM/programming?
-The LEGO\textsuperscript{\textregistered} Mindstorms\textsuperscript{\textregistered} EV3\footnote{\url{http://mindstorms.lego.com}} is expensive ($\approx$ 400\,\$) but offers a wide range of possibilities, especially at the mechanical level using LEGO\textsuperscript{\textregistered} bricks and at the software level with its graphical programming environment.
-At a lower price, LEGO\textsuperscript{\textregistered} proposes the WeDo\textsuperscript{\textregistered} system ($\approx$ 80\,\$).
-The WeDo\textsuperscript{\textregistered} is similar to EV3 but only reads one sensor and controls one actuator and must always be connected to a desktop computer.
-% FIXME: SM: The WeDo falls a bit from the sky here, although I see the underlying idea...
-The BeeBot robot\footnote{\url{http://www.beebot.org.uk}} is a cheap ($\approx$ 90\,\$) mobile system.
-However, it is not a real robot but rather an automaton: There is no perception of the environment, just a good encoder for precise movements and a simple programming interface based on buttons giving directions for the movements.
+Among those, only a handful are commercially available, limiting their validation by educational scientists, who are typically not roboticists.
+This results in 90\,\% of publications about validation of educational results made on LEGO\textsuperscript{\textregistered} Mindstorms\textsuperscript{\textregistered}~\cite{benitti2012exploring}, a widely available commercial product.
+The latest version, EV3\footnote{\url{http://mindstorms.lego.com}}, is expensive ($\approx$ 400\,\$) but offers a wide range of possibilities, especially at the mechanical level using LEGO\textsuperscript{\textregistered} bricks and at the software level with its graphical programming environment.
+%NOTE: SM: commented WeDo to save space and streamline story
+%At a lower price, LEGO\textsuperscript{\textregistered} proposes the WeDo\textsuperscript{\textregistered} system ($\approx$ 80\,\$).
+%The WeDo\textsuperscript{\textregistered} is similar to EV3 but only reads one sensor and controls one actuator and must always be connected to a desktop computer.
+In comparison, the BeeBot robot\footnote{\url{http://www.beebot.org.uk}} is a cheap ($\approx$ 90\,\$) mobile system.
+However, it is not a real robot but rather an automaton: There is no perception of the environment, just a good encoder for precise movements and a simple programming interface based on buttons giving directions for the movement.
 
-Among these widely used robots, none is part of an open source hardware project.
-%The only move in this direction comes from LEGO\textsuperscript{\textregistered}, that made public all the schematics of the EV3 in an hardware developer kit, to allow the design of extensions.
-% FIXME: if the LEGO license is not a true open hardware one, I would drop the previous sentence, it does not bring much
+Among widely used robots, none is part of an open source hardware project.
 In niche projects, only few open source hardware educational robots are used in schools:
@@
 These are probably two important factors for its success in schools~\cite{2008highfield,2008demichele} and the availability of a rich set of educational material\footnote{\url{https://www.learningplace.com.au/deliver/content.asp?pid=38840} and \\ \url{http://beebots.skola.edu.mt/}}.
-However, the total lack of sensors reduces the BeeBot to an automaton, easy to use but limited to very simple activities for young kids, without much potential in teaching computational thinking or robotics concepts.
-% FIXME: SM: I would even say that the BeeBot targets a specfic concept for kids, which is the learning of sequences, and is very good at it.
-I think we could find a strong educational argument to show how limited it is.
-maybe Morgane can help.
+However, the total lack of sensors reduces the BeeBot to an automaton, easy to use but limited to activities implementable by sequences of motions, without any closed-loop control that would allow teaching computational thinking or robotics concepts.
 
 For more complex robots, both the development of educational material and the training of teachers require a huge work, demanding a good mix between robotics and educational skills. 
-Moreover, educational material changes very much from school to school, as education is very dependent from local educational programs and languages, for instance. 
-In the particular case of Switzerland, for instance, every of the 26 states composing the country has its own school program and its own language among the four national ones. 
+Moreover, educational material varies from school to school, for instance requirements are very dependent on local educational programs and languages. 
+For example, in the particular case of Switzerland, every of the 26 states composing the country has its own school program and its own language among the four national ones. 
 Educational material and the related training must therefore be adapted to each situation.
@@
 This model is an extension to education of what is already practiced in projects such as Arduino~\cite{Jamieson2010}.
-
-% FIXME: SM: I am lacking a link, what is the contribution of the paper in that context?
+In htis paper, we study a case of implementation of this model.
 
@@
 	\includegraphics[width=.8\columnwidth]{figures/front_2016}
-	%FIXME: SM: maybe we should have a ruler on the side to show the scale?
-	\vskip 1.5em
+	\vskip 1.0em
 	\includegraphics[width=\columnwidth]{figures/wirelessThymioVecto}
@@
 The robot embeds a battery, rechargable by \textsc{usb}, that provides an autonomy of about 2 hours.
-
-%We start by presenting here the first six main axes, before moving to the next section devoted the open hardware specificities.
-% FIXME: SM: what do you mean by specificities here?
+More details on the robot and the previous research results can be found in~\cite{RiedoPhD, magnenat2014}.
 
@@
 These basic behaviors allow people starting Thymio to immediately interact with it, while illustrating the many possibilities of the robot.
-The behaviors are interactive and oriented toward the user. % SM: FIXME: not clear what "oriented toward the user" means
+The behaviors are interactive.
 For each behavior, the robot's body has a different color, allowing the user to easily recognize it.
@@
 
-% FIXME: SM: is there a reason to limit images to 90% width?
 \begin{figure}
 \centering
-\includegraphics[width=.9\columnwidth]{figures/vpl}
+\includegraphics[width=\columnwidth]{figures/vpl}
 \caption{The visual programming language.}
@@
 \centering
-\includegraphics[width=.9\columnwidth]{figures/blockly}
+\includegraphics[width=\columnwidth]{figures/blockly}
 \caption{The Blockly programming environment.}
@@
 \centering
-\includegraphics[width=.9\columnwidth]{figures/aseba-studio}
+\includegraphics[width=\columnwidth]{figures/aseba-studio}
 \caption{Aseba Studio, the integrated development environment}
@@
 Both were targeting similar users and have been sold in similar quantities.
-What we can observe after more than 10 years is that Khepera generated many royalties for the university, but less relative academic visibility than e-puck (Fig.~\ref{fig:khepuck}).
-% FIXME: SM: "many royalties" is unclear
+What we can observe after more than 10 years is that Khepera generated royalties for the university, but less relative academic visibility than e-puck (Fig.~\ref{fig:khepuck}).
 The e-puck robot, being open hardware and with an image better linked to the university, generated no income for the university but much more visibility.
@@
 Therefore the link with the professional activity and the particular project is evident.
-On another side, the motivation of contributing to society is much stronger in our team, as developing a robot targeting education has a strong societal component.
-Moreover this project has a strong scientific motivation, with several ongoing studies on the acceptance by teachers and the impact on children.
-Also this motivation is therefore stronger in our team.
-% FIXME: SM: duplication about "our team"
-Although this seems a minor issue, sharing a strong fundamental motivation such as education or scientific achievements, is a key element for a strong community~\cite{Stahlbrost2011}, especially if interdisciplinary like ours.
+On another side, the motivation of contributing to society is strong in our team, as developing a robot targeting education has an important societal component.
+Moreover, our project also has a strong scientific motivation, with several ongoing studies on the acceptance by teachers and the effect on children's learning.
+Therefore, sharing a strong fundamental motivation such as education or scientific achievements, is a key element for building a solid community~\cite{Stahlbrost2011}, especially if it is interdisciplinary like ours.
 
@@
 Among the fundamental choices when starting an open source hardware project, there is the choice of the type of production. 
-In the definition of open hardware we gave, it is stated that one should offer an ``hardware whose design is made publicly available so that anyone can ... make... the ... hardware ''.
-% FIXME: SM: I do not understand the dots
+In the definition of open hardware we gave, it is stated that one should offer an ``hardware whose design is made publicly available so that anyone can [\dots] make [\dots] hardware ''.
 Behind the ``anyone'' should we consider every single person or only companies able to produce the product?
@@
 
-% FIXME: SM: in case of lack of space, we can strip this paragraph away
-Another key element in supporting repair by the users are the calibration methods. 
-When choosing very low cost components, one gets into large dispersion of the characteristics. 
-In our case, for instance, the right and left wheel motors can have very different electrical characteristics, resulting in different speed evaluations and therefore in the robot not going straight when the two motors regulators have the same set speed.
-To correct this problem, one interesting solution is to introduce factory calibration, storing correction factors that are specific for each robot, in the microcontroller memory. 
-To allow the user to replace a broken motor, it is essential to give her also the possibility to re-calibrate the robot and adjust the parameters of the new motors.
+Another key element in supporting repairs by end users is the documentation of calibration methods. 
+When choosing very low cost components, one faces large dispersion of characteristics. 
+For example, in the Thymio the right and left wheel motors can differ in their electrical characteristics, resulting in the robot not going straight for similar speed commands to both wheels.
+To correct this problem, we introduced factory calibration, storing in the microcontroller memory correction factors that are specific for each robot. 
+To allow the user to replace a broken motor, it is essential to give her also the possibility to re-calibrate the robot and adjust the parameters of the new motor.
 In Thymio, this results in the design and the documentation of calibration processes that can be performed by anyone, getting close to the original definition of open source hardware.
@@
 On this operating system the LEGO\textsuperscript{\textregistered} Mindstorms\textsuperscript{\textregistered} software tools are not available, and their proprietary nature is not compatible with the philosophy dictated by the government.
-Thymio was therefore extremely welcome, and the state of Geneva organized ther own training courses for teachers, ported several educational materials into their curriculum and organized a lending system for all schools that cannot afford to buy the robots.
-%FIXME: SM: what do you mean by "ported several educational materials into their curriculum", which direction?
-%Others states are starting following this model.
+Thymio was therefore extremely welcome, and the state of Geneva organized their own training courses for teachers, adapted existing educational materials to their own curriculum and organized a lending system for all schools that cannot afford to buy the robots.
 
 % FIXME: SM: in case of space limit, I would remove the following paragraph
-Finally, the non-profit nature of the project, including the production infrastructure, has contributed to reduce the obstacle toward the acquisition of this platform for many schools, where teachers have a strong societal motivation and are reluctant to use tax-payers money to pay the divident of some shareholders.
+%Finally, the non-profit nature of the project, including the production infrastructure, has contributed to reduce the obstacle toward the acquisition of this platform for many schools, where teachers have a strong societal motivation and are reluctant to use tax-payers money to pay the divident of some shareholders.
 
@@
 More than 62\% of the answers to our survey report about projects that are managed by the founder of the initiative and not by the most expert in the field or by a manager, which is perhaps not the best objective choice.
-% FIXME: SM: I do not totally agree, successful open source software project are heavily engineering-oriented, maybe it is interesting to highlight the effect of hardware here
 It is indeed very difficult to evolve from a phase where the engineering is the core of the project, to a new phase where communication and management play an increasing role.
+This problem is deeply connected with the one of acquiring sustained funding for the project, in order to build a professional team able to support the product.
 The same holds for several other factors, like the efforts in documentation, production and distribution of the product. 
@@
 
-In the Thymio project we had very similar issues and tried to address them by having several partners, each of them specialized in one aspect: design, production and sale, creation of educational material, etc.
-Despite this approach, we had also problems of management of the community, and issues in finding the right long term vision, especially having so many partners. 
-% FIXME: SM: I would highlight the problem of sustained funding
-The long term vision of a sales person is not the same of an engineer, and those confrontations result in conflicts. 
+In the Thymio project we faced similar issues and tried to address them by having several partners, each of them specialized in one aspect: design, production and sale, creation of educational material, etc.
+Despite this approach, we also encountered difficulties in nurturing an active community covering all aspects of the project, and issues in finding a long term vision that both injects life into the project and is financially sustainable.
+This is especially difficult with a large number of partners, as differing views can result in conflicts. 
 A key element that allowed to keep cohesion in our community was to share, among all partners, strong basic values about improvement of society through education.
@@
 
-% FIXME: SM: it is not clear what you mean by "performances" in the following sentence
-The introduction of robots in formal education is a very challenging task, not only because of technical requirements such as low cost, performances and interactivity, but also because of factors depending on the school environment, such as the diversity of the educational programs, the dependence on local structures and languages, or the required training of teachers.
+The introduction of robots in formal education is a very challenging task, not only because of technical requirements such as low cost and interactivity, but also because of factors depending on the school environment, such as the diversity of the educational programs, the dependence on local structures and languages, or the required training of teachers.
 Most of the current robotic tools for education capitalize upon existing hardware platforms, programming environments, or robot architectures.
-Because they do not consider all elements at stake, these tools fail to meet some challenges of the field.
-% FIXME: SM: I do not totally agree, Mindstorms was very successful in schools in the US, our paper is kind of weak with respect to such work
+Because they do not consider all elements at stake, these tools fail to meet some challenges of formal education.
 On the contrary, in this paper we presented a global approach that tackles these questions in a holistic way, based on open source and non-profit principles.
@@
 
-By making a survey among a panel of people contributing to open source hardware projects, we could observe that our project shares some characteristics with the majority of the projects represented in the survey.
-We also had problems of management of the community, we chose production and distribution methods that are broadly applied, and we share most of the motivational elements that are behind other projects.
+By making a survey among contributors to open source hardware projects, we could observe that our project shares some characteristics with the majority of the projects represented in the survey. 
+We also faced difficulties in establishing the community, we chose production and distribution methods that are broadly applied, and we share most of the motivational elements that are behind other projects.
 We identified an underestimated legal issue for open source hardware projects in the licensing term of \textsc{cad} software.
@@
 This fits a political trend toward open source in education and in science in general.
-We also found a solution to the problem of production methods by splitting our hardware in two categories, enabling both advanced technology and the inclusion of a broad community of end users.
-% FIXME: SM: although I see what you want to say, I feel that the gender issue falls a bit from the sky here
-This addresses durability of equipment and gender issues that are critical in education.
+We also found a solution to the problem of production methods by splitting our hardware in two categories, enabling both advanced technology for the robot and a large variety of accessories.
+Hence, Thymio appeals to a broad community of end users, which addresses durability of equipment and gender issues that are critical in education.
 

### 2016-04-14 paper_ram.tex
Attempt to improve section on problems of opne-source hardware

@@
 
-\subsection{Problems of open-source projects}
+\subsection{Problems of open-source hardware projects}
 
@@
 
-The very specific structure of an open project has also drawbacks. 
+The very specific structure of an open-source hardware project has also drawbacks. 
 Asked about problems met in their project (Fig.~\ref{fig:problems}), the participants to our survey mention the management of the community, the evolution of the project, the efforts in documentation, production and distribution and the limited resources.
-There are here two kinds of problems: one of management of people, the other in the development of the project after the first design.
-The management of the project seems a recurrent problem and is related to the nature of this type of structure, mainly very flat and built on academic people with engineering background. 
-More than 62\% of the answers to our survey report about projects that are managed by the founder of the initiative and not by the most expert in the field or by a manager, which is perhaps not the best objective choice.
-It is indeed very difficult to evolve from a phase where the engineering is the core of the project, to a new phase where communication and management play an increasing role.
-This problem is deeply connected with the one of acquiring sustained funding for the project, in order to build a professional team able to support the product.
-The same holds for several other factors, like the efforts in documentation, production and distribution of the product. 
-Also here there is a change of activity from the very creative and appealing design phase, to a phase of documentation and, in some cases, production and distribution which is not at all the same type of activity and requires different skills.
-This shift requires extra efforts and motivation, which are often missing, especially for activities like documentation that are not appreciated by engineers.
-In general many respondents to our survey identified the evolution of the project as a general problematic issue, which summarizes the several aspects mentioned above.
+These problems are mainly caused by the mismatch between the conditions of the initial phase of the project, and the needs of long term sustainability.
+Initially, these projects are created by academic people with engineering background, typically in a flat structure.
+However, long term sustainability requirements are closer to the ones of a successful startup, and involve communication, marketing, and professional management in order to acquire sustained funding.
+Yet, 62\% of the participants to our survey report projects that are managed by the founder of the initiative, who might not be the best person to lead such a structure.
+This differs from open-source software, because initiating hardware production requires the investment of a large amount of money, and modifying hardware is much slower than modifying software.
+Moreover, in the case of robotics, although software plays a critical part, the perception of the product is highly centered on the hardware.
+Therefore, in open hardware projects, there is a structural conflict between the open-source ideals and the concrete financial and logistic constraints.
+
+% There are here two kinds of problems: one of management of people, the other in the development of the project after the first design.
+% The management of the project seems a recurrent problem and is related to the nature of this type of structure, mainly very flat and built on academic people with engineering background. 
+% More than 62\% of the answers to our survey report about projects that are managed by the founder of the initiative and not by the most expert in the field or by a manager, which is perhaps not the best objective choice.
+% It is indeed very difficult to evolve from a phase where the engineering is the core of the project, to a new phase where communication and management play an increasing role.
+% This problem is deeply connected with the one of acquiring sustained funding for the project, in order to build a professional team able to support the product.
+% The same holds for several other factors, like the efforts in documentation, production and distribution of the product. 
+% Also here there is a change of activity from the very creative and appealing design phase, to a phase of documentation and, in some cases, production and distribution which is not at all the same type of activity and requires different skills.
+% This shift requires extra efforts and motivation, which are often missing, especially for activities like documentation that are not appreciated by engineers.
+% In general many respondents to our survey identified the evolution of the project as a general problematic issue, which summarizes the several aspects mentioned above.
 

### 2016-04-14 paper_ram.tex
Finished edition of problems of open-source hardware projects

@@
 Yet, 62\% of the participants to our survey report projects that are managed by the founder of the initiative, who might not be the best person to lead such a structure.
-This differs from open-source software, because initiating hardware production requires the investment of a large amount of money, and modifying hardware is much slower than modifying software.
-Moreover, in the case of robotics, although software plays a critical part, the perception of the product is highly centered on the hardware.
-Therefore, in open hardware projects, there is a structural conflict between the open-source ideals and the concrete financial and logistic constraints.
+This differs from open-source software, because in our model initiating hardware production requires the investment of a large amount of money, and modifying hardware is much slower than modifying software.
+Moreover, in the case of robotics, although software plays a critical part, the perception of the product is highly centered on the hardware, creating tensions between hardware- and software-oriented people.
+Therefore, to grow such open hardware projects, there is a structural conflict between open-source ideals (for example meritocratic decision process) and the concrete financial and logistic constraints.
 
@@
 
-In the Thymio project we faced similar issues and tried to address them by having several partners, each of them specialized in one aspect: design, production and sale, creation of educational material, etc.
+In the Thymio project we have addressed these problems by having several partners, each of them specialized in one aspect: design, production and sale, creation of educational material, etc.
 Despite this approach, we also encountered difficulties in nurturing an active community covering all aspects of the project, and issues in finding a long term vision that both injects life into the project and is financially sustainable.

### 2016-10-18 paper_ram.tex
radical cut

@@
 L\'ea Pereyre$^{1}$, Philippe R\'etornaz$^{1}$ and St\'ephane Magnenat$^{1}$% <-this % stops a space
-\thanks{*This research was partially supported by the Swiss National Center of Competence in Research ``Robotics'' (Thymio robot development and deployment), partially by GebertRuf Stiftung (design of accessories), partially by the Swiss NSF project CRAGP2 151543 ``Robotics in schools'', and partially by the EU-FP7 project ASSISIbf, no.
+\thanks{*This research was supported by the Swiss National Center of Competence in Research ``Robotics'' (Thymio robot development and deployment), by GebertRuf Stiftung (design of accessories), by the Swiss NSF project CRAGP2 151543 ``Robotics in schools'', and by the EU-FP7 project ASSISIbf, no.
 601074 (survey on open hardware).
-Many thank to Luc Bergeron and his team at écal.ch for the industrial design of Thymio; Didier Roy, David Sherman and all the INRIA team for their contributions and diffusion in France; Gordana Gerber for the educational material; all the Mobsya team for the effort in production and sale; and Fabian Hahn for the port of Blockly.}% <-this % stops a space``anyone''
+Many thank to Luc Bergeron and his team at \'ecal.ch for the industrial design of Thymio; Didier Roy, David Sherman and all the INRIA team for their contributions and diffusion in France; Gordana Gerber for the educational material; all the Mobsya team for the effort in production and sale; and Fabian Hahn for the port of Blockly.}% <-this % stops a space``anyone''
 \thanks{$^{1}$Francesco Mondada, Manon Briod, L\'ea Pereyre, Philippe R\'etornaz and St\'ephane Magnenat are with the Laboratoire de Syst\`emes Robotiques,
@@
         {\tt\small firstname.lastname@epfl.ch}}%
-\thanks{$^{2}$Michael Bonani and Fanny Riedo are with Mobsya Association, Ecublens, Switzerland
+\thanks{$^{2}$Michael Bonani and Fanny Riedo are with Mobsya Association, Renens, Switzerland
         {\tt\small michael.bonani@mobsya.org} and {\tt\small fanny.riedo@mobsya.org}}%
@@
 Mobile robots are valuable tools for education because of both the enthusiasm they raise and the multidisciplinary nature of their technology.
-Because autonomous mobile robots sense the environment and take actions based on their perception, they seem to display \emph{intentions of their own}.
-This fascinates the users of the robot and creates feelings of accomplishment and power when these users engage in creative activities with the robots.
-The permeating presence of robots in science fiction and their projected use in our society increases these emotions by giving a sense of touching the future.
-Children can spend hours looking at a robot interacting with its environment and adults can approach robots as a new hobby that replaces radio and electronics kits.
-Robots embed various technologies and therefore give access to a wide range of fields, such as complex mechanics, sensors, wireless transmission, mathematics, computer science, etc.
-The emotional potential of mobile robots and their varied technology makes them a potential ideal tool for education.
+They give access to a wide range of fields, such as complex mechanics, sensors, wireless transmission, mathematics, computer science, etc.
+As mobile robots sense the environment and take actions based on their perception, they seem to display \emph{intentions of their own}~\cite{Mondada2001}.
+%This fascinates the users of the robot and creates feelings of accomplishment and power when these users engage in creative activities with the robots.
+This impression of intelligent technology, the permeating presence of robots in science fiction and their projected use in our society give a sense of touching the future.
+%Children can spend hours looking at a robot interacting with its environment and adults can approach robots as a new hobby that replaces radio and electronics kits.
+%The emotional potential of mobile robots and their varied technology makes them a potential ideal tool for education.
 
-However, robots are still not as widespread in schools as they could be, for several reasons:
+Despite their potential as educational tools, robots are still not as widespread in schools as they could be. Among the possible reasons, we believe that the following five are playing a key role:
 \begin{enumerate}
-\item Although there are many robotics projects developing innovative and interesting educational robots, few reach a sufficient maturity to become distributed and accessible to schools. 
-\item A robot performing interesting behaviors is a complex piece of technology and therefore expensive. 
-Indeed, for having an educational value and providing an interesting level of interaction, a robot must embed a wide set of sensors and actuators.
-Existing platforms with these features cost several hundred Euros.
-This prevents most schools, which have a limited budget for equipment, to acquire interesting robots.
-\item Introducing robotic tools into teaching activities requires investment in time and training for the teachers.
-In Europe, despite a trend to increase the use of technological tools in the learning process, teachers are still insufficiently trained and are reluctant to introduce these tools in their teaching activity~\cite{CERI2008}.
-Therefore, to be accepted by teachers, robots must be both accessible with minimal effort and accompanied by well-prepared educational material. 
-Moreover, infrastructure should be provided to allow the sharing of educational material among teachers.
+\item Although there are many research projects developing innovative and interesting educational robots, few reach a sufficient maturity to become distributed and accessible to schools. 
+\item A versatile robot performing interesting behaviors is a complex piece of technology and therefore expensive. 
+%Indeed, for having an educational value and providing an interesting level of interaction, a robot must embed a wide set of sensors and actuators.
+%Existing platforms with these features cost several hundred Euros.
+This prevents most schools, which have a limited budget for equipment, to acquire educational robots.
+\item Introducing robotic tools into teaching activities requires investment in time and training for the teachers~\cite{CERI2008}.
+%In Europe, despite a trend to increase the use of technological tools in the learning process, teachers are still insufficiently trained and are reluctant to introduce these tools in their teaching activity~\cite{CERI2008}.
+Therefore, to be accepted by teachers, robots must be both accessible with minimal effort and accompanied by well-prepared educational material shared among colleagues. 
+%Moreover, infrastructure should be provided to allow the sharing of educational material among teachers.
 \item Robot construction, use, and programming is often perceived as a boyish activity in our society~\cite{leonard2009lego,nourbakhsh2009robot}.
@@
 \item Finally, many teachers are reluctant to follow volatile trends, especially if these are based on purely commercial arguments.
-Teachers prefer to invest in stable tools, and this is the opposite as the trend of current consumer technology, which is often very volatile and based on continuous renewal rather than on durability.
+Teachers prefer to invest in stable tools, and this is the opposite as the trend of current consumer technology.
+%, which is often very volatile and based on continuous renewal rather than on durability.
 \end{enumerate}
@@
 By \textit{open source hardware} we mean, following the definition of the Open Source Hardware Association\footnote{\url{http://www.oshwa.org/definition/}}, ``hardware whose design is made publicly available so that anyone can study, modify, distribute, make, and sell the design or hardware based on that design''.
-In this paper, we show that this concept, implemented in the Thymio project gathering a community of users, developers and manufacturers, brings a strong added value to the robot and to the educational methods.
+In this paper, we show that this concept, implemented in the Thymio project and gathering a community of users, developers and manufacturers, brings a strong added value to the robot and to the educational methods.
 In addition, we compare our experience with other robotics open source hardware projects not focused on education, and highlight challenges and opportunities specific to education.
@@
 
-There is a large number of publications presenting educational robots with different prices and features, from low-cost systems for instance gtargeting education in Africa~\cite{Rubenstein2015,Gyebi2015} to extremely sophisticated humanoids~\cite{Hood2015,Mazzoni2016}.
+There is a large number of publications presenting educational robots, from low-cost systems targeting Africa~\cite{Rubenstein2015,Gyebi2015} to extremely sophisticated humanoids~\cite{Hood2015,Mazzoni2016}.
 Among those, only a handful are commercially available, limiting their validation by educational scientists, who are typically not roboticists.
@@
 %The WeDo\textsuperscript{\textregistered} is similar to EV3 but only reads one sensor and controls one actuator and must always be connected to a desktop computer.
-In comparison, the BeeBot robot\footnote{\url{http://www.beebot.org.uk}} is a cheap ($\approx$ 90\,\$) mobile system.
-However, it is not a real robot but rather an automaton: There is no perception of the environment, just a good encoder for precise movements and a simple programming interface based on buttons giving directions for the movement.
+%In comparison, the BeeBot robot\footnote{\url{http://www.beebot.org.uk}} is a cheap ($\approx$ 90\,\$) mobile system.
+%However, it is not a real robot but rather an automaton: There is no perception of the environment, just a good encoder for precise movements and a simple programming interface based on buttons giving directions for the movement.
+Among the other players on the market, one can cite the Edison robot\footnote{\url{https://meetedison.com/}}, extremely low cost (49\,\$) and compatible with LEGO\textsuperscript{\textregistered} bricks.
+The low price has pushed to drastic design choices; small size, weak motors, few sensors, communication with the computer by audio jack and standard AAA batteries not included in the price.
+These choices strongly limits its possible usage.
+%Among widely used robots, none is part of an open source hardware project.
 
-Among widely used robots, none is part of an open source hardware project.
-In niche projects, only few open source hardware educational robots are used in schools:
+Among the robots available on the market, only few are open source and are used in schools:
 Scribbler2, produced and sold by Parallax\footnote{\url{http://www.parallax.com}} ($\approx$ 180\,\$), is a 188\,mm large robot, designed to run on the ground and equipped with few light sensors, one distance sensor, two ground sensors and few \textsc{led} displays.
@@
 It is programmable with a graphical or a textual code interface.  
-The main weakness of Scribbler2 is its performance per price ratio and the limited compatibility with other systems. 
+The main weakness of Scribbler2 is its limited number of sensors and the limited compatibility with other systems. 
 Moreover, there seems to be no active community around its development.
@@
 Several simulators allow to run experiments of high complexity. 
-There is an active community of users and developers around it.
+%There is an active community of users and developers around it.
 Its main weakness is its high price ($\approx$ 870\,\$).
@@
 
-With respect to these robots, Thymio has a compact size (120~mm), many interaction possibilities, an affordable price and a large set of sensors.
-Its design makes the best use of the recent cheap components coming from the explosive growth of the mobile devices market. 
+With respect to these robots, Thymio has a compact size (110\,mm), many interaction possibilities, an affordable price and a large set of sensors.
+%Its design makes the best use of the recent cheap components coming from the explosive growth of the mobile devices market. 
 To the best of our knowledge, beside Thymio there are no educational products providing a similar integration of sensors and actuators at a price below 130\,\$.
 
-Teachers are one of the main target user group of the Thymio project.
-For teachers, the motivation to use robotic tools in their classes depends on many factors~\cite{chevalier2016}.
+Teachers, from primary school up to high school, are one of the main target user group of the Thymio project.
+They decide which tools are used in their class and are key people in the eduction ecosystem.
+For teachers, the motivation to use robotic tools depends on many factors~\cite{chevalier2016}.
 Among them, the availability of material and training plays a key role.
-The BeeBot mentioned previously, because of its radical simplicity, is one of the tools that needs the least training. 
-Moreover its use in learning activities is based on a simple grid on a printed map on the ground, allowing fast and easy integration in existing lessons.
-These are probably two important factors for its success in schools~\cite{2008highfield,2008demichele} and the availability of a rich set of educational material\footnote{\url{https://www.learningplace.com.au/deliver/content.asp?pid=38840} and \\ \url{http://beebots.skola.edu.mt/}}.
-However, the total lack of sensors reduces the BeeBot to an automaton, easy to use but limited to activities implementable by sequences of motions, without any closed-loop control that would allow teaching computational thinking or robotics concepts.
-
-For more complex robots, both the development of educational material and the training of teachers require a huge work, demanding a good mix between robotics and educational skills. 
-Moreover, educational material varies from school to school, for instance requirements are very dependent on local educational programs and languages. 
-For example, in the particular case of Switzerland, every of the 26 states composing the country has its own school program and its own language among the four national ones. 
-Educational material and the related training must therefore be adapted to each situation.
+%The BeeBot mentioned previously, because of its radical simplicity, is one of the tools that needs the least training. 
+%Moreover its use in learning activities is based on a simple grid on a printed map on the ground, allowing fast and easy integration in existing lessons.
+%These are probably two important factors for its success in schools~\cite{2008highfield,2008demichele} and the availability of a rich set of educational material\footnote{\url{https://www.learningplace.com.au/deliver/content.asp?pid=38840} and \\ \url{http://beebots.skola.edu.mt/}}.
+%However, the total lack of sensors reduces the BeeBot to an automaton, easy to use but limited to activities implementable by sequences of motions, without any closed-loop control that would allow teaching computational thinking or robotics concepts.
+The development of educational material and courses to train the teachers requires a huge effort, based on a good mix between robotics and educational skills. 
+Moreover, educational material varies from school to school, as requirements are very dependent on local educational programs and languages. 
+%For example, in Switzerland, every of the 26 states composing the country has its own school program and its own language among the four national ones. 
+%Educational material and the related training must therefore be adapted to each situation.
 Even for large companies like LEGO\textsuperscript{\textregistered}, such a development is not sustainable. 
@@
 An open source community regrouping developers, manufacturers and end users is therefore a very interesting model to address the distributed development and sharing of educational material, and the diffusion of training sessions.
-This model is an extension to education of what is already practiced in projects such as Arduino~\cite{Jamieson2010}.
-In htis paper, we study a case of implementation of this model.
+%This model is an extension to education of what is already practiced in projects such as Arduino~\cite{Jamieson2010}.
+In this paper, we study a case of implementation of this model.
 
@@
 
-We designed the Thymio robot along seven main axes: a low price to address a larger number of users, a feature set that suits both genders and multiple ages from young children to adults, a combination of sensors, actuators, and programming features that facilitates learning, a mechanical design that allows creative extensions by the users themselves, a set of ready-to-use behaviors to get a demonstration right out of the box, an accessible programming environment, and an open source community contributing to the design and the dissemination.
-The result is a miniature (10\,cm in side) differential-wheeled robot suited for use on a desktop (Fig.~\ref{fig:schematique}, top).
+We designed the Thymio robot along seven main axes: a low price to address a larger number of users; a feature set that suits both genders and multiple ages from young children to adults; a mechanical design that promotes creativity; a combination of sensors, actuators, and programming features that facilitates learning: a set of ready-to-use programs to get fast access to robotic behaviors; an accessible programming environment; and an open source community contributing to the design and the dissemination.
+The result is a miniature (110\,mm in side) differential-wheeled robot suited for use on a desktop (Fig.~\ref{fig:schematique}, top).
 The robot is robust enough to be mishandled by children, as it can fall from a table without breaking itself.
 It features a translucent white hull and a wide range of sensors and actuators (Fig.~\ref{fig:schematique}, bottom).
-The robot embeds a battery, rechargable by \textsc{usb}, that provides an autonomy of about 2 hours.
+The robot embeds a battery, rechargeable by \textsc{usb}, that provides an autonomy of about 3 to 5 hours.
 More details on the robot and the previous research results can be found in~\cite{RiedoPhD, magnenat2014}.
@@
 
-Our main concern during the mechatronic design of Thymio was to keep a low production cost. 
-Price is a key factor for the adoption of the robot by schools~\cite{kradolfer2014sociological}.
-In our previous experience with designing robots, most of the costs were due to the electronics and the sensors~\cite{thesis_michael}, not to the mechanical parts.
-As we wanted Thymio to be an affordable robot, we selected electronic components with great care and explored different design solutions in order to limit the number of expensive components.
-In particular, we focused on low-cost sensors that allow rich interaction possibilities, with both the environment and the user.
-The resulting Thymio robot possesses a large number of sensors:
-Seven horizontal infrared distance sensors, with a range of about 13\,cm.
-They enable many interactions with the environment, like following an object or avoiding an obstacle.
-Two ground infrared sensors allow to follow a line or to detect the edge of a table.
-A three-axis accelerometer measures the orientation of the robot with respect to gravity, and can detect shocks and free falls.
+Price is key in the adoption of robots by schools~\cite{kradolfer2014sociological}.
+Therefore the design of Thymio had to target a low production cost, while including a broad range of functionalities to enable flexibility in its use. 
+As in this type of robot the main cost comes from electronics and sensors~\cite{thesis_michael}, we focused on low-cost sensors that allow rich interaction possibilities, with both the environment and the user.
+
+The resulting Thymio robot possesses a large number of simple sensors:
+Seven horizontal proximity sensors, two ground sensors, a three-axis accelerometer, a thermistor, and a microphone.
 Five capacitive touch buttons organized as a direction pad form an intuitive user interface.
@@
 A remote control receiver provides additional distant buttons.
-A microphone can record sounds and detect claps. 
-Finally, a thermistor measures temperature.
-Most of these sensors cost less then 0.2\,\$, the most expensive being the accelerometer with a cost of about 0.8\,\$, which is an acceptable price given the possibilities it brings to the robot.
+%A microphone can record sounds and detect claps. 
+%Finally, a thermistor measures temperature.
+Most of these devices cost less then 0.2\,\$, the most expensive being the accelerometer with a cost of about 0.8\,\$, which is an acceptable price given the possibilities it brings to the robot.
 %The omnipresence of this type of sensors in smartphones and compact cameras has drastically lowered their prices over the last decade.
-We also chose low-cost toy motors and control them in speed (max. 13\,cm/s) using a custom-designed electronics.
-By measuring the back-electromotive force, speed control is possible without additional encoders.
+We also chose low-cost toy motors and control them in speed (max. 13\,cm/s) measuring the back-electromotive force and therefore removing the need of additional encoders.
 
-However, choosing the optimal combination of sensors and processor is a difficult problem, as hardware interfaces are limited on a microcontroller and therefore compromises must be made.
+%However, choosing the optimal combination of sensors and processor is a difficult problem, as hardware interfaces are limited on a microcontroller and therefore compromises must be made.
 We evaluated several microcontrollers and chose the PIC24F from Microchip because it integrates a \textsc{usb} interface and can drive capacitive touch buttons directly, saving additional components. 
-This microcontroller controls all sensors and actuators, excepted the battery recharging logic, which uses a specific chip for safety reasons.
+This microcontroller controls all sensors and actuators, excepted the internal Li-Po battery recharging logic, which uses a specific chip for safety reasons.
 %This choice of a single microcontroller led us to non-trivial design decisions like using low-cost shift registers to drive the many \textsc{led}s.
@@
 
-In addition, we conceived Thymio to fit the requirements of series production.
-This is a very critical choice in open source hardware projects, and we will discuss this issue in detail in the next section.
-We produced until now more than 10'000 robots by batches of 1'000 units.
-Because manual work is expensive in western countries, we produce through subcontractors in China.
-Indeed, while most of the electronics consists of \textsc{smd} components mounted by a robot, some are soldered (sensors, microphone, \textsc{led}s) by hand and the robot is also assembled by hand.
-We streamlined the production of plastic parts (Thymio has only 5 parts while our previous prototype had 11) and opted for injection molding. 
-The plastic hull consists of two main parts screwed together to facilitate the vertical assembly and disassembly of the robot.
-Moreover, when selecting features we took into consideration the complexity of the injection mold.
+For our specific design, we needed custom made mechanical parts.
+To reduce the price, all mechanical parts are in injected plastic, for a total production cost of less than 4\,\$.
 
+Our choice of electronic components implies a production that is either fully automatized or includes some manual operations. 
+A final assembly by the user is excluded by the certification required in the case of use by children. 
+As a full automatization requires investments that are beyond the possibilities of this project, the current production is based on some manual operations. 
+We therefore decided to produce in China, where manual work is less expensive.
+%This is a very critical choice in open source hardware projects, and we will discuss this issue in detail in the next section.
+We produced until now more than 15,000 robots by batches of 1,000 units, with a cost per robot of 35\,\$.
+%Because manual work is expensive in western countries, we produce through subcontractors in China.
+%Indeed, while most of the electronics consists of \textsc{smd} components mounted by a robot, some are soldered (sensors, microphone, \textsc{led}s) by hand and the robot is also assembled by hand.
+%We streamlined the production of plastic parts (Thymio has only 5 parts while our previous prototype had 11) and opted for injection molding. 
+%The plastic hull consists of two main parts screwed together to facilitate the vertical assembly and disassembly of the robot.
+%Moreover, when selecting features we took into consideration the complexity of the injection mold.
+%The design of the mechanical parts has been aligned with these choicesthat are injected, resulting in all custom parts of Thymio costing less than 4\,\$.
+%In these conditions Thymio costs 35\,\$. 
+The strict quality control, the production management, the support to the users and the margins for distributors bring to a selling price of 130\,\$.
+
+\iffalse
 \begin{table}
@@
 \end{table}
+\fi
 
-Thymio finally costs 35\,\$ per unit when produced in a batch of 1\,k units (see Table~\ref{tbl:thymio-price}), with fixed fees of 15\,k\$ (13\,k\$ for the mold and 2\,k\$ for the \textsc{ce} certification).
-This allows a selling price of \textsc{chf}\,129 ($\approx$ 130\,\$).
+%Thymio finally costs 35\,\$ per unit when produced in a batch of 1\,k units, with fixed fees of 15\,k\$ (13\,k\$ for the mold and 2\,k\$ for the \textsc{ce} certification).
+%This allows a selling price of \textsc{chf}\,129 ($\approx$ 130\,\$).
 
@@
 
-While we optimized Thymio for cost, we also targeted a large range of users thanks to an important contribution by industrial designers of the University of Art and Design of Lausanne\footnote{\url{http://www.ecal.ch}}.
-The variety of sensors, the multiple ways of interacting with the robot, its hull design, and its customization make Thymio accessible to girls and boys of different age groups.
-The white hull is initially neutral and children can choose their own color using the \textsc{rgb} \textsc{led}s or attach accessories to the hull.
-The robot can be used from children as young as 6 years old to adults:
+Several design choices, like the variety of sensors, the multiple ways of interacting with the robot, the neutral hull design, the various programming environments, and the possible customization with accessories, contribute to make Thymio accessible to girls and boys of different age groups from kindergarden to university~\cite{Riedo2013}.
+These design choices were made and implemented thanks to an important contribution by industrial designers of the University of Art and Design of Lausanne\footnote{\url{http://www.ecal.ch}}.
+The white neutral hull is a key element in this set of choices, and it is the opposite than the technical look chosen for the Edison or the LEGO\textsuperscript{\textregistered} robots. 
+These last robots are implicitly targeting a group of people interested in technical systems, mostly males, while Thymio is open to both genders and a larger target audience, from 6 years old kids to adults.
+%The white hull is initially neutral and children can choose their own color using the \textsc{rgb} \textsc{led}s or attach accessories to the hull.
+
+\iffalse
+This impacts also on the type of activities:
 The youngest play with the robot through a set of basic behaviors (see section \ref{sec:behaviors}) and can use Thymio as a support for handicrafts or construction.
@@
 Finally, because of the open-source and open-hardware nature of the project, advanced programmers can improve their skills in C by modifying the firmware or in electronics by disassembling the robot.
-
-\subsection{Facilitating Learning}
-
-When designing Thymio, we took care of providing many incentives for the users to learn new things throughout their direct interaction with the robot.
-This translates into specific hardware and software choices.
-
-At the hardware level, we render visible the activity of the sensors by adding a \textsc{led} next to each of them, for a total of 39 \textsc{led}s.
-These \textsc{led}s locally color the hull and allow the user to see immediately where and when the robot perceives a change in its environment.
-There is a \textsc{led} next to each proximity sensor that lights up as soon as an object is close enough to be seen, and shines brighter the closer the object gets.
-Similarly, a combination of blue and red \textsc{led}s show the temperature, and the infrared remote control receiver and the microphone both have \textsc{led}s that flash when they detect something.
-On the top of the robot, a circle of 8 \textsc{led}s shows the 3D inclination of the robot thanks to the accelerometer. 
-This circle is also used to reflect some of the behaviors of Thymio.
-Finally, two strong \textsc{rgb} \textsc{led}s color the whole top of the robot.
-In addition to a visual feedback, the capacitive buttons also have an acoustic feedback.
-The link between a sensor and its feedback can be turned off when programming the robot to use the \textsc{led}s and the loudspeaker for other purposes.
-
-At the software level, we provide a set of programming environments (see section \ref{sec:aseba}) that enable beginners to discover programming progressively.
-First, we teach them the basic rules of programming using a purely visual interface, then they discover the construction of syntax trees by assembling graphical blocks, and finally we provide a full text-based coding environment with advanced debugging tools, such as real-time inspection of the variables of the robot and plotting features, providing a visual way to understand time-related concepts.
+\fi
 
@@
 
-We wanted the robot to be a starting point for users to invent their own creations.
-Therefore, we thought Thymio as a support for handicrafts and constructions.
-Its white color is meant to be seen as a blank page that can be decorated and drawn upon, and its shape allows easy integration into a structure.
-The robot features a hole in its middle to insert a marker for drawing.
-It also has a slot for a microSD card, allowing users to add their own sounds or to load a pre-compiled code by the simple insertion of a dedicated card.
-
-The square format of the hull allows to use the robot as a base for driving the user's own constructions.
-To that end, Thymio is compatible with LEGO\textsuperscript{\textregistered} bricks:
-There are four attachment positions on the top of the robot and two LEGO\textsuperscript{\textregistered} Technics crosses on each side.
-Additionally, the two wheels have attachment points, which permits to use them to actuate elements elsewhere in the structure (Fig.~\ref{fig:example-construction}, third row left) or to lift the robot's own weight (Fig.~\ref{fig:example-construction}, third row right).
+%We wanted the robot to be a starting point for users to invent their own creations.
+%Therefore, we thought Thymio as a support for handicrafts and constructions.
+The white neutral hull is also meant to represent a blank page that can be decorated and drawn upon, and its shape allows easy integration into a larger structure.
+%The robot features a hole in its middle to insert a marker for drawing.
+%It also has a slot for a microSD card, allowing users to add their own sounds or to load a pre-compiled code by the simple insertion of a dedicated card.
+The square format of the hull facilitates the use of the robot as a base of the user's own constructions.
+To that end, Thymio is compatible with LEGO\textsuperscript{\textregistered} bricks, both on the body than on the wheels. 
+This last connection can be used to actuate elements elsewhere in the added structure (Fig.~\ref{fig:example-construction}, third row left) or to lift the robot's own weight (Fig.~\ref{fig:example-construction}, third row right).
+%There are four attachment positions on the top of the robot and two LEGO\textsuperscript{\textregistered} Technics crosses on each side.
+%Additionally, the two wheels have attachment points, which permits to use them to actuate elements elsewhere in the structure (Fig.~\ref{fig:example-construction}, third row left) or to lift the robot's own weight (Fig.~\ref{fig:example-construction}, third row right).
 Therefore we chose more powerful motors than strictly necessary to move the robot around.
-The same fixation points used for LEGO\textsuperscript{\textregistered} bricks can be used to attach paper structures.
-Paper can allow to simply change the body shape and add some body movements, as illustrated in Fig.~\ref{fig:example-construction} by the orca, opening and closing its mouth while moving forward or by the bat, moving its wings.
+%The same fixation points used for LEGO\textsuperscript{\textregistered} bricks can be used to attach paper structures.
+Paper can also be used to change the body shape and/or add some body movements, as illustrated in Fig.~\ref{fig:example-construction} by the orca, opening and closing its mouth while moving forward or by the bat, moving its wings.
 But paper and cardboard can also radically change the locomotion principle, as illustrated in the second row of Fig.~\ref{fig:example-construction} by the zombie, where the wheels of the robot activate the legs.
@@
 
+\subsection{Facilitating Learning}
+
+When designing Thymio, we took care of providing many incentives for the users to learn new things throughout their direct interaction with the robot.
+This translates into specific hardware and software choices.
+
+At the hardware level, we render visible the activity of the various robot components by adding a \textsc{led} next to each of them, for a total of 39 \textsc{led}s.
+These \textsc{led}s locally color the hull and allow the user to see immediately where and when the robot perceives a change in its environment: proximity of objects, changes in the ground color, temperature, sound, or accelerations.
+%\textsc{led}s show, by their light intesity or their color, the measurements of the proximity sensors, the temperature sensor, the microphone, and the accelerometer. 
+Some \textsc{led}s display data exchanges, from the infrared remote control receiver or with the microSD card. 
+The capacitive buttons give both a visual and an acoustic feedback.
+%There is a \textsc{led} next to each proximity sensor that lights up as soon as an object is close enough to be seen, and shines brighter the closer the object gets.
+%Similarly, a combination of blue and red \textsc{led}s show the temperature, and the infrared remote control receiver and the microphone both have \textsc{led}s that flash when they detect something.
+%On the top of the robot, a circle of 8 \textsc{led}s shows the 3D inclination of the robot thanks to the accelerometer. 
+%This circle is also used to reflect some of the behaviors of Thymio.
+%Finally, two strong \textsc{rgb} \textsc{led}s color the whole top of the robot.
+%In addition to a visual feedback, the capacitive buttons also have an acoustic feedback.
+The link between a sensor and its feedback can be turned off when programming the robot to use the \textsc{led}s and the loudspeaker for other purposes.
+
+At the software level, we provide a set of programming environments (see section \ref{sec:aseba}) that enable beginners to discover programming progressively.
+First, we teach them the basic rules of programming using a purely visual interface, then they discover the construction of syntax trees by assembling graphical blocks, and finally we provide a full text-based coding environment with advanced debugging tools, such as real-time inspection of the variables of the robot and plotting features, providing a visual way to understand time-related concepts.
+%Robotics experts can program Thymio using the \textsc{ros}\footnote{accessible through the \emph{asebaros} bridge, see \url{http://www.ros.org/wiki/asebaros}} environment~\cite{quigley2009ros}.
+
 \subsection{Fast access to robotics behaviors}
@@
 Many existing robots need to be built or configured before having the first operational behavior. 
-This is a barrier for many users and we wanted to avoid it carefully, having a robot able to show interesting behaviors right out of the box.
-Therefore Thymio provides 6 different basic behaviors (see table \ref{tbl:basic-behaviors}), which are accessible as soon as the robot is started and are stored in flash permanently, also if the Thymio is reprogrammed by the user.
+For instance the Edison robot needs to read a bar code and the mBot needs to be assembled.
+This can be a barrier for school activities and we wanted to avoid it carefully, having a robot able to show interesting behaviors right out of the box.
+Therefore Thymio has six different basic behaviors accessible as soon as the robot is started and stored in flash permanently.
 These basic behaviors allow people starting Thymio to immediately interact with it, while illustrating the many possibilities of the robot.
-The behaviors are interactive.
-For each behavior, the robot's body has a different color, allowing the user to easily recognize it.
-The user can navigate between the colors with the buttons and select the behavior she/he wants to use.
-These behaviors can be exploited in a construction to create reactivity without the use of programming, such as in the paper creations shown in the top four images of Fig.~\ref{fig:example-construction}.
+%The behaviors are interactive.
+%For each behavior, the robot's body has a different color, allowing the user to easily recognize it.
+%The user can navigate between the colors with the buttons and select the behavior she/he wants to use.
+%These behaviors can be exploited in a construction to create reactivity without the use of programming, such as in the paper creations shown in the top four images of Fig.~\ref{fig:example-construction}.
+The user can already start creating constructions on top of these basic behavior without the need of programming, such as in the paper creations shown in the top four images of Fig.~\ref{fig:example-construction}.
 
+\iffalse
 \begin{table}
@@
 \end{table}
+\fi
 
-\subsection{Programming}
+\subsection{Programming environment}
 \label{sec:aseba}
@@
 This capability is provided both with the \ac{vpl}~\cite{Magnenat2015} and in the scripting environment, by displaying the content of variables in real time through texts or plots.
-The \ac{ide} also provides a help documentation\footnote{currently in English, French, German, Italian and Spanish}.
-In addition, the wiki pages of the project provide tutorials on programming with Thymio and a detailed description of the programming interface.
+%The \ac{ide} also provides a help documentation\footnote{currently in English, French, German, Italian and Spanish}.
+%In addition, the wiki pages of the project provide tutorials on programming with Thymio and a detailed description of the programming interface.
 %The latter was designed with consistency in mind: variables and functions are available in namespaces and similar features (such as the different \textsc{led}s) are accessed the same way.
@@
 
-The choice of developing a robot within an open source hardware project has an impact on the robot design and the way this device is used in the community of users. 
+The last but not least design choice among the seven mentioned in the previous section, is to have an open source hardware design.
+This choice has an impact on the robot design and the way this device is used in the community of users. 
 In this section we analyze more in detail the implications of this choice in the context of educational robotics.
-We compare these implications with the result of a survey that got 35 answers from people active in various open source hardware projects based on a worldwide call for contributions. 
-Among these 35 answers, 11 come from project leaders, 13 from core design team members, 8 from contributors and 3 from enthusiastic users.
-54\% of these respondents are between 25 and 35 years old, 26\% between 35 and 50 years old, 9\% are less than 25 years old and 11\% are older than 50.  
-67\% either work in an academic environment or are students. 
-34 of the 35 answers come from male respondents.
+We compare these implications with the result of a survey that got 35 answers from 11 project leaders, 13 core design team members, 8 contributors and 3 enthusiastic users of open hardware projects.
+%people active in various open source hardware projects based on a worldwide call for contributions. 
+%Among these 35 answers, 11 come from project leaders, 13 from core design team members, 8 from contributors and 3 from enthusiastic users.
+%Although 35 answer are not many, the high proportion of project leaders and core design members allows to extract some meaningful observations.
+%54\% of these respondents are between 25 and 35 years old, 26\% between 35 and 50 years old, 9\% are less than 25 years old and 11\% are older than 50.  
+%For instance, it is interesting to observe that 34 of the 35 answers come from male respondents.
 
@@
 
-We can distinguish two levels of motivation, the institutional and the personal one.
-The institution initiating a project is mainly looking for recognition and/or money. 
+%We can distinguish two levels of motivation, the institutional and the personal one.
+%The institution initiating a project is mainly looking for recognition and/or money. 
 Our group has a good experience in disseminating robotic hardware with the Khepera~\cite{MonFraIen93} and e-puck~\cite{mondada2009puck} robots.
@@
 Both were targeting similar users and have been sold in similar quantities.
-What we can observe after more than 10 years is that Khepera generated royalties for the university, but less relative academic visibility than e-puck (Fig.~\ref{fig:khepuck}).
+What we can observe after more than 10 years is that Khepera generated royalties for the university, but the name of the robot was associated with the name of the company producing it, not the university that developed it (Fig.~\ref{fig:khepuck}).
 The e-puck robot, being open hardware and with an image better linked to the university, generated no income for the university but much more visibility.
-Khepera was more linked with the name of the company producing it, the e-puck robot was nearly not linked with the producer when mentioned in the literature.
-In the case of Thymio, developed in a research project, the institutional motivation was to get visibility more than money. 
-Therefore, based on the past experience, the open source hardware strategy seemed more adapted.
+%Khepera was more linked with the name of the company producing it, the e-puck robot was nearly not linked with the producer when mentioned in the literature.
+In the case of Thymio the institutional motivation was to get visibility more than money. 
+Therefore, from an institutional point of view, the open source hardware strategy seemed more adapted.
 
@@
 \includegraphics[width=\columnwidth]{figures/others_robots}
-\caption{Comparison of the impact of e-puck and Khepera.
+\caption{Comparison between e-Puck and Khepera in term of mentions in google scholar, either associated with the university of with the company producing them.
 %NOTE: SM: this is an analysis and should not be in caption
@@
 
-The personal motivation to participate to such a project is very different than the institutional one. 
-When asked about their personal motivation, the people participating to the survey give as main motivation a link to their professional activity and to the specific project, followed by a more general motivation by the nature of the project and finally by the engineering practice (Fig.~\ref{fig:motivation}).
-The more abstract goal of improving our society is mentioned only by few respondents.
-
+Beside the institutional motivation, there is the personal motivation of each contributor.
+%The personal motivation to participate to such a project is very different than the institutional one. 
+When asked about their personal motivation, the people participating to the survey give as main motivation a link to their professional activity and to the specific project.
+%, followed by a more general motivation by the nature of the project and finally by the engineering practice (Fig.~\ref{fig:motivation}). 
+Very few mention a more abstract goal like improving our society.
+\iffalse
 \begin{figure}
@@
 \end{figure}
+\fi
 
-In the Thymio project, most engineering contributors were hired for a research project developing the robot. 
-Industrial designers also contributed as part of an institutional project.
-Therefore the link with the professional activity and the particular project is evident.
-On another side, the motivation of contributing to society is strong in our team, as developing a robot targeting education has an important societal component.
+Also in the Thymio project, most contributors participate to the development as part of their job, but the motivation to contribute to society is stronger, as developing a robot targeting education has an important societal component.
+%were professionally involved in the professionals involved in the development of the robot because XXXX
+%Industrial designers also contributed as part of an institutional project.
+%Therefore the link with the professional activity and the particular project is evident.
+%On another side, the motivation of contributing to society is strong in our team, as developing a robot targeting education has an important societal component.
 Moreover, our project also has a strong scientific motivation, with several ongoing studies on the acceptance by teachers and the effect on children's learning.
-Therefore, sharing a strong fundamental motivation such as education or scientific achievements, is a key element for building a solid community~\cite{Stahlbrost2011}, especially if it is interdisciplinary like ours.
+Sharing a strong fundamental motivation, such as education or scientific achievements, is a key element for building a solid community~\cite{Stahlbrost2011}, especially if it is interdisciplinary like ours.
 
-It is also interesting to look at what people expect as benefits in participating in such a project (Fig.~\ref{fig:getout}).
-We can observe both a technological component in getting better results and impact, and a human-relations component that is not in the basic definition of open hardware but results from the community created around the project.
-People expect to work with other people to achieve more together, to create a network, and to learn through new experience and contacts.
-
+What do people expect as benefits in participating in such a project? Fig.~\ref{fig:getout} shows the answers from the survey.
+We can observe both a technological and a human-relations component, resulting from the community created around the project. 
+%People expect to work with other people to achieve more together, to create a network, and to learn through new experience and contacts. 
 In the Thymio project we had similar expectations. 
@@
 We established a wiki\footnote{http://www.thymio.org} as the meeting point for the learners, the robot developers, and the teachers.
-This wiki introduces the robot, explains the basic behaviors, gives access to the programming environment and its documentation, and provides code samples and examples of constructions.
+%This wiki introduces the robot, explains the basic behaviors, gives access to the programming environment and its documentation, and provides code samples and examples of constructions.
 It is open for editing by anyone, and although we initially provided most of the material, the other members of the community have started to contribute.
-One of the most important contributions to the wiki, in term of effort, has been its translation into four languages.
+One of the important contributions from people outside the core design team was the translation of the wiki into four languages.
 
-In addition to those elements, in our project we expected another important benefit from open hardware in term of image: we wanted a match between the non-profit nature of the project and the non-profit nature of education in general.
+A last important motivation for having an open hardware project was because of the resulting image: we wanted a match between the non-profit nature of the project and the non-profit nature of education in general.
 
@@
 When starting an open source hardware project, one of the typical questions is which license to use when disseminating the project source files. 
+All Thymio documentation and plans are distributed under the CC-BY-SA 3.0 license and software under the LGPL license.
 We will not discuss this matter here, as it is a very common and well-covered issue.
@@
 There is another licensing issue which is not well known and that we discovered very late in our project: the constraints of the license of the mechanics and electronics \textsc{cad} tools. 
-Indeed, when asked about this issue, the participants to our survey seem generally not aware of the fact that \textsc{cad} licenses can be very restrictive about the way the source files can be published (Fig.~\ref{fig:aware}).
-More than the half or respondents to our survey state that they were not aware that this could be an issue, and only one third checked the license of their \textsc{cad} software. 
+Indeed, when asked about this issue, 57\% of the participants to our survey are not aware of the fact that \textsc{cad} licenses can be very restrictive about the way the source files can be published, and only one third checked the license on this aspect.
 %More than the half or respondents to our survey, and more than 60\% among the project leaders, state that they were not aware that this could be an issue, and only one third checked the license of their \textsc{cad} software.
-Some stated that they did not checked the license because they were sure they can freely publish their design, but one third of them are using software not allowing the publication of source files when using educational licenses, and all these respondents are academics or students.
+%Some stated that they did not checked the license because they were sure they can freely publish their design, but one third of them are using software not allowing the publication of source files when using educational licenses, and all these respondents are academics or students.
 % FIXME: SM: there is a bit of duplication with previous sentence, I think these can be compressed
-This issue is very serious, as most contributors of open source hardware projects are academics and use academic or educational licenses. 
-Most licenses do not allow to use the design created with this version of the software for commercial activities. 
+This issue is very serious, as most contributors of open source hardware projects are academics and use educational licenses, not allowing commercial use of the resulting designs. 
 As producing or selling the product is part of the definition of open hardware, these licenses simply forbid publication under the standard open source hardware conditions.
 
+\iffalse
 \begin{figure}
@@
 \end{figure}
-
+\fi
 To clarify this issue we contacted twelve of the major editors of mechanical \textsc{cad} and \textsc{pcb} routing software.
-We asked them if their educational license allows the publication of the source files, and we specified that the published files could have been downloaded by a company to produce the system for commercial purposes.
+We asked them if their educational license allows the publication of the source files.
+%, and we specified that the published files could have been downloaded by a company to produce the system for commercial purposes.
 Fig.~\ref{fig:editors} summarizes the result of this survey.
@@
 This can multiply by factors of hundreds the price of the \textsc{cad} license.
-This blocking factor for open hardware also applies to the publication of scientific results, as promoted by many governments in the last decade and generally called ``open science''.
-The change of policy in \textsc{cad} files publication that we obtained is a sign of the very positive trend set by open hardware and by open science in general.
+%This blocking factor for open hardware also applies to the publication of scientific results, as promoted by many governments in the last decade and generally called ``open science''.
+%The change of policy in \textsc{cad} files publication that we obtained is a sign of the very positive trend set by open hardware and by open science in general.
 
@@
 
-As a conclusion, this legal issue is totally underestimated by both people participating to the projects and by the \textsc{cad} editors. 
-In a period where editors are looking for additional revenues and are attacking universities for misuse of licenses\footnote{We are aware, just in Switzerland, of two situations where large amounts of money are asked to universities for non-respect of software licenses.}, this can be a very dangerous situation.
+As a conclusion, this legal issue is totally underestimated by both people participating to the projects and by the \textsc{cad} editors, and is a potential threat for many projects. 
+%In a period where editors are looking for additional revenues and are attacking universities for misuse of licenses\footnote{We are aware, just in Switzerland, of two situations where large amounts of money are asked to universities for non-respect of software licenses.}, this can be a very dangerous situation.
 
-\subsection{Who designs and produces the hardware}
+\subsection{Who designs, produces, and supports the hardware}
 
-Among the fundamental choices when starting an open source hardware project, there is the choice of the type of production. 
+%Among the fundamental choices when starting an open source hardware project, there is the choice of the type of production. 
 In the definition of open hardware we gave, it is stated that one should offer an ``hardware whose design is made publicly available so that anyone can [\dots] make [\dots] hardware ''.
 Behind the ``anyone'' should we consider every single person or only companies able to produce the product?
-This choice has implications on who designs the system and how.
-In our project we have two different types of hardware: the robot itself and the accessories used in specific activities.
+%This choice has implications on who designs the system and how.
+
+In our project we decided to have two different types of hardware: the robot itself and the accessories. % used in specific activities.
 The robot is the expensive part and has a very neutral design, allowing adaptation to specific situations.
@@
 For the robot itself, we opted for the interpretation of the definition that considers under ``anyone'' only the professional structures able to mass-produce hardware, for a reason of price and complexity of the product.
-When looking for the best performance per price ratio, we decided to use techniques that require heavy equipment.
-For instance we decided to produce all mechanical parts by injection molds. 
-This ensures a price per part of some cents, which cannot be achieved by others techniques.
-Letting end-user produce their own parts would results in higher prices for less performances. 
-This choice has another implication: only a core team of highly skilled engineers can contribute to the project. 
-It is not trivial to design a mechanical part that can be molded in the simplest possible way.
+%When looking for the best performance per price ratio, we decided to use techniques that require heavy equipment.
+%For instance we decided to produce all mechanical parts by injection molds. 
+%This ensures a price per part of some cents, which cannot be achieved by others techniques.
+%Letting end-user produce their own parts would results in higher prices for less performances. 
+%This choice has another implication: only a core team of highly skilled engineers can contribute to the project. 
+%It is not trivial to design a mechanical part that can be molded in the simplest possible way.
 
@@
 
+\iffalse
 \begin{figure}
@@
 \end{figure}
+\fi
 
-Our survey shows that the most common production method is a centralized one by commercial companies (Fig.~\ref{fig:production}).
-Indeed, large open source hardware projects like Arduino are based on this model.
-One could argue that this model makes these robotic projects close to standard commercial models, leading to a lack of involvement from users and loosing the open source spirit.
-In our case, we decided to alleviate this effect by creating a non-profit association, called Mobsya, in charge of producing and selling the robots. 
+%Our survey shows that the most common production method is a centralized one by commercial companies.% (Fig.~\ref{fig:production}).
+%Indeed, large open source hardware projects like Arduino are based on this model.
+%One could argue that this model makes these robotic projects close to standard commercial models, leading to a lack of involvement from users and loosing the open source spirit.
+%In our case, we decided to alleviate this effect by creating a non-profit association, called Mobsya, in charge of producing and selling the robots. 
 %While anybody else could produce the Thymio follwing a more buiness-oriented model, the initial production has been done respecting a coherent and global non-profit strategy, in line with educational values of the community of users.
-Moreover, the users are better integrated in the second layer of open source hardware consisting of accessories, where anybody can contribute their own designs and put them on the wiki.
+%Moreover, the users are better integrated in the second layer of open source hardware consisting of accessories, where anybody can contribute their own designs and put them on the wiki.
 
-Another strong difference between a standard consumer product and our vision of open hardware is that Thymio should be durable.
-Schools do not have large budgets for technological tools and make strong investment in training of their teachers when adopting a new tool.
-Therefore, the lifetime of the products should be as long as possible. 
-The open hardware approach gives to the user, or to a generic technician, better conditions to repair the system.
+%Another strong difference between a standard consumer product and our vision of open hardware is that Thymio should be durable.
+Another strong element of our vision of open hardware is that Thymio should be durable.
+%Schools do not have large budgets for technological tools and make strong investment in training of their teachers when adopting a new tool.
+As schools invest in long term training of teachers, for instance, the lifetime of the products should also be as long as possible. 
+The open hardware approach fits well to this requirement, as it gives to the user, or to a generic technician, better conditions to repair the system, having the specifications of all components.
+This is not the case of commercial robots like Edison, for instance.
 Supporting this type of operation has an impact on the robot design; for instance Thymio can be easily opened with four standard screws, and we introduced connectors between key elements such as motors, speaker, the battery and the main \textsc{pcb}.
-
 Another key element in supporting repairs by end users is the documentation of calibration methods. 
@@
 For example, in the Thymio the right and left wheel motors can differ in their electrical characteristics, resulting in the robot not going straight for similar speed commands to both wheels.
-To correct this problem, we introduced factory calibration, storing in the microcontroller memory correction factors that are specific for each robot. 
-To allow the user to replace a broken motor, it is essential to give her also the possibility to re-calibrate the robot and adjust the parameters of the new motor.
+To correct this problem, we introduced factory calibration. %, storing in the microcontroller memory correction factors that are specific for each robot. 
+To allow the user to replace a broken motor, it is essential to give him also the possibility to re-calibrate the robot and adjust the parameters of the new motor.
 In Thymio, this results in the design and the documentation of calibration processes that can be performed by anyone, getting close to the original definition of open source hardware.
 
+\iffalse
 \subsection{Added value of open source hardware in distribution}
@@
 A key element that allowed to keep cohesion in our community was to share, among all partners, strong basic values about improvement of society through education.
+\fi
 
@@
 
+%XXXXX present a better overview of the significant improvements that the design choices bring in relation to the design goals  XXXXXXX
 The introduction of robots in formal education is a very challenging task, not only because of technical requirements such as low cost and interactivity, but also because of factors depending on the school environment, such as the diversity of the educational programs, the dependence on local structures and languages, or the required training of teachers.
-Most of the current robotic tools for education capitalize upon existing hardware platforms, programming environments, or robot architectures.
-Because they do not consider all elements at stake, these tools fail to meet some challenges of formal education.
-On the contrary, in this paper we presented a global approach that tackles these questions in a holistic way, based on open source and non-profit principles.
-The result is the Thymio robot and its surrounding open source community including engineers, art designers, production and sales people, and teachers.
-The open source hardware and the related contributions are split in two categories: the robot itself, produced by a non-profit organization and requiring very advanced skills in both design and production, and the accessories for activities, based on techniques broadly accessible and with a larger panel of contributors.
+Most of the current educational robotic activities capitalize upon one strong element, for instance the technical innovation, but miss to match the formal education requirements.
+%Because they do not consider all elements at stake, these tools fail to meet some challenges of formal education.
+%On the contrary, in this paper we presented a global approach that tackles these questions in a holistic way, based on open source and non-profit principles.
+%The result is the Thymio robot and its surrounding open source community including engineers, art designers, production and sales people, and teachers.
+%The open source hardware and the related contributions are split in two categories: the robot itself, produced by a non-profit organization and requiring very advanced skills in both design and production, and the accessories for activities, based on techniques broadly accessible and with a larger panel of contributors.
 
-The open source hardware development and dissemination of Thymio addresses several of the issues found in educational robotics.
-It allowed to broadly distribute the robot with minimal changes dues to management of intellectual properties, royalties, financial support and so on.
-This was achieved with an excellent match between the philosophy of the project and the one of the community of users in education.
-In particular, the open source approach allows to provide a durable robot, easy to maintain and repair, with at the same time a community of users providing educational material and mutual support.
+The open source hardware approach used in the Thymio project addresses this and several other issues found in educational robotics.
+The inclusion of education scientists, teachers and designers is possible because of the open nature of the project and the split between core technology, produced in a central way, and accessories, accessible with DIY approaches.
+This split allowed to ensure also a gender- and age-neutral basic robot.
+This large inclusion of users and contributors allowed the production, in parallel to the robot technology, of a large set of pedagogic scenarios.
+
+The philosophy of open source and free access to information fits extremely well with the community of users in education, and was reinforced by producing the robot in a non-profit structure.
+This appraoch allowed to broadly distribute the robot with minimal changes dues to management of intellectual properties, royalties, financial support and so on.
+Moreover, the open source approach allows to provide a durable robot, easy to maintain and repair, with at the same time a community of users providing educational material and mutual support.
 
 By making a survey among contributors to open source hardware projects, we could observe that our project shares some characteristics with the majority of the projects represented in the survey. 
-We also faced difficulties in establishing the community, we chose production and distribution methods that are broadly applied, and we share most of the motivational elements that are behind other projects.
-We identified an underestimated legal issue for open source hardware projects in the licensing term of \textsc{cad} software.
-Indeed only very few editors have licenses allowing the publication of source files created with educational license.
-Two surveys among the users and the \textsc{cad} editors show that both people contributing or leading open source hardware projects and the editors themselves are not aware of this issue.
+%We also faced difficulties in establishing the community, we chose production and distribution methods that are broadly applied, and we share most of the motivational elements that are behind other projects.
+We identified for instance an underestimated legal issue for open source hardware projects in the licensing term of \textsc{cad} software.
+%Indeed only very few editors have licenses allowing the publication of source files created with educational license.
+%Two surveys among the users and the \textsc{cad} editors show that both people contributing or leading open source hardware projects and the editors themselves are not aware of this issue.
 Finally, we could show some elements, specific to educational robotics, that differentiate our project from other open source hardware projects.
 In particular, our project takes advantage of an alignment between the principles underlying open source and the nature of education institutions. 
-This fits a political trend toward open source in education and in science in general.
+%This fits a political trend toward open source in education and in science in general.
 We also found a solution to the problem of production methods by splitting our hardware in two categories, enabling both advanced technology for the robot and a large variety of accessories.
-Hence, Thymio appeals to a broad community of end users, which addresses durability of equipment and gender issues that are critical in education.
+Hence, Thymio appeals to a broad community of end users in education, addressing durability and inclusion at several levels.
 

### 2016-10-19 paper_ram.tex
finalization

@@
 L\'ea Pereyre$^{1}$, Philippe R\'etornaz$^{1}$ and St\'ephane Magnenat$^{1}$% <-this % stops a space
-\thanks{*This research was supported by the Swiss National Center of Competence in Research ``Robotics'' (Thymio robot development and deployment), by GebertRuf Stiftung (design of accessories), by the Swiss NSF project CRAGP2 151543 ``Robotics in schools'', and by the EU-FP7 project ASSISIbf, no.
+\thanks{*This research was supported by the Swiss National Center of Competence in Research ``Robotics'' (Thymio robot development and deployment), by GebertRuf Stiftung (design of accessories), by the Swiss NSF project CRAGP2 151543 ``Robotics in schools,'' and by the EU-FP7 project ASSISIbf, no.
 601074 (survey on open hardware).
-Many thank to Luc Bergeron and his team at \'ecal.ch for the industrial design of Thymio; Didier Roy, David Sherman and all the INRIA team for their contributions and diffusion in France; Gordana Gerber for the educational material; all the Mobsya team for the effort in production and sale; and Fabian Hahn for the port of Blockly.}% <-this % stops a space``anyone''
+Many thanks to Luc Bergeron and his team at \'ecal.ch for the industrial design of Thymio; Didier Roy, David Sherman and all the INRIA team for their contributions and diffusion in France; Gordana Gerber for the educational material; all the Mobsya team for the effort in production and sale; and Fabian Hahn for the port of Blockly.}% <-this % stops a space``anyone''
 \thanks{$^{1}$Francesco Mondada, Manon Briod, L\'ea Pereyre, Philippe R\'etornaz and St\'ephane Magnenat are with the Laboratoire de Syst\`emes Robotiques,
-        Ecole Polytechnique F\'ed\'erale de Lausanne (EPFL), Switzerland
-        {\tt\small firstname.lastname@epfl.ch}}%
-\thanks{$^{2}$Michael Bonani and Fanny Riedo are with Mobsya Association, Renens, Switzerland
+        Ecole Polytechnique F\'ed\'erale de Lausanne (EPFL), Switzerland.
+        {General rule for EPFL emails: \tt\small firstname.lastname@epfl.ch}}%
+\thanks{$^{2}$Michael Bonani and Fanny Riedo are with Mobsya Association, Renens, Switzerland.
         {\tt\small michael.bonani@mobsya.org} and {\tt\small fanny.riedo@mobsya.org}}%
@@
 
-Mobile robots are valuable tools for education because of both the enthusiasm they raise and the multidisciplinary nature of their technology.
-They give access to a wide range of fields, such as complex mechanics, sensors, wireless transmission, mathematics, computer science, etc.
+Mobile robots are valuable tools for education because of both the enthusiasm they raise and the multidisciplinary nature of robotic technology.
+Mobile robots give access to a wide range of fields, such as complex mechanics, sensors, wireless transmission, mathematics, and computer science, among others.
 As mobile robots sense the environment and take actions based on their perception, they seem to display \emph{intentions of their own}~\cite{Mondada2001}.
 %This fascinates the users of the robot and creates feelings of accomplishment and power when these users engage in creative activities with the robots.
-This impression of intelligent technology, the permeating presence of robots in science fiction and their projected use in our society give a sense of touching the future.
+This impression of intelligence, the permeating presence of robots in science fiction, and their projected use in our society give a sense of touching the future.
 %Children can spend hours looking at a robot interacting with its environment and adults can approach robots as a new hobby that replaces radio and electronics kits.
@@
 
-Despite their potential as educational tools, robots are still not as widespread in schools as they could be. Among the possible reasons, we believe that the following five are playing a key role:
+Despite their potential as educational tools, robots are still not as widespread in schools as they could be. Among the possible reasons, we believe that the following five play a key role:
 \begin{enumerate}
-\item Although there are many research projects developing innovative and interesting educational robots, few reach a sufficient maturity to become distributed and accessible to schools. 
+\item Although many research projects are developing innovative and interesting educational robots, few reach sufficient maturity to become distributed and accessible to schools. 
 \item A versatile robot performing interesting behaviors is a complex piece of technology and therefore expensive. 
@@
 %Existing platforms with these features cost several hundred Euros.
-This prevents most schools, which have a limited budget for equipment, to acquire educational robots.
+This prevents most schools, which have a limited budget for equipment, from acquiring educational robots.
 \item Introducing robotic tools into teaching activities requires investment in time and training for the teachers~\cite{CERI2008}.
@@
 %Moreover, infrastructure should be provided to allow the sharing of educational material among teachers.
-\item Robot construction, use, and programming is often perceived as a boyish activity in our society~\cite{leonard2009lego,nourbakhsh2009robot}.
+\item Robot construction, use, and programming is often perceived as a boyish activity in our Western society~\cite{leonard2009lego,nourbakhsh2009robot}.
 % For instance, several authors report that girls have such a perception of the LEGO Minstorms~\cite{leonard2009lego,nourbakhsh2009robot}.
@@
 \item Finally, many teachers are reluctant to follow volatile trends, especially if these are based on purely commercial arguments.
-Teachers prefer to invest in stable tools, and this is the opposite as the trend of current consumer technology.
-%, which is often very volatile and based on continuous renewal rather than on durability.
+Teachers prefer to invest in stable tools, in contrast to trends in current consumer technology. %, which is often very volatile and based on continuous renewal rather than on durability.
 \end{enumerate}
@@
 Open source hardware projects can address several of these issues in a different way than closed-source purely commercial products. 
-By \textit{open source hardware} we mean, following the definition of the Open Source Hardware Association\footnote{\url{http://www.oshwa.org/definition/}}, ``hardware whose design is made publicly available so that anyone can study, modify, distribute, make, and sell the design or hardware based on that design''.
-In this paper, we show that this concept, implemented in the Thymio project and gathering a community of users, developers and manufacturers, brings a strong added value to the robot and to the educational methods.
-In addition, we compare our experience with other robotics open source hardware projects not focused on education, and highlight challenges and opportunities specific to education.
+By \textit{open source hardware} we mean, following the definition of the Open Source Hardware Association,\footnote{\url{http://www.oshwa.org/definition/}} ``hardware whose design is made publicly available so that anyone can study, modify, distribute, make, and sell the design or hardware based on that design.''
+In this paper, we show that this concept, implemented in the Thymio project through a community of users, developers and manufacturers, brings a strong added value to the robot and to the educational methods.
+In addition, we compare our experience with other robotics open source hardware projects not focused on education and highlight challenges and opportunities specific to education.
 
@@
 
-There is a large number of publications presenting educational robots, from low-cost systems targeting Africa~\cite{Rubenstein2015,Gyebi2015} to extremely sophisticated humanoids~\cite{Hood2015,Mazzoni2016}.
+Many publications present educational robots, from low-cost systems targeting Africa~\cite{Rubenstein2015,Gyebi2015} to extremely sophisticated humanoids~\cite{Hood2015,Mazzoni2016}. 
 Among those, only a handful are commercially available, limiting their validation by educational scientists, who are typically not roboticists.
-This results in 90\,\% of publications about validation of educational results made on LEGO\textsuperscript{\textregistered} Mindstorms\textsuperscript{\textregistered}~\cite{benitti2012exploring}, a widely available commercial product.
-The latest version, EV3\footnote{\url{http://mindstorms.lego.com}}, is expensive ($\approx$ 400\,\$) but offers a wide range of possibilities, especially at the mechanical level using LEGO\textsuperscript{\textregistered} bricks and at the software level with its graphical programming environment.
+As a result, 90\% of publications about validation of educational results have been based on LEGO\textsuperscript{\textregistered} Mindstorms\textsuperscript{\textregistered}~\cite{benitti2012exploring}, a widely available commercial product.
+The latest version, EV3\footnote{\url{http://mindstorms.lego.com}}, is expensive ($\approx$ US\$400) but offers a wide range of possibilities, especially at the mechanical level using LEGO\textsuperscript{\textregistered} bricks and at the software level with its graphical programming environment.
 %NOTE: SM: commented WeDo to save space and streamline story
@@
 %However, it is not a real robot but rather an automaton: There is no perception of the environment, just a good encoder for precise movements and a simple programming interface based on buttons giving directions for the movement.
-Among the other players on the market, one can cite the Edison robot\footnote{\url{https://meetedison.com/}}, extremely low cost (49\,\$) and compatible with LEGO\textsuperscript{\textregistered} bricks.
-The low price has pushed to drastic design choices; small size, weak motors, few sensors, communication with the computer by audio jack and standard AAA batteries not included in the price.
-These choices strongly limits its possible usage.
+Among the other players on the market is the Edison robot\footnote{\url{https://meetedison.com/}}, which is extremely low cost (US\$49) and compatible with LEGO\textsuperscript{\textregistered} bricks.
+The low price has pushed drastic design choices: small size, weak motors, few sensors, communication with the computer by audio jack and standard AAA batteries not included in the price.
+These choices strongly limit its possible usage.
 %Among widely used robots, none is part of an open source hardware project.
 
-Among the robots available on the market, only few are open source and are used in schools:
-Scribbler2, produced and sold by Parallax\footnote{\url{http://www.parallax.com}} ($\approx$ 180\,\$), is a 188\,mm large robot, designed to run on the ground and equipped with few light sensors, one distance sensor, two ground sensors and few \textsc{led} displays.
+Among the robots available on the market, only a few are open source and are used in schools:
+Scribbler2, produced and sold by Parallax\footnote{\url{http://www.parallax.com}} ($\approx$ US\$180), is a 188\,mm large robot, designed to run on the ground and equipped with a few light sensors, one distance sensor, two ground sensors, and few \textsc{led} displays.
 It runs on standard AA batteries and has a hacker port for interfacing electronic extensions.
 It is programmable with a graphical or a textual code interface.  
-The main weakness of Scribbler2 is its limited number of sensors and the limited compatibility with other systems. 
+The main weakness of Scribbler2 is its limited number of sensors and compatibility with other systems. 
 Moreover, there seems to be no active community around its development.
-The e-puck~\cite{mondada2009puck} robot is a robot targeting university-level education.
+The e-puck~\cite{mondada2009puck} robot targets university-level education.
 Well equipped with sensors and actuators, modular and compact, it can be programmed with industry-standard environments.
-Several simulators allow to run experiments of high complexity. 
+Several simulators allow running highly complex experiments. 
 %There is an active community of users and developers around it.
-Its main weakness is its high price ($\approx$ 870\,\$).
-Finch\footnote{\url{http://www.finchrobot.com/}} ($\approx$ 99\,\$) is a very simple robot that has been designed around a wired connection to the computer. 
-This connection reduces electronics requirements, such as batteries or wireless communication, and allows to implement the control entirely on the computer.
-This results in a very broad set of possible programming languages available, which is the real force of this robot. 
+Its main weakness is its high price ($\approx$ US\$870).
+Finch\footnote{\url{http://www.finchrobot.com/}} ($\approx$ US\$99) is a very simple robot that has been designed around a wired connection to the computer. 
+This connection reduces electronics requirements, such as batteries or wireless communication, and allows control to be implemented entirely on the computer.
+This results in availability of a very broad set of possible programming languages, which is the real force of this robot. 
 However, the cable does not allow real autonomy and mobility.
 Finally, the mBot\footnote{\url{http://www.makeblock.cc/mbot/}} is a mobile platform based on an Arduino board.
-Its electronics is simple and inexpensive and the robot only features a couple of sensors, which allows to drastically reduce its price ($\approx$ 75\,\$), but also limits the perception possibilities and therefore the span of use.
+Its electronics is simple and inexpensive and the robot only features a couple of sensors, which allows drastic reduction in its price ($\approx$~US\$75), but also limits the perception possibilities and therefore the span of use.
 
-With respect to these robots, Thymio has a compact size (110\,mm), many interaction possibilities, an affordable price and a large set of sensors.
+With respect to these robots, Thymio has a compact size (110\,mm), many interaction possibilities, an affordable price (US\$130), and a large set of sensors.
 %Its design makes the best use of the recent cheap components coming from the explosive growth of the mobile devices market. 
-To the best of our knowledge, beside Thymio there are no educational products providing a similar integration of sensors and actuators at a price below 130\,\$.
+To the best of our knowledge, beside Thymio there are no educational products providing a similar integration of sensors and actuators at a lower price.
 
-Teachers, from primary school up to high school, are one of the main target user group of the Thymio project.
-They decide which tools are used in their class and are key people in the eduction ecosystem.
+Teachers, from primary school up to high school, are a primary target user group of the Thymio project.
+They decide which tools are used in their class and are key people in the education ecosystem.
 For teachers, the motivation to use robotic tools depends on many factors~\cite{chevalier2016}.
-Among them, the availability of material and training plays a key role.
+Among them, the availability of materials and training plays a key role.
 %The BeeBot mentioned previously, because of its radical simplicity, is one of the tools that needs the least training. 
@@
 %However, the total lack of sensors reduces the BeeBot to an automaton, easy to use but limited to activities implementable by sequences of motions, without any closed-loop control that would allow teaching computational thinking or robotics concepts.
-The development of educational material and courses to train the teachers requires a huge effort, based on a good mix between robotics and educational skills. 
+The development of educational material and courses to train teachers requires a huge effort, based on a good mix between robotics and educational skills. 
 Moreover, educational material varies from school to school, as requirements are very dependent on local educational programs and languages. 
@@
 Even for large companies like LEGO\textsuperscript{\textregistered}, such a development is not sustainable. 
-To solve this problem, a crowd-sourcing approach can be used.
-An active community of users can contribute to the development of the material in a distributed manner, adapting the material to the local situation.
-LEGO\textsuperscript{\textregistered} itself is moving toward this direction by promoting communities of users~\cite{Hienerth2014} with a user-producer interaction close to those of open hardware projects.
-An open source community regrouping developers, manufacturers and end users is therefore a very interesting model to address the distributed development and sharing of educational material, and the diffusion of training sessions.
+A crowd-sourcing approach may solve this problem; An active community of users can contribute to the development of the material in a distributed manner, adapting the material to the local situation.
+LEGO\textsuperscript{\textregistered} itself is moving this direction by promoting communities of users~\cite{Hienerth2014} with a user-producer interaction similar to that of open hardware projects.
+An open source community regrouping developers, manufacturers, and end users is therefore a very interesting model to address the distributed development and sharing of educational material and the diffusion of training sessions.
 %This model is an extension to education of what is already practiced in projects such as Arduino~\cite{Jamieson2010}.
@@
 	\includegraphics[width=\columnwidth]{figures/wirelessThymioVecto}
-	\caption{The Thymio robot and its main components for the Wireless and the \textsc{usb}-connected versions.}
+	\caption{The Thymio robot and its main components for the wireless- and the \textsc{usb}-connected versions.}
 	\label{fig:schematique}
@@
 
-We designed the Thymio robot along seven main axes: a low price to address a larger number of users; a feature set that suits both genders and multiple ages from young children to adults; a mechanical design that promotes creativity; a combination of sensors, actuators, and programming features that facilitates learning: a set of ready-to-use programs to get fast access to robotic behaviors; an accessible programming environment; and an open source community contributing to the design and the dissemination.
-The result is a miniature (110\,mm in side) differential-wheeled robot suited for use on a desktop (Fig.~\ref{fig:schematique}, top).
-The robot is robust enough to be mishandled by children, as it can fall from a table without breaking itself.
+We designed the Thymio robot along seven main axes: a low price to address a larger number of users; a feature set that suits both genders and multiple ages from young children to adults; a mechanical design that promotes creativity; a combination of sensors, actuators, and programming features that facilitates learning: a set of ready-to-use programs to quickly access robotic behaviors; an accessible programming environment; and an open source community contributing to design and dissemination.
+The result is a miniature differential-wheeled robot suited for use on a desktop (Fig.~\ref{fig:schematique}, top).
+The robot is robust enough to be mishandled by children; it can fall from a table without breaking.
 It features a translucent white hull and a wide range of sensors and actuators (Fig.~\ref{fig:schematique}, bottom).
-The robot embeds a battery, rechargeable by \textsc{usb}, that provides an autonomy of about 3 to 5 hours.
+The robot has an embedded battery, rechargeable by \textsc{usb}, that provides 3 to 5 hours of power.
 More details on the robot and the previous research results can be found in~\cite{RiedoPhD, magnenat2014}.
@@
 Price is key in the adoption of robots by schools~\cite{kradolfer2014sociological}.
-Therefore the design of Thymio had to target a low production cost, while including a broad range of functionalities to enable flexibility in its use. 
-As in this type of robot the main cost comes from electronics and sensors~\cite{thesis_michael}, we focused on low-cost sensors that allow rich interaction possibilities, with both the environment and the user.
+Thus, the design of Thymio targeted low production costs while including a broad range of functionalities enabling flexibility.
+%Therefore the design of Thymio had to target a low production cost, while including a broad range of functionalities to enable flexibility in its use. 
+As in this type of robot the main cost comes from electronics and sensors~\cite{thesis_michael}, we focused on low-cost sensors that allow rich interactions with both the environment and the user.
 
 The resulting Thymio robot possesses a large number of simple sensors:
-Seven horizontal proximity sensors, two ground sensors, a three-axis accelerometer, a thermistor, and a microphone.
+seven horizontal proximity sensors, two ground sensors, a three-axis accelerometer, a thermistor, and a microphone.
 Five capacitive touch buttons organized as a direction pad form an intuitive user interface.
@@
 %Finally, a thermistor measures temperature.
-Most of these devices cost less then 0.2\,\$, the most expensive being the accelerometer with a cost of about 0.8\,\$, which is an acceptable price given the possibilities it brings to the robot.
+Most of these devices cost less then US\$0.20, the most expensive being the accelerometer with a cost of about US\$0.80, which is an acceptable price given the possibilities it brings to the robot.
 %The omnipresence of this type of sensors in smartphones and compact cameras has drastically lowered their prices over the last decade.
-We also chose low-cost toy motors and control them in speed (max. 13\,cm/s) measuring the back-electromotive force and therefore removing the need of additional encoders.
+We also chose low-cost toy motors and control them in speed (max. 13\,cm/s) measuring the back-electromotive force and therefore removing the need for additional encoders.
 
@@
 We evaluated several microcontrollers and chose the PIC24F from Microchip because it integrates a \textsc{usb} interface and can drive capacitive touch buttons directly, saving additional components. 
-This microcontroller controls all sensors and actuators, excepted the internal Li-Po battery recharging logic, which uses a specific chip for safety reasons.
+This microcontroller controls all sensors and actuators, with the exception of the internal Li-Po battery recharging logic, which uses a specific chip for safety reasons.
 %This choice of a single microcontroller led us to non-trivial design decisions like using low-cost shift registers to drive the many \textsc{led}s.
@@
 
-For our specific design, we needed custom made mechanical parts.
-To reduce the price, all mechanical parts are in injected plastic, for a total production cost of less than 4\,\$.
+For our specific design, we needed custom-made mechanical parts.
+To reduce the price, all mechanical parts are injected plastic, for a total production cost of less than US\$4.
 
-Our choice of electronic components implies a production that is either fully automatized or includes some manual operations. 
-A final assembly by the user is excluded by the certification required in the case of use by children. 
+Our choice of electronic components implies production that is either fully automatized or includes some manual operations. 
+Full assembly is required in order for the robots to be certified for use by children. 
+%A final assembly by the user is excluded by the certification required in the case of use by children. 
 As a full automatization requires investments that are beyond the possibilities of this project, the current production is based on some manual operations. 
@@
 %This is a very critical choice in open source hardware projects, and we will discuss this issue in detail in the next section.
-We produced until now more than 15,000 robots by batches of 1,000 units, with a cost per robot of 35\,\$.
+We have thus far produced more than 15,000 robots in batches of 1,000 units, with a cost per robot of US\$35.
 %Because manual work is expensive in western countries, we produce through subcontractors in China.
@@
 %In these conditions Thymio costs 35\,\$. 
-The strict quality control, the production management, the support to the users and the margins for distributors bring to a selling price of 130\,\$.
+The strict quality control, the production management, the support to the users, and the margins for distributors result in a final selling price of US\$130.
 
@@
 
-\subsection{Multi-age and Gender-neutral Feature Set}
+\subsection{Multi-Age and Gender-Neutral Feature Set}
 \label{sec:multi}
 
-Several design choices, like the variety of sensors, the multiple ways of interacting with the robot, the neutral hull design, the various programming environments, and the possible customization with accessories, contribute to make Thymio accessible to girls and boys of different age groups from kindergarden to university~\cite{Riedo2013}.
-These design choices were made and implemented thanks to an important contribution by industrial designers of the University of Art and Design of Lausanne\footnote{\url{http://www.ecal.ch}}.
-The white neutral hull is a key element in this set of choices, and it is the opposite than the technical look chosen for the Edison or the LEGO\textsuperscript{\textregistered} robots. 
-These last robots are implicitly targeting a group of people interested in technical systems, mostly males, while Thymio is open to both genders and a larger target audience, from 6 years old kids to adults.
+Several design choices, like the variety of sensors, the multiple ways of interacting with the robot, the neutral hull design, the various programming environments, and the possible customization with accessories, contribute to make Thymio accessible to girls and boys of different age groups from kindergarten to university~\cite{Riedo2013}.
+These design choices were made and implemented thanks to an important contribution by industrial designers of the University of Art and Design of Lausanne.\footnote{\url{http://www.ecal.ch}}
+The white neutral hull is a key element in this set of choices, and it is the opposite of the technical look chosen for the Edison or the LEGO\textsuperscript{\textregistered} robots. 
+These latter two robots implicitly target a group of people interested in technical systems, mostly males, while Thymio is open to both genders and a larger target audience.
 %The white hull is initially neutral and children can choose their own color using the \textsc{rgb} \textsc{led}s or attach accessories to the hull.
@@
 %Therefore, we thought Thymio as a support for handicrafts and constructions.
-The white neutral hull is also meant to represent a blank page that can be decorated and drawn upon, and its shape allows easy integration into a larger structure.
+The white neutral hull is also meant to represent a blank page that can be decorated and drawn upon, and the hull's shape allows easy integration into a larger structure.
 %The robot features a hole in its middle to insert a marker for drawing.
 %It also has a slot for a microSD card, allowing users to add their own sounds or to load a pre-compiled code by the simple insertion of a dedicated card.
-The square format of the hull facilitates the use of the robot as a base of the user's own constructions.
+The square format of the hull facilitates the use of the robot as a base for the user's own constructions.
 To that end, Thymio is compatible with LEGO\textsuperscript{\textregistered} bricks, both on the body than on the wheels. 
@@
 %Additionally, the two wheels have attachment points, which permits to use them to actuate elements elsewhere in the structure (Fig.~\ref{fig:example-construction}, third row left) or to lift the robot's own weight (Fig.~\ref{fig:example-construction}, third row right).
-Therefore we chose more powerful motors than strictly necessary to move the robot around.
+Therefore, we chose more powerful motors than strictly necessary to move the robot around.
 %The same fixation points used for LEGO\textsuperscript{\textregistered} bricks can be used to attach paper structures.
-Paper can also be used to change the body shape and/or add some body movements, as illustrated in Fig.~\ref{fig:example-construction} by the orca, opening and closing its mouth while moving forward or by the bat, moving its wings.
+Paper can also be used to change the body shape or add body movements, as illustrated in Fig.~\ref{fig:example-construction} by the orca, opening and closing its mouth while moving forward, or by the bat, moving its wings.
 But paper and cardboard can also radically change the locomotion principle, as illustrated in the second row of Fig.~\ref{fig:example-construction} by the zombie, where the wheels of the robot activate the legs.
-The paper structure can be also used to interact with the sensors, as illustrated in the second row of Fig.~\ref{fig:example-construction} by the bear, who extends his paw in front of the sensors to drive its iceberg (the robot).
+The paper structure can also be used to interact with the sensors, as illustrated in the second row of Fig.~\ref{fig:example-construction} by the bear, which extends its paw in front of the sensors to drive its iceberg (the robot).
 The same fixation points can be used to attach 3D printed customized parts, as illustrated by the winder shown in the fourth row of Fig.~\ref{fig:example-construction}.
-Moreover, one can use paper to create environments, either flat with patterns that can be used in association with the ground sensors (Fig.~\ref{fig:example-construction}, bottom) or 3D object, such as the trees beside the zombie in Fig.~\ref{fig:example-construction}.
+Moreover, one can use paper to create environments, either flat with patterns that can be used in association with the ground sensors (Fig.~\ref{fig:example-construction}, bottom) or 3D objects, such as the trees beside the zombie in Fig.~\ref{fig:example-construction}.
 Finally, it is also possible to link several Thymio by software, allowing the coordination of complex multi-Thymio robotic structures.
@@
 
-When designing Thymio, we took care of providing many incentives for the users to learn new things throughout their direct interaction with the robot.
+When designing Thymio, we took care to provide many incentives for the users to learn new things throughout their direct interaction with the robot.
 This translates into specific hardware and software choices.
 
-At the hardware level, we render visible the activity of the various robot components by adding a \textsc{led} next to each of them, for a total of 39 \textsc{led}s.
+At the hardware level, we render visible the activity of the various robot components by adding an \textsc{led} next to each of them, for a total of 39 \textsc{led}s.
 These \textsc{led}s locally color the hull and allow the user to see immediately where and when the robot perceives a change in its environment: proximity of objects, changes in the ground color, temperature, sound, or accelerations.
 %\textsc{led}s show, by their light intesity or their color, the measurements of the proximity sensors, the temperature sensor, the microphone, and the accelerometer. 
-Some \textsc{led}s display data exchanges, from the infrared remote control receiver or with the microSD card. 
-The capacitive buttons give both a visual and an acoustic feedback.
+Some \textsc{led}s display data exchanges from the infrared remote control receiver or with the microSD card. 
+The capacitive buttons give both visual and acoustic feedback.
 %There is a \textsc{led} next to each proximity sensor that lights up as soon as an object is close enough to be seen, and shines brighter the closer the object gets.
@@
 %In addition to a visual feedback, the capacitive buttons also have an acoustic feedback.
-The link between a sensor and its feedback can be turned off when programming the robot to use the \textsc{led}s and the loudspeaker for other purposes.
+The link between a sensor and its feedback can be turned off when programming the robot so that the \textsc{led}s and loudspeaker can be used for other purposes.
 
-At the software level, we provide a set of programming environments (see section \ref{sec:aseba}) that enable beginners to discover programming progressively.
+At the software level, we provide a set of programming environments (see the Programming Environment section below) that enable beginners to discover programming progressively.
 First, we teach them the basic rules of programming using a purely visual interface, then they discover the construction of syntax trees by assembling graphical blocks, and finally we provide a full text-based coding environment with advanced debugging tools, such as real-time inspection of the variables of the robot and plotting features, providing a visual way to understand time-related concepts.
@@
 
-\subsection{Fast access to robotics behaviors}
+\subsection{Fast Access to Robotics Behaviors}
 \label{sec:behaviors}
 
-Many existing robots need to be built or configured before having the first operational behavior. 
-For instance the Edison robot needs to read a bar code and the mBot needs to be assembled.
-This can be a barrier for school activities and we wanted to avoid it carefully, having a robot able to show interesting behaviors right out of the box.
-Therefore Thymio has six different basic behaviors accessible as soon as the robot is started and stored in flash permanently.
+Many existing robots need to be built or configured before showing any operational behavior. 
+For instance, the Edison robot needs to read a bar code and the mBot needs to be assembled.
+This can be a barrier for school activities, one we wanted to avoid; rather, we wanted a robot able to show interesting behaviors right out of the box.
+Therefore, Thymio has six different basic behaviors, stored in flash permanently, accessible as soon as the robot is started.
 These basic behaviors allow people starting Thymio to immediately interact with it, while illustrating the many possibilities of the robot.
@@
 %These behaviors can be exploited in a construction to create reactivity without the use of programming, such as in the paper creations shown in the top four images of Fig.~\ref{fig:example-construction}.
-The user can already start creating constructions on top of these basic behavior without the need of programming, such as in the paper creations shown in the top four images of Fig.~\ref{fig:example-construction}.
+The user can begin creating constructions on top of these basic behaviors without the need for programming, such as in the paper creations shown in the top four images of Fig.~\ref{fig:example-construction}.
 
@@
 
-\subsection{Programming environment}
+\subsection{Programming Environment}
 \label{sec:aseba}
 
-Thymio runs the Aseba open-source programming environment~\cite{aseba}.
+Thymio runs the Aseba open source programming environment~\cite{aseba}.
 Aseba is designed to enable novices to program robots easily.
@@
 A virtual machine allows instantaneous upload and safe execution of programs.
-On the desktop side, Aseba provides an \ac{ide} featuring a \ac{vpl} (Fig.~\ref{fig:vpl}), a scripting language (Fig.~\ref{fig:aseba-studio}), and a mixed language, Blockly\footnote{\url{https://developers.google.com/blockly/}}, to assemble scripts graphically (Fig.~\ref{fig:blockly}).
+On the desktop side, Aseba provides an \ac{ide} featuring a \ac{vpl} (Fig.~\ref{fig:vpl}), a scripting language (Fig.~\ref{fig:aseba-studio}), and a mixed language, Blockly,\footnote{\url{https://developers.google.com/blockly/}} to assemble scripts graphically (Fig.~\ref{fig:blockly}).
 These different languages cover the abilities of children of different ages and the progression of skills of learners.
@@
 
-The \ac{ide} integrates real-time feedback of the status of program execution, as this feature was recognized of critical importance to properly learn programming~\cite{sorva2013notional}.
+The \ac{ide} integrates real-time feedback on the status of program execution, as this feature was recognized as of critical importance to properly learn programming~\cite{sorva2013notional}.
 This capability is provided both with the \ac{vpl}~\cite{Magnenat2015} and in the scripting environment, by displaying the content of variables in real time through texts or plots.
@@
 Aseba integrates with \textsc{ros}~\cite{quigley2009ros} through the \emph{asebaros}\footnote{\url{http://www.ros.org/wiki/asebaros}} bridge.
-\textsc{Ros} is one of the most widely used software framework in robotics research, and this integration allows to run sophisticated algorithms, such as simultaneous localization and mapping, in conjunction with  Thymio.
+\textsc{Ros} is one of the most widely used software frameworks in robotics research, and this integration allows running sophisticated algorithms, such as simultaneous localization and mapping, in conjunction with  Thymio.
 This makes the robot suitable for university-level education.
@@
 \includegraphics[width=\columnwidth]{figures/aseba-studio}
-\caption{Aseba Studio, the integrated development environment}
+\caption{Aseba Studio, the integrated development environment.}
 \label{fig:aseba-studio}
@@
 
-The last but not least design choice among the seven mentioned in the previous section, is to have an open source hardware design.
+The last but not least design choice among the seven mentioned in the previous section is the open source hardware design.
 This choice has an impact on the robot design and the way this device is used in the community of users. 
-In this section we analyze more in detail the implications of this choice in the context of educational robotics.
-We compare these implications with the result of a survey that got 35 answers from 11 project leaders, 13 core design team members, 8 contributors and 3 enthusiastic users of open hardware projects.
+In this section, we analyze in more detail the implications of this choice in the context of educational robotics.
+We compare these implications with the results of a survey that returned 35 answers from 11 project leaders, 13 core design team members, 8 contributors, and 3 enthusiastic users of open hardware projects.
 %people active in various open source hardware projects based on a worldwide call for contributions. 
@@
 %The institution initiating a project is mainly looking for recognition and/or money. 
-Our group has a good experience in disseminating robotic hardware with the Khepera~\cite{MonFraIen93} and e-puck~\cite{mondada2009puck} robots.
-Khepera was disseminated with a proprietary strategy, e-puck with an open source hardware one.
-Both were targeting similar users and have been sold in similar quantities.
+Our group has good experience in disseminating robotic hardware with the Khepera~\cite{MonFraIen93} and e-puck~\cite{mondada2009puck} robots.
+Khepera was disseminated with a proprietary strategy, e-puck with an open source hardware strategy.
+Both targeted similar users and have been sold in similar quantities.
 What we can observe after more than 10 years is that Khepera generated royalties for the university, but the name of the robot was associated with the name of the company producing it, not the university that developed it (Fig.~\ref{fig:khepuck}).
-The e-puck robot, being open hardware and with an image better linked to the university, generated no income for the university but much more visibility.
+The e-puck robot, with open hardware and an image better linked to the university, generated no income for the university but much more visibility.
 %Khepera was more linked with the name of the company producing it, the e-puck robot was nearly not linked with the producer when mentioned in the literature.
-In the case of Thymio the institutional motivation was to get visibility more than money. 
-Therefore, from an institutional point of view, the open source hardware strategy seemed more adapted.
+In the case of Thymio, the institutional motivation was toward visibility more than money. 
+Therefore, from an institutional point of view, the open source hardware strategy seemed more adapted to the desired outcomes.
 
@@
 \includegraphics[width=\columnwidth]{figures/others_robots}
-\caption{Comparison between e-Puck and Khepera in term of mentions in google scholar, either associated with the university of with the company producing them.
+\caption{Comparison between e-Puck and Khepera in terms of mentions in Google Scholar, either associated with the university or with the company producing them.
 %NOTE: SM: this is an analysis and should not be in caption
@@
 
-Beside the institutional motivation, there is the personal motivation of each contributor.
+Along with the institutional motivation, each contributor has a personal motivation.
+%Beside the institutional motivation, there is the personal motivation of each contributor.
 %The personal motivation to participate to such a project is very different than the institutional one. 
-When asked about their personal motivation, the people participating to the survey give as main motivation a link to their professional activity and to the specific project.
+When asked about their personal motivation, the people participating in the survey cited links to their professional activity and to the specific project.
 %, followed by a more general motivation by the nature of the project and finally by the engineering practice (Fig.~\ref{fig:motivation}). 
-Very few mention a more abstract goal like improving our society.
+A few mentioned a more abstract goal like improving our society.
 \iffalse
@@
 
-Also in the Thymio project, most contributors participate to the development as part of their job, but the motivation to contribute to society is stronger, as developing a robot targeting education has an important societal component.
+While most contributors participated in the Thymio project as part of their job, they showed a strong motivation to contribute to society, as developing a robot targeting education has an important societal component.
 %were professionally involved in the professionals involved in the development of the robot because XXXX
@@
 %On another side, the motivation of contributing to society is strong in our team, as developing a robot targeting education has an important societal component.
-Moreover, our project also has a strong scientific motivation, with several ongoing studies on the acceptance by teachers and the effect on children's learning.
+Moreover, our project also has a strong scientific motivation; several studies are ongoing concerning the program's acceptance by teachers and the effect on children's learning.
 Sharing a strong fundamental motivation, such as education or scientific achievements, is a key element for building a solid community~\cite{Stahlbrost2011}, especially if it is interdisciplinary like ours.
 
-What do people expect as benefits in participating in such a project? Fig.~\ref{fig:getout} shows the answers from the survey.
+What benefits do people expect from participating in such a project? 
+Fig.~\ref{fig:getout} shows the answers from our survey.
 We can observe both a technological and a human-relations component, resulting from the community created around the project. 
 %People expect to work with other people to achieve more together, to create a network, and to learn through new experience and contacts. 
-In the Thymio project we had similar expectations. 
-Working together with several partners was for everybody a win-win situation, and creating a community of users was the only solution to allow the development of high quality accessories and educational material.
+Developers of the Thymio project had similar expectations.
+%In the Thymio project we had similar expectations. 
+Working together with several partners was for everybody a win-win situation, and creating a community of users was the only solution that allowed the development of high quality accessories and educational material.
 We established a wiki\footnote{http://www.thymio.org} as the meeting point for the learners, the robot developers, and the teachers.
 %This wiki introduces the robot, explains the basic behaviors, gives access to the programming environment and its documentation, and provides code samples and examples of constructions.
-It is open for editing by anyone, and although we initially provided most of the material, the other members of the community have started to contribute.
+It is open for editing by anyone, and although we initially provided most of the material, other members of the community have started to contribute.
 One of the important contributions from people outside the core design team was the translation of the wiki into four languages.
@@
 
-\subsection{License of project and license of tools}
+\subsection{License of Project and License of Tools}
 
 When starting an open source hardware project, one of the typical questions is which license to use when disseminating the project source files. 
-All Thymio documentation and plans are distributed under the CC-BY-SA 3.0 license and software under the LGPL license.
+All Thymio documentation and plans are distributed under the CC-BY-SA 3.0 license and the software is distributed under the LGPL license.
 We will not discuss this matter here, as it is a very common and well-covered issue.
 
-There is another licensing issue which is not well known and that we discovered very late in our project: the constraints of the license of the mechanics and electronics \textsc{cad} tools. 
-Indeed, when asked about this issue, 57\% of the participants to our survey are not aware of the fact that \textsc{cad} licenses can be very restrictive about the way the source files can be published, and only one third checked the license on this aspect.
+There is another licensing issue which is less well known and that we discovered very late in our project: the constraints of the licensing of the mechanical and electronic \textsc{cad} tools. 
+Indeed, when asked about this issue, 57\% of the participants in our survey were not aware of the fact that \textsc{cad} licenses can be very restrictive about the way source files can be published, and only one third checked the license on this aspect.
 %More than the half or respondents to our survey, and more than 60\% among the project leaders, state that they were not aware that this could be an issue, and only one third checked the license of their \textsc{cad} software.
@@
 % FIXME: SM: there is a bit of duplication with previous sentence, I think these can be compressed
-This issue is very serious, as most contributors of open source hardware projects are academics and use educational licenses, not allowing commercial use of the resulting designs. 
+This issue is very serious, as most contributors to open source hardware projects are academics and use educational licenses that do not allow commercial use of the resulting designs. 
 As producing or selling the product is part of the definition of open hardware, these licenses simply forbid publication under the standard open source hardware conditions.
@@
 %, and we specified that the published files could have been downloaded by a company to produce the system for commercial purposes.
-Fig.~\ref{fig:editors} summarizes the result of this survey.
+Fig.~\ref{fig:editors} summarizes the results of this survey.
 Only three of the twelve editors have education licenses allowing this type of publication. 
-Two others explicitly mentioned the possibility to ask for permission before publication.
-A large mechanical \textsc{cad} editor was puzzled by our questions and after realizing the impact of the license, introduced a special condition allowing publication of files in clearly-labeled open source hardware projects.
-In previous situations of open source publication, the same editor asked the universities to purchase commercial licenses to permit publication.
+Two others explicitly mentioned the possibility if permission were requested before publication.
+A large mechanical \textsc{cad} editor was puzzled by our questions and after realizing the impact of the license, introduced a special condition allowing publication of files in clearly labeled open source hardware projects.
+In previous situations concerning open source publication, the same editor asked the universities to purchase commercial licenses to permit publication.
 This can multiply by factors of hundreds the price of the \textsc{cad} license.
@@
 \includegraphics[width=\columnwidth]{figures/table}
-\caption{Publication possibilities as function of the \textsc{cad} editors, situation at the end of March 2016.}
+\caption{Publication possibilities as function of the \textsc{cad} editors, as of the end of March 2016.}
 \label{fig:editors}
@@
 
-As a conclusion, this legal issue is totally underestimated by both people participating to the projects and by the \textsc{cad} editors, and is a potential threat for many projects. 
+This legal issue is totally underestimated by both people participating in such projects and by the \textsc{cad} editors, and it is a potential threat for many projects. 
 %In a period where editors are looking for additional revenues and are attacking universities for misuse of licenses\footnote{We are aware, just in Switzerland, of two situations where large amounts of money are asked to universities for non-respect of software licenses.}, this can be a very dangerous situation.
 
-\subsection{Who designs, produces, and supports the hardware}
+\subsection{Who Designs, Produces, and Supports the Hardware}
 
 %Among the fundamental choices when starting an open source hardware project, there is the choice of the type of production. 
-In the definition of open hardware we gave, it is stated that one should offer an ``hardware whose design is made publicly available so that anyone can [\dots] make [\dots] hardware ''.
-Behind the ``anyone'' should we consider every single person or only companies able to produce the product?
+In the definition of open hardware given at the beginning of this paper, it is stated that one should offer ``hardware whose design is made publicly available so that anyone can [\dots] make [\dots] hardware.''
+Should ``anyone'' mean every single person or only companies able to produce the product?
 %This choice has implications on who designs the system and how.
 
-In our project we decided to have two different types of hardware: the robot itself and the accessories. % used in specific activities.
+In our project, we decided to have two different types of hardware: the robot itself and the accessories. % used in specific activities.
 The robot is the expensive part and has a very neutral design, allowing adaptation to specific situations.
-This adaptation is achieved by custom accessories that increase the attraction of the robot in its specific role, enabling activities for different ages and genders.
+This adaptation is achieved by custom accessories that increase the attractiveness of the robot in its specific role, enabling activities for different ages and genders.
 
-For the robot itself, we opted for the interpretation of the definition that considers under ``anyone'' only the professional structures able to mass-produce hardware, for a reason of price and complexity of the product.
+For the robot itself, we opted to interpret ``anyone'' as only the professional structures able to mass-produce hardware based on the price and complexity of the product. 
 %When looking for the best performance per price ratio, we decided to use techniques that require heavy equipment.
@@
 The philosophy of open source and free access to information fits extremely well with the community of users in education, and was reinforced by producing the robot in a non-profit structure.
-This appraoch allowed to broadly distribute the robot with minimal changes dues to management of intellectual properties, royalties, financial support and so on.
+This approach allowed to broadly distribute the robot with minimal changes dues to management of intellectual properties, royalties, financial support and so on.
 Moreover, the open source approach allows to provide a durable robot, easy to maintain and repair, with at the same time a community of users providing educational material and mutual support.

### 2016-10-20 paper_ram.tex
final changes

@@
 \author{Francesco Mondada$^{1}$, Michael Bonani$^{2}$, Fanny Riedo$^{2}$, Manon Briod$^{1}$, \\
-L\'ea Pereyre$^{1}$, Philippe R\'etornaz$^{1}$ and St\'ephane Magnenat$^{1}$% <-this % stops a space
+L\'ea Pereyre$^{1}$, Philippe R\'etornaz$^{1}$ and St\'ephane Magnenat$^{2,3}$% <-this % stops a space
 \thanks{*This research was supported by the Swiss National Center of Competence in Research ``Robotics'' (Thymio robot development and deployment), by GebertRuf Stiftung (design of accessories), by the Swiss NSF project CRAGP2 151543 ``Robotics in schools,'' and by the EU-FP7 project ASSISIbf, no.
@@
 Many thanks to Luc Bergeron and his team at \'ecal.ch for the industrial design of Thymio; Didier Roy, David Sherman and all the INRIA team for their contributions and diffusion in France; Gordana Gerber for the educational material; all the Mobsya team for the effort in production and sale; and Fabian Hahn for the port of Blockly.}% <-this % stops a space``anyone''
-\thanks{$^{1}$Francesco Mondada, Manon Briod, L\'ea Pereyre, Philippe R\'etornaz and St\'ephane Magnenat are with the Laboratoire de Syst\`emes Robotiques,
+\thanks{$^{1}$Francesco Mondada, Manon Briod, L\'ea Pereyre, and Philippe R\'etornaz are with the Laboratoire de Syst\`emes Robotiques,
         Ecole Polytechnique F\'ed\'erale de Lausanne (EPFL), Switzerland.
         {General rule for EPFL emails: \tt\small firstname.lastname@epfl.ch}}%
-\thanks{$^{2}$Michael Bonani and Fanny Riedo are with Mobsya Association, Renens, Switzerland.
+\thanks{$^{2}$Michael Bonani, Fanny Riedo and St\'ephane Magnenat are with Mobsya Association, Renens, Switzerland.
         {\tt\small michael.bonani@mobsya.org} and {\tt\small fanny.riedo@mobsya.org}}%
+\thanks{$^{3}$St\'ephane Magnenat is also with Game Technology Center, ETH Zurich, Switzerland. 
+        {\tt\small stephane@magnenat.net}}%
 }
@@
 %However, it is not a real robot but rather an automaton: There is no perception of the environment, just a good encoder for precise movements and a simple programming interface based on buttons giving directions for the movement.
-Among the other players on the market is the Edison robot\footnote{\url{https://meetedison.com/}}, which is extremely low cost (US\$49) and compatible with LEGO\textsuperscript{\textregistered} bricks.
-The low price has pushed drastic design choices: small size, weak motors, few sensors, communication with the computer by audio jack and standard AAA batteries not included in the price.
+Among the recent new players on the market is the Edison robot\footnote{\url{https://meetedison.com/}}, which is extremely low cost (US\$49), robust, and compatible with LEGO\textsuperscript{\textregistered} bricks.
+The low price has pushed drastic design choices: very few sensors, three buttons and two \textsc{led}s as user interface, and a unidirectional communication with the computer by audio jack.
 These choices strongly limit its possible usage.
@@
 %Educational material and the related training must therefore be adapted to each situation.
-Even for large companies like LEGO\textsuperscript{\textregistered}, such a development is not sustainable. 
+%Even for large companies like LEGO\textsuperscript{\textregistered}, such a development is not sustainable. 
 A crowd-sourcing approach may solve this problem; An active community of users can contribute to the development of the material in a distributed manner, adapting the material to the local situation.
@@
 The resulting Thymio robot possesses a large number of simple sensors:
-seven horizontal proximity sensors, two ground sensors, a three-axis accelerometer, a thermistor, and a microphone.
+seven horizontal proximity sensors, two infrared sensors pointing to the ground, a three-axis accelerometer, a thermistor, and a microphone.
 Five capacitive touch buttons organized as a direction pad form an intuitive user interface.
@@
 %The omnipresence of this type of sensors in smartphones and compact cameras has drastically lowered their prices over the last decade.
-We also chose low-cost toy motors and control them in speed (max. 13\,cm/s) measuring the back-electromotive force and therefore removing the need for additional encoders.
+We also chose low-cost toy motors and control them in speed (max. 13\,cm/s) measuring the back-electromotive force.
 
@@
 
-Our choice of electronic components implies production that is either fully automatized or includes some manual operations. 
+Our choice of electronic components implies different degrees of automatization in the production. 
 Full assembly is required in order for the robots to be certified for use by children. 
 %A final assembly by the user is excluded by the certification required in the case of use by children.
-As a full automatization requires investments that are beyond the possibilities of this project, the current production is based on some manual operations. 
-We therefore decided to produce in China, where manual work is less expensive.
+As a full automatization requires investments that are beyond the possibilities of this project, the current production combines automatization for most components of the \textsc{pcb} and manual operations for the final assembly, and is performed in China due to low cost of manual work.
+We have thus far produced more than 16,000 robots in batches of 2,000 units, with a cost per robot of US\$39.
+The strict quality control, the management of the production, the after-sales support, part of the development costs of the software, and the margins
+for distributors result in a final selling price of US\$130.
+
+%Our choice of electronic components implies production that is either fully automatized or includes some manual operations. 
+%Full assembly is required in order for the robots to be certified for use by children. 
+%A final assembly by the user is excluded by the certification required in the case of use by children. 
+%As a full automatization requires investments that are beyond the possibilities of this project, the current production is based on some manual operations. 
+%We therefore decided to produce in China, where manual work is less expensive.
 %This is a very critical choice in open source hardware projects, and we will discuss this issue in detail in the next section.
-We have thus far produced more than 15,000 robots in batches of 1,000 units, with a cost per robot of US\$35.
+%We have thus far produced more than 15,000 robots in batches of 1,000 units, with a cost per robot of US\$35.
 %Because manual work is expensive in western countries, we produce through subcontractors in China.
@@
 
-Several design choices, like the variety of sensors, the multiple ways of interacting with the robot, the neutral hull design, the various programming environments, and the possible customization with accessories, contribute to make Thymio accessible to girls and boys of different age groups from kindergarten to university~\cite{Riedo2013}.
+Several design choices, such as the variety of sensors, the multiple ways of interacting with the robot, the neutral hull design, the various programming environments, and the possible customization with accessories, contribute to make Thymio accessible to girls and boys of different age groups from kindergarten to university~\cite{Riedo2013}.
 These design choices were made and implemented thanks to an important contribution by industrial designers of the University of Art and Design of Lausanne.\footnote{\url{http://www.ecal.ch}}
-The white neutral hull is a key element in this set of choices, and it is the opposite of the technical look chosen for the Edison or the LEGO\textsuperscript{\textregistered} robots. 
+The white neutral hull is a key element in this set of choices, and it is the opposite of the technical look chosen for the LEGO\textsuperscript{\textregistered} robots.
+The look of Edison, designed after Thymio, is also technical due to its transparent cover.
 These latter two robots implicitly target a group of people interested in technical systems, mostly males, while Thymio is open to both genders and a larger target audience.
@@
 
-\subsection{Programming Environment}
+\subsection{Programming Environments}
 \label{sec:aseba}
@@
 
-\section{Open source hardware, choices and impact}
+\section{Open Source Hardware, Choices and Impact}
 
-The last but not least design choice among the seven mentioned in the previous section is the open source hardware design.
+The final key design choice among the seven mentioned in the previous section is the open source project.
 This choice has an impact on the robot design and the way this device is used in the community of users. 
 In this section, we analyze in more detail the implications of this choice in the context of educational robotics.
-We compare these implications with the results of a survey that returned 35 answers from 11 project leaders, 13 core design team members, 8 contributors, and 3 enthusiastic users of open hardware projects.
+We compare these implications with the results of a survey we performed among several open source hardware communities.
+we collected 35 answers from 11 project leaders, 13 core design team members, 8 contributors, and 3 enthusiastic users of open hardware projects.
 %people active in various open source hardware projects based on a worldwide call for contributions. 
@@
 Both targeted similar users and have been sold in similar quantities.
-What we can observe after more than 10 years is that Khepera generated royalties for the university, but the name of the robot was associated with the name of the company producing it, not the university that developed it (Fig.~\ref{fig:khepuck}).
-The e-puck robot, with open hardware and an image better linked to the university, generated no income for the university but much more visibility.
+What we can observe after more than 10 years is that Khepera generated royalties for the university, but the name of the robot was mostly associated with the name of the company producing it, not the university that developed it (see Fig.~\ref{fig:khepuck}).
+The e-puck robot, with open hardware and an image better linked to the university, generated no income for the university but much more relative visibility.
 %Khepera was more linked with the name of the company producing it, the e-puck robot was nearly not linked with the producer when mentioned in the literature.
@@
 \includegraphics[width=\columnwidth]{figures/others_robots}
-\caption{Comparison between e-Puck and Khepera in terms of mentions in Google Scholar, either associated with the university or with the company producing them.
+\caption{Comparison between e-puck and Khepera in terms of mentions in Google Scholar, either associated with the university or with the company producing them.
 %NOTE: SM: this is an analysis and should not be in caption
@@
 %, followed by a more general motivation by the nature of the project and finally by the engineering practice (Fig.~\ref{fig:motivation}). 
-A few mentioned a more abstract goal like improving our society.
+A few mentioned a more general goal like improving our society.
 \iffalse
@@
 The open hardware approach fits well to this requirement, as it gives to the user, or to a generic technician, better conditions to repair the system, having the specifications of all components.
-This is not the case of commercial robots like Edison, for instance.
+This is not the case of proprietary robots like Edison, for instance.
 Supporting this type of operation has an impact on the robot design; for instance Thymio can be easily opened with four standard screws, and we introduced connectors between key elements such as motors, speaker, the battery and the main \textsc{pcb}.
@@
 
-%XXXXX present a better overview of the significant improvements that the design choices bring in relation to the design goals  XXXXXXX
 The introduction of robots in formal education is a very challenging task, not only because of technical requirements such as low cost and interactivity, but also because of factors depending on the school environment, such as the diversity of the educational programs, the dependence on local structures and languages, or the required training of teachers.
